---
title: "Fatal consequences of feline coronavirus infection are associated with virus persistence and a distinct adaptive immune repertoire - BCR repertoire analysis pipeline"
author: "Tom Hiron"
date: "2026"
output: 
  html_document:
    theme: "simplex"
    toc: true
    toc_float: true
    highlight: kate
    code_folding: show
    self_contained: true
---
```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      echo = TRUE, 
                      eval = TRUE,
                      fig.path = 'figures/',
                      fig.ext = 'png',
                      cache = FALSE,
                      cache.path = 'cache/')
```

Pipeline to reproduce B Cell Receptor (BCR) repertoire analysis in the manuscript "Fatal consequences of feline coronavirus infection are associated with virus persistence and a distinct adaptive immune repertoire", Hiron et al. 2025, https://www.biorxiv.org/content/10.1101/2025.10.02.679983v1.

## Data availability

Sequencing data supporting the findings of this study have been deposited in the National Center for Biotechnology Information (NCBI) Sequence Read Archive under BioProject ID PRJNA1330231.

## Preparation of feline V(D)J sequence libraries

Created custom library from de novo search and rearranged BCR chains from Lu et al 2017.

IGHV genes: [IgDetective](https://github.com/Immunotools/IgDetective) and manual annotation of fca126 chrB3 using predicted 7mer_23bp_9mer positions. 
IGHJ genes: [StochHMM](https://github.com/KorfLab/StochHMM) on fca126 chrB3 with the J gene model included in the supplementary information of this paper: (https://www.sciencedirect.com/science/article/pii/S0165242718300941?via%3Dihub)
IGHD genes: [StochHMM](https://github.com/KorfLab/StochHMM) on fca126 chrB3 with the D gene model also included in the supplementary information of the above paper.
IGHC genes: Derived from sequences in [Lu et al 2017](https://www.nature.com/articles/s41598-017-12725-5), and from BLAST searches of canine IGHC regions against fca126. 

For this version of the reference library, an additional 5 IGHV genes detected by BLAST search of previously detected IGHV alleles against fca126 are included. This was done due to the presence of quite distant alleles (detected with MiXCR's findAlleles) detected in the rearranged IGH sequences during preliminary analysis. Some distant alleles did not map to new genes, so there may still be additional IGHV genes not found due to absence from fca126 (either due to assembly errors or structural variation). 

IGH gene FASTAs used as follows to prepare custom reference library in [MiXCR](https://mixcr.com/mixcr/guides/create-custom-library/):

```{r mixcr_custom_lib, engine = 'bash', eval = FALSE}
mixcr buildLibrary --debug \
  --v-genes-from-fasta IGHV.fa \
  --v-gene-feature VRegion \
  --j-genes-from-fasta IGHJ.fa \
  --d-genes-from-fasta IGHD.fa \
  --c-genes-from-fasta IGHC.fa \
  --chain IGH \
  --taxon-id 9685 \
  --species cat \
  feline-IGH.json.gz
```

## Assembly of IGH clonotypes from RNA-seq samples with MiXCR

MiXCR v4.7.0.

```{r mixcr_assemble_rnaseq_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J mixcr_igh_rnaseq_secondary.job 
#SBATCH -o mixcr_igh_rnaseq_secondary.%A_%a.out 
#SBATCH -c 4

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
inpath=fastq
outpath=mixcr/igh

cd $inpath

fastqs=()
for fastq in *_1.fastq.gz
do
        fastq=${fastq%_1.fastq.gz}
        fastqs[${#fastqs[@]}]=$fastq
done

cd $path

#put task here
file1=$inpath/${fastqs[$i]}_1.fastq.gz
file2=$inpath/${fastqs[$i]}_2.fastq.gz
outpath=$outpath/${fastqs[$i]}

mixcr analyze rna-seq -t 4 \
    --species cat \
    --library feline-IGH.json.gz \
    $file1 \
    $file2 \
    $outpath
```

```{r mixcr_assemble_rnaseq_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
inpath=fastq

cd $inpath

fastqs=()
for fastq in *_1.fastq.gz
do
        fastq=${fastq%_1.fastq.gz}
        fastqs[${#fastqs[@]}]=$fastq
done
fastqcount=${#fastqs[@]}

cd $path

sbatch --array 1-$fastqcount:1 mixcr_assemble_rnaseq_secondary.sh
```

## Clonotype analysis

Immunarch (https://immunarch.com/).

Load MiXCR data.

```{r immunarch_load}
library(immunarch)
library(stringr)

igh_mixcr <- repLoad('mixcr/IGH/')

# set metadata and sample names
igh_mixcr$meta$Sample <- names(igh_mixcr$data)
sampleGroups <- factor(c(rep('Control', 21), rep('FIP', 21)), levels = c('Control', 'FIP'))

names(igh_mixcr$data) <- str_remove(names(igh_mixcr$data), 'feline_')
names(igh_mixcr$data) <- str_remove(names(igh_mixcr$data), '_merged')
names(igh_mixcr$data) <- str_remove(names(igh_mixcr$data), '.clones_IGH')

igh_mixcr$meta$Group <- sampleGroups

igh_mixcr$meta$Sample <- str_remove(igh_mixcr$meta$Sample, 'feline_')
igh_mixcr$meta$Sample <- str_remove(igh_mixcr$meta$Sample, '_merged')
igh_mixcr$meta$Sample <- str_remove(igh_mixcr$meta$Sample, '.clones_IGH')

# add tissue to metadata
igh_mixcr$meta$Tissue <- ifelse(grepl('liver', igh_mixcr$meta$Sample), 'Liver', ifelse(grepl('lung', igh_mixcr$meta$Sample), 'Lung', 'MLN'))
```

## Clone and clonotype stats - all tissues

```{r immunarch_basic}
# first, plot the number of clones detected in each sample, faceted by tissue
igh.clns <- data.frame(lapply(igh_mixcr$data, function(x) sum(x$Clones)))
igh.clns <- as.data.frame(t(igh.clns))
igh.clns$sample <- rownames(igh.clns)
igh.clns$group <- ifelse(grepl('control', igh.clns$sample), 'Control', 'FIP')
igh.clns$tissue <- ifelse(grepl('liver', igh.clns$sample), 'Liver', ifelse(grepl('lung', igh.clns$sample), 'Lung', 'MLN'))
colnames(igh.clns) <- c('Clones', 'Sample', 'Group', 'Tissue')
#write to table to add IHC results
# write.table(igh.clns, file = 'tables/rnaseq_igh_clns.txt', sep = '\t')

library(ggplot2)
library(ggrepel)
library(cowplot)

igh.clones.gg <- ggplot(igh.clns, aes(Sample, log10(Clones), fill = Group)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  # geom_hline(yintercept = c(3, 3.30103, 3.69897), color = 'black', lty = 'dashed') +
  # scale_y_continuous(expand = c(0,0), limits = c(0, 5)) +
  facet_wrap(vars(Tissue), nrow = 1, scales = 'free_x') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# number of clonotypes
igh.exp_vol <- repExplore(igh_mixcr$data, .method = 'volume')
igh.exp_vol$Group <- igh.clns$Group
igh.exp_vol$Tissue <- igh.clns$Tissue
igh.clntypes.gg <- ggplot(igh.exp_vol, aes(Sample, log10(Volume), fill = Group)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  # scale_y_continuous(expand = c(0,0), limits = c(0, 4.5)) +
  facet_wrap(vars(Tissue), nrow = 1, scales = 'free_x') +
  labs(x = 'Sample', y = 'Clonotypes') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

plot_grid(igh.clones.gg, igh.clntypes.gg,
          align = 'v', 
          ncol = 1,
          axis = 'lr')

# plot clones vs clonotypes
igh.clns.clntypes <- igh.clns
igh.clns.clntypes$Volume <- igh.exp_vol$Volume
# read in IHC data
rnaseq.ihc <- read.csv('rnaseq_igh_ihc.txt', sep = '\t', header = TRUE)
igh.clns.clntypes <- merge(igh.clns.clntypes, rnaseq.ihc, by = 'Sample')
# also add to metadata
igh_mixcr$meta <- merge(igh_mixcr$meta, rnaseq.ihc, by = 'Sample', all.x = TRUE)

ggplot(igh.clns.clntypes, aes(log10(Clones), log10(Volume), color = IHC)) +
  geom_point() +
  geom_text_repel(aes(label = Sample)) +
  theme_bw()

# CDR3 lengths
igh.exp_len <- repExplore(igh_mixcr$data, .method = "len", .col = "aa")
vis(igh.exp_len, .by = c('Group'), .meta = igh_mixcr$meta)

# finally, plot clones and clonotypes by tissue and group
clns.box <- ggplot(igh.clns.clntypes, aes(Group, log10(Clones), group = Group, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Group), width = 0.1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  stat_compare_means() +
  labs(x = '', y = expression(log[10]*' clones'), fill = '', color = '') +
  facet_wrap(vars(Tissue), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())

clntyp.box <- ggplot(igh.clns.clntypes, aes(Group, log10(Volume), group = Group, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Group), width = 0.1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  stat_compare_means() +
  labs(x = '', y = expression(log[10]*' clonotypes'), fill = '', color = '') +
  facet_wrap(vars(Tissue), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())

plot_grid(clns.box, clntyp.box, nrow = 2)
```

Filter out CDR3s of aberrant length (prior to downsampling). 

```{r immunarch_filtering}
# filter clonotypes by CDR3 length 
for(i in seq_along(igh_mixcr$data)) {
  igh_mixcr$data[[i]]$CDR3.length <- nchar(igh_mixcr$data[[i]]$CDR3.aa)
}

igh_mixcr.f <- repFilter(igh_mixcr, .method = 'by.clonotype', .query = list(CDR3.length = lessthan(35)))

# visualise new CDR3 aa length distribution
igh.filt_len <- repExplore(igh_mixcr.f$data, .method = "len", .col = "aa")
vis(igh.filt_len, .by = 'Group', .meta = igh_mixcr.f$meta)

table(igh_mixcr.f$meta$Group)
igh.f.exp_vol <- repExplore(igh_mixcr.f$data, .method = "volume")
vis(igh.f.exp_vol)
# write.table(igh.f.exp_vol, file = 'tables/igh_clonotypes.txt', sep = '\t')
```

## Downsample MLN clonotypes

```{r immunarch_mln_downsample}
min(unlist(lapply(igh_mixcr.f$data[grep("mln", names(igh_mixcr.f$data))], function(x) sum(x$Clones))))
# IGH - set to 51568 - min sample clones after filtering (control_174_mln)
mln_mixcr.f <- list()
mln_mixcr.f$data <- repFilter(igh_mixcr.f, .method = "by.meta", .query = list(Tissue = include("MLN")))$data
mln_mixcr.f$meta <- repFilter(igh_mixcr.f, .method = "by.meta", .query = list(Tissue = include("MLN")))$meta
mln_mixcr.f2 <- list()
mln_mixcr.f2$data <- repFilter(mln_mixcr.f, .method = "by.repertoire", .query = list(n_clones = morethan(50000)))$data
mln_mixcr.f2$meta <- repFilter(mln_mixcr.f, .method = "by.repertoire", .query = list(n_clones = morethan(50000)))$meta
mln_mixcr.f.ds <- mln_mixcr.f2
mln_mixcr.f.ds$data <- repSample(mln_mixcr.f2$data, .method = 'downsample', .n = 51568)
# export for additional analyses
# saveRDS(mln_mixcr.f.ds, file = 'mln_mixcr_f_ds.rds')
```

```{r immunarch_mln_stats}
library(ggplot2)
library(ggpubr)
library(cowplot)

# get clonality
mln_rare <- repClonality(mln_mixcr.f.ds$data, .method = 'rare')

# re-plot with new sample order based on Group/FCoV abundance
mln.sample_order <- c("control_54_mln", "control_69_mln", "control_135_mln", "control_136_mln", "control_174_mln", "control_188_mln", "control_164_mln", "fip_142_mln", "fip_192_mln", "fip_37_mln", "fip_96_mln", "fip_120_mln", "fip_172_mln", "fip_68_mln", "fip_86_mln")

mln_rare <- mln_rare[mln.sample_order,]

# borrowed preprocessing code from immunarch github - https://github.com/immunomind/immunarch/blob/master/R/vis.R
  tmp <- mln_rare
  if (is.null(dim(tmp))) {
    tmp <- t(as.matrix(tmp))
    mln_rare <- t(as.matrix(mln_rare))
  }

  colnames_new <- c()
  prev_value <- "1"
  start_index <- 1
  if (colnames(mln_rare)[1] == "1") {
    colnames_new <- "1"
    prev_value <- "2"
    start_index <- 2
  }
  for (i in start_index:ncol(mln_rare)) {
    colnames_new <- c(colnames_new, paste0(prev_value, " - ", colnames(mln_rare)[i]))
    if (colnames(mln_rare)[i] != "MAX") {
      prev_value <- as.character(as.integer(colnames(mln_rare)[i]) + 1)
    }
  }
  colnames(tmp) <- colnames_new

  for (i in 2:ncol(mln_rare)) {
    tmp[, i] <- mln_rare[, i] - mln_rare[, i - 1]
  }
  mln.res <- tmp

  mln.res <- as.data.frame(mln.res)
  mln.res$Sample <- row.names(mln.res)
  mln.res <- reshape2::melt(mln.res)
  colnames(mln.res) <- c("Sample", "Counts", "Value")
  
  mln.res$Sample <- factor(mln.res$Sample, levels = mln.sample_order)

ggplot(mln.res, aes(Sample, Value, group = Counts, fill = Counts)) +
  geom_bar(stat = 'identity', position = 'stack', color = 'black', width = 1) +
  scale_y_continuous(expand = c(0,0), labels = seq(0,100,25)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c('#FF4B20','#FF9F30','#DCDFA2','#A8E6F3','#63ABEC','#0348A6')) +
  labs(x = '', y = 'Occupied repertoire space (%)', fill = 'Clonotype counts') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# diversity estimates
# chao
alpha.chao <- repDiversity(mln_mixcr.f.ds$data, 'chao1')

# nice boxplots
alpha.chao.df <- as.data.frame(alpha.chao)
alpha.chao.df$IHC <- mln_mixcr.f.ds$meta$IHC

compare_means(Estimator ~ IHC, data = alpha.chao.df)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

alpha.chao.gg <- ggplot(alpha.chao.df, aes(IHC, Estimator, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 40000)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 37000) +
  labs(x = '', y = 'Chao1 index') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# gini
mln.div_gini <- repDiversity(mln_mixcr.f.ds$data, 'gini')
mln.div_gini <- as.data.frame(mln.div_gini)
mln.div_gini$group <- mln_mixcr.f.ds$meta$Group
mln.div_gini$IHC <- mln_mixcr.f.ds$meta$IHC

# add rna-seq fcov abundance
fcov <- read.csv('fcov_per_sample.txt', sep = '\t', header = TRUE)
fcov$Sample <- rownames(fcov)
fcov.mln <- fcov[fcov$tissue == 'MLN',]
mln.div_gini$Sample <- rownames(mln.div_gini)
mln.div_gini <- merge(mln.div_gini, fcov.mln, by = 'Sample')
mln.div_gini$fcov <- ifelse(mln.div_gini$value > -6.385045, '+', '-')

# get p values
compare_means(V1 ~ fcov, data = mln.div_gini)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

alpha.gini.gg <- ggplot(mln.div_gini, aes(fcov, V1, group = fcov, fill = fcov)) +
  geom_boxplot(outlier.shape = NA, show.legend = TRUE, alpha = 0.5) +
  geom_jitter(aes(color = group), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  scale_y_continuous(limits = c(0, 1)) +
  stat_compare_means() +
  stat_compare_means(label.y = 1) +
  labs(x = '', y = 'Gini coefficient') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# correlation between gini and fcov
mln.div_gini2 <- mln.div_gini[mln.div_gini$group == 'FIP',]
mln.div_gini2$group <- c('FIP', 'FIP (focal renal)', 'FIP', 'FIP (focal renal)', rep('FIP', 4))

ggplot(mln.div_gini2, aes(value, V1)) +
  geom_smooth(method = 'lm', se = FALSE, color = 'darkgrey', lty = 'dashed') +
  geom_point(aes(colour = group)) +
  labs(x = expression('FCoV '*log[2]*' CPM'), y = 'Gini coefficient', colour = '') +
  scale_color_manual(values = c('#8F3092', '#D071AD')) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Calculate isotype usage for MLN samples.

```{r isotype_mln}
# for list of dfs, get IGHC name and frequency
isotype.mln <- lapply(mln_mixcr.f.ds$data, `[`, c('C.name', 'Proportion'))

library(tidyverse)
# remove alignment scores and secondary alignments, and collapse IGHG subtype
isotype.mln <- isotype.mln %>% 
  map(~ .x %>% mutate(C.name = gsub('\\(.{2,7}\\)', '', C.name))) %>%
  map(~ .x %>% mutate(C.name = gsub(', .*', '', C.name))) %>%
  map(~ .x %>% mutate(C.name = gsub('IGHG1|IGHG2|IGHG3', 'IGHG', C.name)))

# normalise as proportion of unique clonotypes
isotype.mln.summ <- lapply(isotype, function(x) {
  x %>%
  group_by(C.name) %>%
  summarise(Count = n()) %>%
  mutate(Proportion = Count/nrow(x)) %>%
  complete(C.name = c('IGHM', 'IGHA', 'IGHD', 'IGHE', 'IGHG'), fill = list(Proportion = 0))
})

# add sample name and combine
isotype.mln.all <- bind_rows(isotype.mln.summ, .id = 'Sample')

# replace NAs with 'Other'
isotype.mln.all$C.name[is.na(isotype.mln.all$C.name)] <- 'Other'

# set sample order
isotype.mln.all$Sample <- factor(isotype.mln.all$Sample, levels = mln.sample_order)

# plot
library(ggplot2)
library(ggsci)

ggplot(isotype.mln.all, aes(Sample, Proportion, fill = C.name, group = Sample)) +
  geom_bar(stat = 'identity', position = 'stack') +
  scale_fill_d3() +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = '', y = 'Proportion of clonotypes', fill = '') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# plot frequencies by group
isotype.mln.all$Group <- ifelse(grepl('control', isotype.mln.all$Sample), 'Control', 'FIP')

ggplot(isotype.mln.all, aes(Group, Proportion, group = Group, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(colour = Group), width = 0.1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  stat_compare_means() +
  facet_wrap(vars(C.name), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())

compare_means(Proportion ~ Group, data = isotype.mln.all, group.by = 'C.name', method = 'wilcox.test', p.adjust.method = 'bonferroni')
```

## Downsample Liver clonotypes

```{r immunarch_liver_downsample}
min(unlist(lapply(igh_mixcr.f$data[grep("liver", names(igh_mixcr.f$data))], function(x) sum(x$Clones))))
# IGH - set to 4568 - drops control_174_liver and fip_120_liver
liver_mixcr.f <- list()
liver_mixcr.f$data <- repFilter(igh_mixcr.f, .method = "by.meta", .query = list(Tissue = include("Liver")))$data
liver_mixcr.f$meta <- repFilter(igh_mixcr.f, .method = "by.meta", .query = list(Tissue = include("Liver")))$meta
liver_mixcr.f2 <- list()
liver_mixcr.f2$data <- repFilter(liver_mixcr.f, .method = "by.repertoire", .query = list(n_clones = morethan(4560)))$data
liver_mixcr.f2$meta <- repFilter(liver_mixcr.f, .method = "by.repertoire", .query = list(n_clones = morethan(4560)))$meta
liver_mixcr.f.ds <- liver_mixcr.f2
liver_mixcr.f.ds$data <- repSample(liver_mixcr.f2$data, .method = 'downsample', .n = 4568)
# export for additional analyses
# saveRDS(liver_mixcr.f.ds, file = 'liver_mixcr_f_ds.rds')
```

```{r immunarch_liver_stats}
library(ggplot2)
library(ggpubr)
library(cowplot)

# get clonality
liver_rare <- repClonality(liver_mixcr.f.ds$data, .method = 'rare')

# re-plot with new sample order based on Group/FCoV abundance
liver.sample_order <- c("control_54_liver", "control_69_liver", "control_135_liver", "control_136_liver", "control_188_liver", "control_164_liver", "fip_142_liver", "fip_192_liver", "fip_96_liver", "fip_172_liver", "fip_37_liver", "fip_68_liver", "fip_86_liver")

liver_rare <- liver_rare[liver.sample_order,]

# borrowed preprocessing code from immunarch github - https://github.com/immunomind/immunarch/blob/master/R/vis.R
  tmp <- liver_rare
  if (is.null(dim(tmp))) {
    tmp <- t(as.matrix(tmp))
    liver_rare <- t(as.matrix(liver_rare))
  }

  colnames_new <- c()
  prev_value <- "1"
  start_index <- 1
  if (colnames(liver_rare)[1] == "1") {
    colnames_new <- "1"
    prev_value <- "2"
    start_index <- 2
  }
  for (i in start_index:ncol(liver_rare)) {
    colnames_new <- c(colnames_new, paste0(prev_value, " - ", colnames(liver_rare)[i]))
    if (colnames(liver_rare)[i] != "MAX") {
      prev_value <- as.character(as.integer(colnames(liver_rare)[i]) + 1)
    }
  }
  colnames(tmp) <- colnames_new

  for (i in 2:ncol(liver_rare)) {
    tmp[, i] <- liver_rare[, i] - liver_rare[, i - 1]
  }
  liver.res <- tmp

  liver.res <- as.data.frame(liver.res)
  liver.res$Sample <- row.names(liver.res)
  liver.res <- reshape2::melt(liver.res)
  colnames(liver.res) <- c("Sample", "Counts", "Value")
  
  liver.res$Sample <- factor(liver.res$Sample, levels = liver.sample_order)

ggplot(liver.res, aes(Sample, Value, group = Counts, fill = Counts)) +
  geom_bar(stat = 'identity', position = 'stack', color = 'black', width = 1) +
  # scale_y_continuous(expand = c(0,0), labels = seq(0,100,25)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c('#FF4B20','#FF9F30','#DCDFA2','#A8E6F3','#63ABEC','#0348A6')) +
  labs(x = '', y = 'Occupied repertoire space (%)', fill = 'Clonotype counts') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# diversity estimates
# chao
alpha.chao <- repDiversity(liver_mixcr.f.ds$data, 'chao1')

# nice boxplots
alpha.chao.df <- as.data.frame(alpha.chao)
alpha.chao.df$IHC <- liver_mixcr.f.ds$meta$IHC

compare_means(Estimator ~ IHC, data = alpha.chao.df)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

alpha.chao.gg <- ggplot(alpha.chao.df, aes(IHC, Estimator, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 40000)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 37000) +
  labs(x = '', y = 'Chao1 index') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# gini
liver.div_gini <- repDiversity(liver_mixcr.f.ds$data, 'gini')
liver.div_gini <- as.data.frame(liver.div_gini)
liver.div_gini$group <- liver_mixcr.f.ds$meta$Group
liver.div_gini$IHC <- liver_mixcr.f.ds$meta$IHC

# get p values
compare_means(V1 ~ IHC, data = liver.div_gini)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

# add rna-seq fcov abundance
fcov.liver <- fcov[fcov$tissue == 'Liver',]
liver.div_gini$Sample <- rownames(liver.div_gini)
liver.div_gini <- merge(liver.div_gini, fcov.mln, by = 'Sample')
liver.div_gini$fcov <- ifelse(liver.div_gini$value > -6.385045, '+', '-')

alpha.gini.gg <- ggplot(liver.div_gini, aes(IHC, V1, group = IHC, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, show.legend = TRUE, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 1)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 1) +
  labs(x = '', y = 'Gini coefficient') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# orrelation between gini and fcov
liver.div_gini2 <- liver.div_gini[liver.div_gini$group == 'FIP',]
liver.div_gini2$group <- c('FIP (focal renal)', 'FIP', 'FIP (focal renal)', rep('FIP', 4))

ggplot(liver.div_gini2, aes(value, V1)) +
  geom_smooth(method = 'lm', se = FALSE, color = 'darkgrey', lty = 'dashed') +
  geom_point(aes(colour = group)) +
  labs(x = expression('FCoV '*log[2]*' CPM'), y = 'Gini coefficient', colour = '') +
  scale_color_manual(values = c('#8F3092', '#D071AD')) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Calculate isotype usage for Liver samples.

```{r isotype_liver}
# for list of dfs, get IGHC name and frequency
isotype.liver <- lapply(liver_mixcr.f.ds$data, `[`, c('C.name', 'Proportion'))

library(tidyverse)
# remove alignment scores and secondary alignments, and collapse IGHG subtype
isotype.liver <- isotype.liver %>% 
  map(~ .x %>% mutate(C.name = gsub('\\(.{2,7}\\)', '', C.name))) %>%
  map(~ .x %>% mutate(C.name = gsub(', .*', '', C.name))) %>%
  map(~ .x %>% mutate(C.name = gsub('IGHG1|IGHG2|IGHG3', 'IGHG', C.name)))

# count of clonotypes by isotype
isotype.liver.summ <- lapply(isotype.liver, function(x) {
  x %>%
  group_by(C.name) %>%
  summarise(Count = n()) %>%
  mutate(Proportion = Count/nrow(x)) %>%
  complete(C.name = c('IGHM', 'IGHA', 'IGHD', 'IGHE', 'IGHG'), fill = list(Proportion = 0))
})

# add sample name and combine
isotype.liver.all <- bind_rows(isotype.liver.summ, .id = 'Sample')

# replace NAs with 'Other'
isotype.liver.all$C.name[is.na(isotype.liver.all$C.name)] <- 'Other'

# set sample order
isotype.liver.all$Sample <- factor(isotype.liver.all$Sample, levels = liver.sample_order)

# plot
library(ggplot2)
library(ggsci)

ggplot(isotype.liver.all, aes(Sample, Proportion, fill = C.name, group = Sample)) +
  geom_bar(stat = 'identity', position = 'stack') +
  scale_fill_d3() +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = '', y = 'Fraction of clonotypes', fill = '') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# plot frequencies by group
isotype.liver.all$Group <- ifelse(grepl('control', isotype.liver.all$Sample), 'Control', 'FIP')

ggplot(isotype.liver.all, aes(Group, Proportion, group = Group, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(colour = Group), width = 0.1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  stat_compare_means() +
  facet_wrap(vars(C.name), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())

compare_means(Proportion ~ Group, data = isotype.liver.all, group.by = 'C.name', method = 'wilcox.test', p.adjust.method = 'bonferroni')
```

## Downsample Lung clonotypes

```{r immunarch_lung_downsample}
min(unlist(lapply(igh_mixcr.f$data[grep("lung", names(igh_mixcr.f$data))], function(x) sum(x$Clones))))
# IGH - set to 11576 
lung_mixcr.f <- list()
lung_mixcr.f$data <- repFilter(igh_mixcr.f, .method = "by.meta", .query = list(Tissue = include("Lung")))$data
lung_mixcr.f$meta <- repFilter(igh_mixcr.f, .method = "by.meta", .query = list(Tissue = include("Lung")))$meta
lung_mixcr.f2 <- list()
lung_mixcr.f2$data <- repFilter(lung_mixcr.f, .method = "by.repertoire", .query = list(n_clones = morethan(11500)))$data
lung_mixcr.f2$meta <- repFilter(lung_mixcr.f, .method = "by.repertoire", .query = list(n_clones = morethan(11500)))$meta
lung_mixcr.f.ds <- lung_mixcr.f2
lung_mixcr.f.ds$data <- repSample(lung_mixcr.f2$data, .method = 'downsample', .n = 11576)
# export for additional analyses
# saveRDS(lung_mixcr.f.ds, file = 'lung_mixcr_f_ds.rds')
```

```{r immunarch_lung_stats}
library(ggplot2)
library(ggpubr)
library(cowplot)

# get clonality
lung_rare <- repClonality(lung_mixcr.f.ds$data, .method = 'rare')

# re-plot with new sample order based on group/FCoV abundance
lung.sample_order <- c("control_54_lung", "control_69_lung", "control_135_lung", "control_136_lung", "control_174_lung", "control_188_lung", "control_164_lung", "fip_142_lung", "fip_192_lung", "fip_68_lung", "fip_86_lung", "fip_172_lung")

lung_rare <- lung_rare[lung.sample_order,]

# borrowed preprocessing code from immunarch github - https://github.com/immunomind/immunarch/blob/master/R/vis.R
  tmp <- lung_rare
  if (is.null(dim(tmp))) {
    tmp <- t(as.matrix(tmp))
    lung_rare <- t(as.matrix(lung_rare))
  }

  colnames_new <- c()
  prev_value <- "1"
  start_index <- 1
  if (colnames(lung_rare)[1] == "1") {
    colnames_new <- "1"
    prev_value <- "2"
    start_index <- 2
  }
  for (i in start_index:ncol(lung_rare)) {
    colnames_new <- c(colnames_new, paste0(prev_value, " - ", colnames(lung_rare)[i]))
    if (colnames(lung_rare)[i] != "MAX") {
      prev_value <- as.character(as.integer(colnames(lung_rare)[i]) + 1)
    }
  }
  colnames(tmp) <- colnames_new

  for (i in 2:ncol(lung_rare)) {
    tmp[, i] <- lung_rare[, i] - lung_rare[, i - 1]
  }
  lung.res <- tmp

  lung.res <- as.data.frame(lung.res)
  lung.res$Sample <- row.names(lung.res)
  lung.res <- reshape2::melt(lung.res)
  colnames(lung.res) <- c("Sample", "Counts", "Value")
  
  lung.res$Sample <- factor(lung.res$Sample, levels = lung.sample_order)

ggplot(lung.res, aes(Sample, Value, group = Counts, fill = Counts)) +
  geom_bar(stat = 'identity', position = 'stack', color = 'black', width = 1) +
  # scale_y_continuous(expand = c(0,0), labels = seq(0,100,25)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c('#FF4B20','#FF9F30','#DCDFA2','#A8E6F3','#63ABEC','#0348A6')) +
  labs(x = '', y = 'Occupied repertoire space (%)', fill = 'Clonotype counts') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# diversity estimates
# chao
alpha.chao <- repDiversity(lung_mixcr.f.ds$data, 'chao1')

# nice boxplots
alpha.chao.df <- as.data.frame(alpha.chao)
alpha.chao.df$IHC <- lung_mixcr.f.ds$meta$IHC

compare_means(Estimator ~ IHC, data = alpha.chao.df)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

alpha.chao.gg <- ggplot(alpha.chao.df, aes(IHC, Estimator, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 40000)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 37000) +
  labs(x = '', y = 'Chao1 index') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# gini
lung.div_gini <- repDiversity(lung_mixcr.f.ds$data, 'gini')
lung.div_gini <- as.data.frame(lung.div_gini)
lung.div_gini$group <- lung_mixcr.f.ds$meta$Group
lung.div_gini$IHC <- lung_mixcr.f.ds$meta$IHC

# add rna-seq fcov abundance
fcov.lung <- fcov[fcov$tissue == 'Lung',]
lung.div_gini$Sample <- rownames(lung.div_gini)
lung.div_gini <- merge(lung.div_gini, fcov.lung, by = 'Sample')
lung.div_gini$fcov <- ifelse(lung.div_gini$value > -6.385045, '+', '-')

# get p values
# compare_means(V1 ~ IHC, data = lung.div_gini)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

alpha.gini.gg <- ggplot(lung.div_gini, aes(IHC, V1, group = IHC, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, show.legend = TRUE, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 1)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 1) +
  labs(x = '', y = 'Gini coefficient') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# correlation between gini and fcov
lung.div_gini2 <- lung.div_gini[lung.div_gini$group == 'FIP',]
lung.div_gini2$group <- c('FIP (focal renal)', 'FIP', 'FIP (focal renal)', rep('FIP', 2))

ggplot(lung.div_gini2, aes(value, V1)) +
  geom_smooth(method = 'lm', se = FALSE, color = 'darkgrey', lty = 'dashed') +
  geom_point(aes(colour = group)) +
  labs(x = expression('FCoV '*log[2]*' CPM'), y = 'Gini coefficient', colour = '') +
  scale_color_manual(values = c('#8F3092', '#D071AD')) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Calculate isotype usage for Lung samples.

```{r isotype_lung}
# for list of dfs, get IGHC name and frequency
isotype.lung <- lapply(lung_mixcr.f.ds$data, `[`, c('C.name', 'Proportion'))

library(tidyverse)
# remove alignment scores and secondary alignments, and collapse IGHG subtype
isotype.lung <- isotype.lung %>% 
  map(~ .x %>% mutate(C.name = gsub('\\(.{2,7}\\)', '', C.name))) %>%
  map(~ .x %>% mutate(C.name = gsub(', .*', '', C.name))) %>%
  map(~ .x %>% mutate(C.name = gsub('IGHG1|IGHG2|IGHG3', 'IGHG', C.name)))

# count of clonotypes by isotype
isotype.lung.summ <- lapply(isotype.lung, function(x) {
  x %>%
  group_by(C.name) %>%
  summarise(Count = n()) %>%
  mutate(Proportion = Count/nrow(x)) %>%
  complete(C.name = c('IGHM', 'IGHA', 'IGHD', 'IGHE', 'IGHG'), fill = list(Proportion = 0))
})

# add sample name and combine
isotype.lung.all <- bind_rows(isotype.lung.summ, .id = 'Sample')

# replace NAs with 'Other'
isotype.lung.all$C.name[is.na(isotype.lung.all$C.name)] <- 'Other'

# set sample order
isotype.lung.all$Sample <- factor(isotype.lung.all$Sample, levels = lung.sample_order)

# plot
library(ggplot2)
library(ggsci)

ggplot(isotype.lung.all, aes(Sample, Proportion, fill = C.name, group = Sample)) +
  geom_bar(stat = 'identity', position = 'stack') +
  scale_fill_d3() +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = '', y = 'Fraction of clonotypes', fill = '') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# plot frequencies by group
isotype.lung.all$Group <- ifelse(grepl('control', isotype.lung.all$Sample), 'Control', 'FIP')

ggplot(isotype.lung.all, aes(Group, Proportion, group = Group, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(colour = Group), width = 0.1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  stat_compare_means() +
  facet_wrap(vars(C.name), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())

compare_means(Proportion ~ Group, data = isotype.lung.all, group.by = 'C.name', method = 'wilcox.test', p.adjust.method = 'bonferroni')
```

## All tissues

Get stacked bar plots showing proportion of repertoire occupied by clonotypes with specific counts, faceted by tissue. 

```{r alltissues_clntypcounts}
# rbind dataframes
all.res <- rbind(liver.res, lung.res, mln.res)

# add tissue
all.res$tissue <- ifelse(grepl('liver', all.res$Sample), 'Liver', 
                         ifelse(grepl('lung', all.res$Sample), 'Lung', 
                                'MLN'))

# plot
ggplot(all.res, aes(Sample, Value, group = Counts, fill = Counts)) +
  geom_bar(stat = 'identity', position = 'stack', color = 'black', width = 1) +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), breaks = seq(0,1,0.2)) +
  scale_fill_manual(values = c('#FF4B20','#FF9F30','#DCDFA2','#A8E6F3','#63ABEC','#0348A6')) +
  labs(x = '', y = 'Proportion of clonotypes', fill = 'Clonotype counts') +
  facet_grid(cols = vars(tissue), scales = 'free_x', space = 'free') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(panel.grid = element_blank())
```

Plot correlation of Gini coefficient and FCoV abundance in all tissues. 

```{r alltissues_ginifcovcorr}
# rbind datasets
all.div_gini <- rbind(liver.div_gini2, lung.div_gini2, mln.div_gini2)

# plot
ggplot(all.div_gini, aes(value, V1)) +
  geom_smooth(method = 'lm', se = FALSE, color = 'darkgrey', lty = 'dashed', fullrange = TRUE) +
  geom_point(aes(colour = group)) +
  labs(x = expression('FCoV '*log[2]*' CPM'), y = 'Gini coefficient', colour = '') +
  scale_color_manual(values = c('#8F3092', '#D071AD')) +
  facet_wrap(vars(tissue), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Get gene usage stats and plot heatmap/correlation for all samples (normalised as proportion of clonotypes).

First fix the functions distributed with immunarch.

```{r immunarch_gene_usage_altfunctions}
geneUsage2 <- function(.data, # df, list, MonetDB
                      .gene = c("hs.trbv", "HomoSapiens.TRBJ", "macmul.IGHV"),
                      .quant = c(NA, "count"),
                      .ambig = c("inc", "exc", "maj"),
                      .type = c("segment", "allele", "family"),
                      .norm = FALSE) {
  .type <- .type[1]
  .ambig <- .ambig[1]
  .quant <- .quant[1]
  .gene <- .gene[1]
  
  if (immunarch:::has_class(.data, "list")) {
    if (length(.data) == 1) {
      return(geneUsage2(.data[[1]], .gene = .gene, .quant = .quant, .ambig = .ambig, .norm = .norm))
    } else {
      genedata <- lapply(.data, geneUsage2,
                         .gene = .gene, .quant = .quant,
                         .ambig = .ambig, .type = .type, .norm = .norm
      )
      
      # Add check if list has 1 element
      for (i in 2:(length(genedata))) {
        if (i > 2) {
          allres <- full_join(allres, genedata[[i]], by = "Names")
          names(allres)[i + 1] <- names(genedata)[i]
        } else {
          allres <- full_join(genedata[[1]], genedata[[2]], by = "Names")
          names(allres)[2] <- names(genedata)[1]
          names(allres)[3] <- names(genedata)[2]
        }
      }
      allres <- allres[order(allres$Names), ]
      
      return(immunarch:::add_class(allres, "immunr_gene_usage"))
    }
  }
  
  "%!in%" <- function(x, y) !("%in%"(x, y))
  
  which_gene <- strsplit(.gene, ".", TRUE)[[1]][2]
  gene_col <- tolower(substr(which_gene, 4, 4))
  if (gene_col == "v") {
    gene_col <- immunarch:::IMMCOL$v
  } else if (gene_col == "d") {
    gene_col <- immunarch:::IMMCOL$d
  } else if (gene_col == "j") {
    gene_col <- immunarch:::IMMCOL$j
  } else {
    stop("The entered gene_col name is invalid")
  }
  
  .gene <- get_genes2(.gene, .type)
  
  # Bad Cases:
  # X.name is segments, .type == "alleles"
  # X.name is families, .type == "segments"
  # X.name is families, .type == "alleles"
  
  if (.ambig == "inc") {
    if (.type == "segment") {
      .data[[gene_col]] <- immunarch:::return_segments(.data[[gene_col]])
    } else if (.type == "family") {
      .data[[gene_col]] <- immunarch:::return_families(.data[[gene_col]])
    }
  }
  
  if (immunarch:::has_class(.data, "data.table")) {
    .data <- .data %>% lazy_dt()
  }
  dataset <- .data %>%
    select(Gene = gene_col, immunarch:::IMMCOL$count) %>%
    group_by(Gene) %>%
    rename(Quant = immunarch:::IMMCOL$count)
  
  if (is.na(.quant)) {
    dataset <- dataset %>%
      summarise(count = n())
  } else {
    dataset <- dataset %>%
      summarise(count = sum(Quant))
  }
  
  dataset <- dataset %>%
    collect(n = Inf)
  
  if (.ambig == "inc") {
    res <- dataset
    names(res) <- c("Names", immunarch:::IMMCOL$count)
  } else {
    gene_col_split <- strsplit(dataset[["Gene"]], ", ", fixed = TRUE, useBytes = TRUE)
    counts <- dataset$count
    names(counts) <- dataset[["Gene"]]
    
    if (.ambig == "exc") {
      included_names <- names(counts) %in% gene_col_split[sapply(gene_col_split, length) == 1]
      chosen <- counts[included_names]
      chosen <- chosen[names(chosen) %in% .gene]
      other <- counts[names(chosen) %!in% .gene]
      res <- data.frame(
        Names = names(chosen),
        Count = unlist(chosen), stringsAsFactors = FALSE
      )
    } else if (.ambig == "maj") {
      maj_counts <- data.frame(Names = sapply(gene_col_split, "[[", 1), Counts = counts, stringsAsFactors = FALSE)
      row.names(maj_counts) <- NULL
      res <- maj_counts %>%
        group_by(Names) %>%
        summarise(Counts = sum(Counts))
    }
  }
  if (length(.gene %in% res[[1]]) < length(.gene)) {
    missing <- data.frame(.gene[!.gene %in% res[[1]]],
                          0,
                          stringsAsFactors = FALSE
    )
    names(missing) <- names(res)
    res <- rbind(res, missing)
  }
  row.names(res) <- NULL
  res <- res[order(res[[1]]), ]
  if (.norm) {
    res[, 2] <- res[, 2] / sum(res[, 2])
  }
  res <- as_tibble(as.data.frame(res))
  immunarch:::add_class(res, "immunr_gene_usage")
}



get_genes2 <- function(.gene = c("hs.trbv", "HomoSapiens.trbj", "macmul.ighv"), .type = c("segment", "allele", "family")) {
  .gene <- .gene[1]
  .type <- .type[1]
  
  # Species
  .gene <- tolower(.gene)
  which_species <- strsplit(.gene, ".", TRUE)[[1]][1]
  if (which_species %in% GENE_SEGMENTS2$alias) {
    which_species_col <- "alias"
  } else {
    if (which_species %in% tolower(GENE_SEGMENTS2$species)) {
      which_species_col <- "species"
    } else {
      stop("Unknown species name / alias")
    }
  }
  
  # Genes
  which_gene <- strsplit(.gene, ".", TRUE)[[1]][2]
  if (!(which_gene %in% GENE_SEGMENTS2[["gene"]][tolower(GENE_SEGMENTS2[[which_species_col]]) == which_species])) {
    stop("Unknown gene name")
  }
  
  # Id's type
  which_type <- paste0(.type, "_id")
  
  sort(unique(GENE_SEGMENTS2[[which_type]][tolower(GENE_SEGMENTS2[[which_species_col]]) == which_species & GENE_SEGMENTS2[["gene"]] == which_gene]))
}
```

```{r alltissues_geneusage}
igh_mixcr.f.gu <- igh_mixcr.f

# tidyverse solution - https://stackoverflow.com/questions/67837922/apply-gsub-over-a-certain-column-in-a-list-of-data-frames
library(tidyverse)
library(pheatmap)
library(RColorBrewer)

igh_mixcr.f.gu$data <- igh_mixcr.f.gu$data %>% 
  map(~ .x %>% mutate(V.name = gsub('\\(.{2,7}\\)', '', V.name),
                      D.name = gsub('\\(.{2,7}\\)', '', D.name),
                      J.name = gsub('\\(.{2,7}\\)', '', J.name)))

# constructed GENE_SEGMENTS entries for feline VDJ genes - formatted as follows
# alias	species	gene	family_id	segment_id	allele_id
# fc	FelisCatus	ighv	IGHV4	IGHV4-1	IGHV4-1*01
# fc	FelisCatus	ighv	IGHV3	IGHV3-2	IGHV3-2*01
# fc	FelisCatus	ighv	IGHV3	IGHV3-3	IGHV3-3*01
# fc	FelisCatus	ighv	IGHV3	IGHV3-4	IGHV3-4*01

# read in and rbind
feline.gene_segments <- read.csv(file = 'gene_segments_cat.txt', sep = '\t', header = T)
GENE_SEGMENTS2 <- rbind(GENE_SEGMENTS, feline.gene_segments)

# calculate and plot gene usage
# IGHV
igh_ighv_gu <- geneUsage2(igh_mixcr.f.gu$data, "fc.ighv", .quant = NA, .ambig = 'maj', .norm = T)
vis(igh_ighv_gu, .by = c("Group"), .plot = 'box', .meta = igh_mixcr.f.gu$meta)

igh_ighv_gu2 <- as.data.frame(igh_ighv_gu)
rownames(igh_ighv_gu2) <- igh_ighv_gu2$Names
igh_ighv_gu2$Names <- NULL
# replace NAs with 0
igh_ighv_gu2[is.na(igh_ighv_gu2)] <- 0

igh_ighv.breaks = seq(-3,3,0.001)
igh_ighv.cols = colorRampPalette(brewer.pal(9, 'RdYlBu'))(length(igh_ighv.breaks))
pheatmap(igh_ighv_gu2, 
         breaks = igh_ighv.breaks,
         color = igh_ighv.cols,
         cluster_rows = TRUE, 
         cluster_cols = FALSE,
         angle_col = 45,
         scale = 'row')

igh_ighv_gu_cor <- geneUsageAnalysis(igh_ighv_gu, .method = "cor", .verbose = F)
vis(igh_ighv_gu_cor, .title = "Gene usage correlation", .leg.title = "Cor", .text.size = 0)

# plot PCA of IGHV usage in all tissues
pca.n <- prcomp(t(igh_ighv_gu2), scale. = TRUE, center = TRUE)

scores.n  = as.data.frame(pca.n$x)
attach(scores.n)

var.capn = summary(pca.n)
pc_var.capn = round(var.capn$importance[2,]*100, 1)
cum_var.capn = round(var.capn$importance[3,]*100, 1)

library(ggplot2)
library(ggdendro)
library(reshape2)

pca_plot <- scores.n[,1:2]
pca_plot$sample <- rownames(pca_plot)
pca_plot$group <- ifelse(grepl('control', pca_plot$sample), 'Control', 'FIP')
pca_plot$group <- ifelse(grepl('142|192', pca_plot$sample), 'FIP (focal renal)', pca_plot$group)
pca_plot$tissue <- ifelse(grepl('liver', pca_plot$sample), 'Liver', 
                         ifelse(grepl('lung', pca_plot$sample), 'Lung', 
                                'MLN'))
pca_plot$cat <- paste(str_split_i(pca_plot$sample, '_', 1), str_split_i(pca_plot$sample, '_', 2), sep = '_')

library(ggrepel)

ggplot(pca_plot, aes(PC1, PC2, fill = group, shape = tissue)) +
  geom_point(size = 3) +
  geom_text_repel(aes(x = PC1, y = PC2, label = sample), max.overlaps = 20, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092', '#D071AD')) +
  scale_shape_manual(values = c(21, 23, 25)) +
  geom_hline(linetype = 3, yintercept = 0) +
  geom_vline(linetype = 3, xintercept = 0) +
  labs(x = paste('PC1: ', pc_var.capn[1], sep = '', '% of variance'), y = paste('PC2: ', pc_var.capn[2], sep = '', '% of variance'), fill = 'Group', shape = 'Tissue') +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  theme_bw() +
  theme(panel.grid = element_blank())

# significance testing
# test for difference between groups
# within group variance is radically different due to control_69
library(vegan)

dist <- vegdist(t(igh_ighv_gu2), method = 'bray')
plot(betadisper(dist, c(rep('Control', 21), rep('FIP', 21))))
boxplot(betadisper(dist, c(rep('Control', 21), rep('FIP', 21))))
anova(betadisper(dist, c(rep('Control', 21), rep('FIP', 21))))

# permanova
igh_ighv_gu4 <- as.data.frame(scale(as.data.frame(t(igh_ighv_gu2)), center = TRUE))
igh_ighv_gu4$group <- c(rep('Control', 21), rep('FIP', 21))
igh_ighv_gu5 <- as.data.frame(t(igh_ighv_gu2))
igh_ighv_gu5$group <- c(rep('Control', 21), rep('FIP', 21))

res <- adonis2(igh_ighv_gu5[,1:64] ~ group, data = igh_ighv_gu5, method = 'bray')

# boxplots for specific genes - 3-30, 3-60, 3-15, 3-5, 3-3
ighv.plot <- igh_ighv_gu2[rownames(igh_ighv_gu2) %in% c('IGHV3-30'),]

ighv.melt <- reshape2::melt(t(ighv.plot))
ighv.melt$Group1 <- ifelse(grepl('control', ighv.melt$Var1), 'Control', 'FIP')
ighv.melt$Group2 <- ifelse(grepl('142|192', ighv.melt$Var1), 'FIP (focal renal)', ighv.melt$Group1)
ighv.melt$tissue <- ifelse(grepl('liver', ighv.melt$Var1), 'Liver', 
                         ifelse(grepl('lung', ighv.melt$Var1), 'Lung', 'MLN'))
# ighv.melt$value <- sqrt(ighv.melt$value)

ggplot(ighv.melt, aes(Group1, value)) +
  geom_boxplot(aes(fill = Group1), outlier.shape = NA, alpha = 0.5, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  geom_jitter(aes(colour = Group2), width = 0.1) +
  scale_color_manual(values = c('#32A7CD', '#8F3092', '#D071AD')) +
  labs(x = '', y = 'Proportion of clonotypes', fill = '', colour = '') +
  facet_wrap(vars(tissue), nrow = 1) +
  stat_compare_means() +
  ggtitle('IGHV3-30') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5))
```

Plot isotype usage for all samples.

```{r alltissues_isotype}
library(ggplot2)
library(ggpubr)
library(ggsci)

# combine isotype usage data from all samples
isotype.multitissue <- list(isotype.mln.all, isotype.liver.all, isotype.lung.all)
names(isotype.multitissue) <- c('MLN', 'Liver', 'Lung')

# bind rows
isotype.multitissue.df <- bind_rows(isotype.multitissue, .id = 'Tissue')

# add group
isotype.multitissue.df$Group <- ifelse(grepl('control', isotype.multitissue.df$Sample), 'Control', 'FIP')

# plot props faceted by tissue and isotype
ggplot(isotype.multitissue.df, aes(Group, Proportion, fill = Group)) +
  geom_boxplot(outlier.shape = NA, show.legend = TRUE, alpha = 0.5) +
  geom_jitter(aes(color = Group), width = 0.1, show.legend = FALSE) +
  labs(x = '', y = 'Proportion of clonotypes') +
  facet_grid(rows = vars(C.name), cols = vars(Tissue), scales = 'free_y') +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  theme_bw() +
  theme(panel.grid = element_blank())

# individual plots and wilcoxon rank sum tests
# IGHG
ighg.multitissue <- lapply(isotype.multitissue, function(x) x[x$C.name == 'IGHG',])

# add group
ighg.multitissue <- lapply(ighg.multitissue, function(x) {
  x$Group = ifelse(grepl('control', x$Sample), 'Control', 'FIP')
  x
})

# bind rows and plot
ighg.flat <- bind_rows(ighg.multitissue, .id = 'Tissue')
ighg.flat$Group2 <- ifelse(grepl('142|192', ighg.flat$Sample), 'FIP (focal renal)', ighg.flat$Group)

ggplot(ighg.flat, aes(Group, Proportion, fill = Group)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE, alpha = 0.5) +
  stat_compare_means(label.y = 0.9, size = 3) +
  scale_y_continuous(expand = c(0,0), limits = c(0,1), breaks = seq(0,1,0.2)) +
  geom_jitter(aes(color = Group2), width = 0.1, show.legend = TRUE) +
  labs(x = '', y = 'Proportion of clonotypes', title = 'IGHG') +
  facet_wrap(vars(Tissue), nrow = 1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092', '#D071AD')) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5))
  
# get p values
lapply(ighg.multitissue, function(x) pairwise.wilcox.test(x$Proportion, x$Group, paired = FALSE, p.adjust.method = 'none'))

# IGHM
ighm.multitissue <- lapply(isotype.multitissue, function(x) x[x$C.name == 'IGHM',])

# add group
ighm.multitissue <- lapply(ighm.multitissue, function(x) {
  x$Group = ifelse(grepl('control', x$Sample), 'Control', 'FIP')
  x
})

# bind rows and plot
ighm.flat <- bind_rows(ighm.multitissue, .id = 'Tissue')

ggplot(ighm.flat, aes(Group, Proportion, fill = Group)) +
  geom_boxplot(outlier.shape = NA, show.legend = TRUE, alpha = 0.5) +
  stat_compare_means(label.y = 1.1, size = 3, method = 't.test') +
  scale_y_continuous(limits = c(0,1.2), breaks = seq(0,1,0.2)) +
  geom_jitter(aes(color = Group), width = 0.1, show.legend = FALSE) +
  labs(x = '', y = 'Fraction of clonotypes') +
  facet_wrap(vars(Tissue), nrow = 1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  theme_bw() +
  theme(panel.grid = element_blank())

# get p values
lapply(ighm.multitissue, function(x) pairwise.wilcox.test(x$Proportion, x$Group, paired = FALSE, p.adjust.method = 'none'))

# IGHA
igha.multitissue <- lapply(isotype.multitissue, function(x) x[x$C.name == 'IGHA',])

# add group
igha.multitissue <- lapply(igha.multitissue, function(x) {
  x$Group = ifelse(grepl('control', x$Sample), 'Control', 'FIP')
  x
})

# bind rows and plot
igha.flat <- bind_rows(igha.multitissue, .id = 'Tissue')

ggplot(igha.flat, aes(Group, Proportion, fill = Group)) +
  geom_boxplot(outlier.shape = NA, show.legend = TRUE, alpha = 0.5) +
  stat_compare_means(label.y = 1.1, size = 3, method = 't.test') +
  scale_y_continuous(limits = c(0,1.2), breaks = seq(0,1,0.2)) +
  geom_jitter(aes(color = Group), width = 0.1, show.legend = FALSE) +
  labs(x = '', y = 'Fraction of clonotypes') +
  facet_wrap(vars(Tissue), nrow = 1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  theme_bw() +
  theme(panel.grid = element_blank())

# get p values
lapply(igha.multitissue, function(x) pairwise.wilcox.test(x$Proportion, x$Group, paired = FALSE, p.adjust.method = 'none'))

# create faceted plot (stacked bars) for all tissues
ggplot(isotype.multitissue.df, aes(Sample, Proportion, fill = C.name)) +
  geom_bar(stat = 'identity', position = 'stack') +
  facet_grid(cols = vars(Tissue), scales = 'free_x', space = 'free') +
  scale_fill_d3() +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = '', y = 'Fraction of clonotypes', fill = '') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# reformat plot as mean
isotype.multissue.list <- split(isotype.multitissue.df, isotype.multitissue.df$Tissue)
isotype.multissue.list <- isotype.multissue.list[-1]

isotype.multissue.list <- lapply(isotype.multissue.list, function(x) {
  x %>%
    group_by(Group, C.name) %>%
    summarise(avg.prop = mean(Proportion), sd.prop = plotrix::std.error(Proportion)) %>%
    arrange(rev(C.name), by.group = TRUE)
})

isotype.multissue.df2 <- bind_rows(isotype.multissue.list, .id = 'Tissue')

ggplot(isotype.multissue.df2, aes(Group, avg.prop, fill = C.name)) +
  geom_bar(stat = 'identity', position = 'stack') +
  scale_y_continuous(expand = c(0,0)) +
  scale_fill_d3() +
  labs(x = '', y = 'Proportion of clonotypes', fill = '') +
  facet_wrap(vars(Tissue), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Network analysis (NAIR)

https://mlizhangx.github.io/Network-Analysis-for-Repertoire-Sequencing-/

```{r alltissues_nair_clonesampling}
# create a list of tissue repertoires for each cat
all.f.list <- list(control_135 = list(control_135_liver = igh_mixcr.f$data$control_135_liver, 
                                      control_135_lung = igh_mixcr.f$data$control_135_lung, 
                                      control_135_mln = igh_mixcr.f$data$control_135_mln),
                   control_136 = list(control_136_liver = igh_mixcr.f$data$control_136_liver, 
                                      control_136_lung = igh_mixcr.f$data$control_136_lung, 
                                      control_136_mln = igh_mixcr.f$data$control_136_mln),
                   control_164 = list(control_164_liver = igh_mixcr.f$data$control_164_liver, 
                                      control_164_lung = igh_mixcr.f$data$control_164_lung, 
                                      control_164_mln = igh_mixcr.f$data$control_164_mln),
                   control_174 = list(control_174_liver = igh_mixcr.f$data$control_174_liver, 
                                      control_174_lung = igh_mixcr.f$data$control_174_lung, 
                                      control_174_mln = igh_mixcr.f$data$control_174_mln),
                   control_188 = list(control_188_liver = igh_mixcr.f$data$control_188_liver, 
                                      control_188_lung = igh_mixcr.f$data$control_188_lung, 
                                      control_188_mln = igh_mixcr.f$data$control_188_mln),
                   control_54 = list(control_54_liver = igh_mixcr.f$data$control_54_liver, 
                                     control_54_lung = igh_mixcr.f$data$control_54_lung, 
                                     control_54_mln = igh_mixcr.f$data$control_54_mln),
                   control_69 = list(control_69_liver = igh_mixcr.f$data$control_69_liver, 
                                     control_69_lung = igh_mixcr.f$data$control_69_lung, 
                                     control_69_mln = igh_mixcr.f$data$control_69_mln),
                   fip_120 = list(fip_120_liver = igh_mixcr.f$data$fip_120_liver, 
                                  fip_120_mln = igh_mixcr.f$data$fip_120_mln),
                   fip_142 = list(fip_142_liver = igh_mixcr.f$data$fip_142_liver, 
                                  fip_142_lung = igh_mixcr.f$data$fip_142_lung, 
                                  fip_142_mln = igh_mixcr.f$data$fip_142_mln),
                   fip_172 = list(fip_172_liver = igh_mixcr.f$data$fip_172_liver, 
                                  fip_172_lung = igh_mixcr.f$data$fip_172_lung, 
                                  fip_172_mln = igh_mixcr.f$data$fip_172_mln),
                   fip_192 = list(fip_192_liver = igh_mixcr.f$data$fip_192_liver, 
                                  fip_192_lung = igh_mixcr.f$data$fip_192_lung, 
                                  fip_192_mln = igh_mixcr.f$data$fip_192_mln),
                   fip_37 = list(fip_37_liver = igh_mixcr.f$data$fip_37_liver, 
                                 fip_37_mln = igh_mixcr.f$data$fip_37_mln),
                   fip_68 = list(fip_68_liver = igh_mixcr.f$data$fip_68_liver, 
                                 fip_68_lung = igh_mixcr.f$data$fip_68_lung, 
                                 fip_68_mln = igh_mixcr.f$data$fip_68_mln),
                   fip_86 = list(fip_86_liver = igh_mixcr.f$data$fip_86_liver, 
                                 fip_86_lung = igh_mixcr.f$data$fip_86_lung, 
                                 fip_86_mln = igh_mixcr.f$data$fip_86_mln),
                   fip_96 = list(fip_96_liver = igh_mixcr.f$data$fip_96_liver, 
                                 fip_96_mln = igh_mixcr.f$data$fip_96_mln))

# collapse lists to single df for each cat
all.f.df <- lapply(all.f.list, function(x) bind_rows(x, .id = 'Sample'))

# add tissue
all.f.df <- lapply(all.f.df, function(x) {
  x$Tissue <- ifelse(grepl('liver', x$Sample), 'Liver',
                              ifelse(grepl('lung', x$Sample), 'Lung', 'MLN'))
  x
})

# build networks
all.f.net <- lapply(all.f.df, function(x) {
  buildNet(data = x,
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

# subsample clusters in each network 1000 times
all.f.clustsamples <- lapply(all.f.net, function(x) {
  replicate(1000, sample(x$cluster_data$cluster_id, 2000, replace = FALSE), simplify = FALSE)
})

# for each sample, extract nodes in sampled clusters
control_135.samples <- lapply(all.f.clustsamples$control_135, function(x) {
  buildNet(data = subset(all.f.net$control_135$node_data[all.f.net$control_135$node_data$cluster_id %in% x,], 
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

control_136.samples <- lapply(all.f.clustsamples$control_136, function(x) {
  buildNet(data = subset(all.f.net$control_136$node_data[all.f.net$control_136$node_data$cluster_id %in% x,], 
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

control_164.samples <- lapply(all.f.clustsamples$control_164, function(x) {
  buildNet(data = subset(all.f.net$control_164$node_data[all.f.net$control_164$node_data$cluster_id %in% x,], 
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

control_174.samples <- lapply(all.f.clustsamples$control_174, function(x) {
  buildNet(data = subset(all.f.net$control_174$node_data[all.f.net$control_174$node_data$cluster_id %in% x,], 
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

control_188.samples <- lapply(all.f.clustsamples$control_188, function(x) {
  buildNet(data = subset(all.f.net$control_188$node_data[all.f.net$control_188$node_data$cluster_id %in% x,], 
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

control_54.samples <- lapply(all.f.clustsamples$control_54, function(x) {
  buildNet(data = subset(all.f.net$control_54$node_data[all.f.net$control_54$node_data$cluster_id %in% x,], 
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

control_69.samples <- lapply(all.f.clustsamples$control_69, function(x) {
  buildNet(data = subset(all.f.net$control_69$node_data[all.f.net$control_69$node_data$cluster_id %in% x,], 
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

fip_142.samples <- lapply(all.f.clustsamples$fip_142, function(x) {
  buildNet(data = subset(all.f.net$fip_142$node_data[all.f.net$fip_142$node_data$cluster_id %in% x,],
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

fip_172.samples <- lapply(all.f.clustsamples$fip_172, function(x) {
  buildNet(data = subset(all.f.net$fip_172$node_data[all.f.net$fip_172$node_data$cluster_id %in% x,],
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

fip_192.samples <- lapply(all.f.clustsamples$fip_192, function(x) {
  buildNet(data = subset(all.f.net$fip_192$node_data[all.f.net$fip_192$node_data$cluster_id %in% x,],
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

fip_37.samples <- lapply(all.f.clustsamples$fip_37, function(x) {
  buildNet(data = subset(all.f.net$fip_37$node_data[all.f.net$fip_37$node_data$cluster_id %in% x,],
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

fip_68.samples <- lapply(all.f.clustsamples$fip_68, function(x) {
  buildNet(data = subset(all.f.net$fip_68$node_data[all.f.net$fip_68$node_data$cluster_id %in% x,],
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

fip_86.samples <- lapply(all.f.clustsamples$fip_86, function(x) {
  buildNet(data = subset(all.f.net$fip_86$node_data[all.f.net$fip_86$node_data$cluster_id %in% x,],
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

fip_96.samples <- lapply(all.f.clustsamples$fip_96, function(x) {
  buildNet(data = subset(all.f.net$fip_96$node_data[all.f.net$fip_96$node_data$cluster_id %in% x,],
                         select = c(1:5, Tissue)),
           seq_col = 'CDR3.nt',
           count_col = 'Proportion',
           color_nodes_by = 'Tissue',
           size_nodes_by = 'Proportion',
           drop_isolated_nodes = FALSE,
           node_stats = TRUE,
           cluster_stats = TRUE)
})

all.samples <- list(fip_142 = fip_142.samples,
                    fip_172 = fip_172.samples,
                    fip_192 = fip_192.samples,
                    fip_120 = fip_120.samples,
                    fip_37 = fip_37.samples,
                    fip_68 = fip_68.samples,
                    fip_86 = fip_86.samples,
                    fip_96 = fip_96.samples,
                    control_135 = control_135.samples,
                    control_136 = control_136.samples,
                    control_164 = control_164.samples,
                    control_174 = control_174.samples,
                    control_188 = control_188.samples,
                    control_54 = control_54.samples,
                    control_69 = control_69.samples)

# for each sample, get the median largest cluster size of all samples
cluster.sizes <- as.data.frame(t(data.frame(lapply(all.samples, function(x) median(sapply(x, function(y) max(y$cluster_data$node_count)))))))

# add group info
cluster.sizes$group <- ifelse(grepl('control', rownames(cluster.sizes)), 'Control', 'FIP')
cluster.sizes$group2 <- ifelse(grepl('142|192', rownames(cluster.sizes)), 'FIP (focal renal)', cluster.sizes$group)

# plot
library(ggplot2)

ggplot(cluster.sizes, aes(group, V1, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5, show.legend = FALSE) +
  geom_jitter(aes(color = group2), width = 0.1, show.legend = TRUE) +
  scale_y_continuous(limits = c(0,80)) +
  scale_color_manual(values = c('#32A7CD', '#8F3092', '#D071AD')) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  labs(x = '', y = 'Median largest cluster size') +
  theme_bw() +
  theme(panel.grid = element_blank())

wilcox.test(cluster.sizes[cluster.sizes$group == 'FIP',]$V1, cluster.sizes[cluster.sizes$group == 'Control',]$V1)

# get the number of multi-tissue clusters (>1 node) in each sample
# adjust this to only MLN and Liver, as these tissues are included for all samples
# cast a matrix of clone counts per tissue, by cluster_id
all.samples.multimat <- suppressMessages(lapply(all.samples, function(x) lapply(x, function(y) as.data.frame(reshape2::acast(y$node_data, cluster_id ~ Tissue, fill = 0)))))

# remove lung
all.samples.multimat <- lapply(all.samples.multimat, function(x) lapply(x, function(y) subset(y, select = c(MLN, Liver))))

# add variable for multi-tissue clusters
all.samples.multimat <- lapply(all.samples.multimat, function(x) 
  lapply(x, function(y) {
    y$multitissue <- rowSums(y > 0) > 1
    y
}))

# get mean number of multi-tissue clusters per sample
mean.multi <- lapply(all.samples.multimat, function(x) mean(unlist(lapply(x, function(y) nrow(y[y$multitissue == TRUE,])))))

# unlist and plot
mean.multi.df <- data.frame(meanmulti = unlist(mean.multi), sample = names(mean.multi))
mean.multi.df$group <- ifelse(grepl('control', rownames(mean.multi.df)), 'Control', 'FIP')
mean.multi.df$group2 <- ifelse(grepl('142|192', rownames(mean.multi.df)), 'FIP (focal renal)', mean.multi.df$group)

ggplot(mean.multi.df, aes(group, meanmulti, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5, show.legend = FALSE) +
  geom_jitter(aes(color = group2), width = 0.1, show.legend = TRUE) +
  # scale_y_continuous(limits = c(0,80)) +
  scale_color_manual(values = c('#32A7CD', '#8F3092', '#D071AD')) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  labs(x = '', y = 'Mean cross-tissue clusters (MLN - Liver)') +
  theme_bw() +
  theme(panel.grid = element_blank())

wilcox.test(mean.multi.df[mean.multi.df$group == 'FIP',]$meanmulti, mean.multi.df[mean.multi.df$group == 'Control',]$meanmulti)
```

Plot representative networks.

```{r alltissues_nair_repnetworks}
library(NAIR)
library(igraph)
library(ggraph)
library(netdiffuseR)
library(qgraph)

# select representative clusters for plotting
# control_135 - median 18
control_135.clustsizes <- data.frame(max.cluster = unlist(lapply(control_135.samples, function(x) max(x$cluster_data$node_count))))
control_135.clustsizes$sample <- row.names(control_135.clustsizes)

# select rows with a max cluster size of 18
control_135.clustsizes.median <- control_135.clustsizes[control_135.clustsizes$max.cluster == 18,]

# randomly select a representative sample
sample(control_135.clustsizes.median$sample, 1)

# plot sample 202
control_135.net <- control_135.samples[[202]]$igraph
V(control_135.net)$size <- control_135.samples[[202]]$node_data$Proportion
V(control_135.net)$tissue <- control_135.samples[[202]]$node_data$Tissue

layout135 = qgraph.layout.fruchtermanreingold(as_edgelist(control_135.net ,names = FALSE), 
                                              vcount = vcount(control_135.net), 
                                              area = 10 * vcount(control_135.net)^2)

ggraph(control_135.net, layout = layout135) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(control_135.net)$tissue), 
                    size = rescale_vertex_igraph(V(control_135.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# control_136 - median 16
control_136.clustsizes <- data.frame(max.cluster = unlist(lapply(control_136.samples, function(x) max(x$cluster_data$node_count))))
control_136.clustsizes$sample <- row.names(control_136.clustsizes)

# select rows with a max cluster size of 18
control_136.clustsizes.median <- control_136.clustsizes[control_136.clustsizes$max.cluster == 16,]

# randomly select a representative sample
sample(control_136.clustsizes.median$sample, 1)

# plot sample 507
control_136.net <- control_136.samples[[507]]$igraph
V(control_136.net)$size <- control_136.samples[[507]]$node_data$Proportion
V(control_136.net)$tissue <- control_136.samples[[507]]$node_data$Tissue

layout136 = qgraph.layout.fruchtermanreingold(as_edgelist(control_136.net ,names = FALSE), 
                                              vcount = vcount(control_136.net), 
                                              area = 10 * vcount(control_136.net)^2)

ggraph(control_136.net, layout = layout136) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(control_136.net)$tissue), 
                    size = rescale_vertex_igraph(V(control_136.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# control_164 - median 73
control_164.clustsizes <- data.frame(max.cluster = unlist(lapply(control_164.samples, function(x) max(x$cluster_data$node_count))))
control_164.clustsizes$sample <- row.names(control_164.clustsizes)

# select rows with a max cluster size of 73
control_164.clustsizes.median <- control_164.clustsizes[control_164.clustsizes$max.cluster == 73,]

# randomly select a representative sample
sample(control_164.clustsizes.median$sample, 1)

# plot sample 612
control_164.net <- control_164.samples[[612]]$igraph
V(control_164.net)$size <- control_164.samples[[612]]$node_data$Proportion
V(control_164.net)$tissue <- control_164.samples[[612]]$node_data$Tissue

layout164 = qgraph.layout.fruchtermanreingold(as_edgelist(control_164.net ,names = FALSE), 
                                              vcount = vcount(control_164.net), 
                                              area = 10 * vcount(control_164.net)^2)

ggraph(control_164.net, layout = layout164) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(control_164.net)$tissue), 
                    size = rescale_vertex_igraph(V(control_164.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# control_174 - median 10
control_174.clustsizes <- data.frame(max.cluster = unlist(lapply(control_174.samples, function(x) max(x$cluster_data$node_count))))
control_174.clustsizes$sample <- row.names(control_174.clustsizes)

# select rows with a max cluster size of 73
control_174.clustsizes.median <- control_174.clustsizes[control_174.clustsizes$max.cluster == 10,]

# randomly select a representative sample
sample(control_174.clustsizes.median$sample, 1)

# plot sample 617
control_174.net <- control_174.samples[[617]]$igraph
V(control_174.net)$size <- control_174.samples[[617]]$node_data$Proportion
V(control_174.net)$tissue <- control_174.samples[[617]]$node_data$Tissue

layout174 = qgraph.layout.fruchtermanreingold(as_edgelist(control_174.net ,names = FALSE), 
                                              vcount = vcount(control_174.net), 
                                              area = 10 * vcount(control_174.net)^2)

ggraph(control_174.net, layout = layout174) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(control_174.net)$tissue), 
                    size = rescale_vertex_igraph(V(control_174.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# control_188 - median 18
control_188.clustsizes <- data.frame(max.cluster = unlist(lapply(control_188.samples, function(x) max(x$cluster_data$node_count))))
control_188.clustsizes$sample <- row.names(control_188.clustsizes)

# select rows with a max cluster size of 73
control_188.clustsizes.median <- control_188.clustsizes[control_188.clustsizes$max.cluster == 18,]

# randomly select a representative sample
sample(control_188.clustsizes.median$sample, 1)

# plot sample 70
control_188.net <- control_188.samples[[70]]$igraph
V(control_188.net)$size <- control_188.samples[[70]]$node_data$Proportion
V(control_188.net)$tissue <- control_188.samples[[70]]$node_data$Tissue

layout188 = qgraph.layout.fruchtermanreingold(as_edgelist(control_188.net ,names = FALSE), 
                                              vcount = vcount(control_188.net), 
                                              area = 10 * vcount(control_188.net)^2)

ggraph(control_188.net, layout = layout188) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(control_188.net)$tissue), 
                    size = rescale_vertex_igraph(V(control_188.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# control_54 - median 13
control_54.clustsizes <- data.frame(max.cluster = unlist(lapply(control_54.samples, function(x) max(x$cluster_data$node_count))))
control_54.clustsizes$sample <- row.names(control_54.clustsizes)

# select rows with a max cluster size of 73
control_54.clustsizes.median <- control_54.clustsizes[control_54.clustsizes$max.cluster == 13,]

# randomly select a representative sample
sample(control_54.clustsizes.median$sample, 1)

# plot sample 676
control_54.net <- control_54.samples[[676]]$igraph
V(control_54.net)$size <- control_54.samples[[676]]$node_data$Proportion
V(control_54.net)$tissue <- control_54.samples[[676]]$node_data$Tissue

layout54 = qgraph.layout.fruchtermanreingold(as_edgelist(control_54.net ,names = FALSE), 
                                              vcount = vcount(control_54.net), 
                                              area = 10 * vcount(control_54.net)^2)

ggraph(control_54.net, layout = layout54) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(control_54.net)$tissue), 
                    size = rescale_vertex_igraph(V(control_54.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# control_69 - median 17
control_69.clustsizes <- data.frame(max.cluster = unlist(lapply(control_69.samples, function(x) max(x$cluster_data$node_count))))
control_69.clustsizes$sample <- row.names(control_69.clustsizes)

# select rows with a max cluster size of 73
control_69.clustsizes.median <- control_69.clustsizes[control_69.clustsizes$max.cluster == 17,]

# randomly select a representative sample
sample(control_69.clustsizes.median$sample, 1)

# plot sample 1000
control_69.net <- control_69.samples[[1000]]$igraph
V(control_69.net)$size <- control_69.samples[[1000]]$node_data$Proportion
V(control_69.net)$tissue <- control_69.samples[[1000]]$node_data$Tissue

layout69 = qgraph.layout.fruchtermanreingold(as_edgelist(control_69.net ,names = FALSE), 
                                              vcount = vcount(control_69.net), 
                                              area = 10 * vcount(control_69.net)^2)

ggraph(control_69.net, layout = layout69) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(control_69.net)$tissue), 
                    size = rescale_vertex_igraph(V(control_69.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# fip_142 - median 15
fip_142.clustsizes <- data.frame(max.cluster = unlist(lapply(fip_142.samples, function(x) max(x$cluster_data$node_count))))
fip_142.clustsizes$sample <- row.names(fip_142.clustsizes)

# select rows with a max cluster size of 73
fip_142.clustsizes.median <- fip_142.clustsizes[fip_142.clustsizes$max.cluster == 15,]

# randomly select a representative sample
sample(fip_142.clustsizes.median$sample, 1)

# plot sample 322
fip_142.net <- fip_142.samples[[322]]$igraph
V(fip_142.net)$size <- fip_142.samples[[322]]$node_data$Proportion
V(fip_142.net)$tissue <- fip_142.samples[[322]]$node_data$Tissue

layout142 = qgraph.layout.fruchtermanreingold(as_edgelist(fip_142.net ,names = FALSE), 
                                              vcount = vcount(fip_142.net), 
                                              area = 10 * vcount(fip_142.net)^2)

ggraph(fip_142.net, layout = layout142) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(fip_142.net)$tissue), 
                    size = rescale_vertex_igraph(V(fip_142.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# fip_172 - median 34
fip_172.clustsizes <- data.frame(max.cluster = unlist(lapply(fip_172.samples, function(x) max(x$cluster_data$node_count))))
fip_172.clustsizes$sample <- row.names(fip_172.clustsizes)

# select rows with a max cluster size of 73
fip_172.clustsizes.median <- fip_172.clustsizes[fip_172.clustsizes$max.cluster == 34,]

# randomly select a representative sample
sample(fip_172.clustsizes.median$sample, 1)

# plot sample 957
fip_172.net <- fip_172.samples[[957]]$igraph
V(fip_172.net)$size <- fip_172.samples[[957]]$node_data$Proportion
V(fip_172.net)$tissue <- fip_172.samples[[957]]$node_data$Tissue

layout172 = qgraph.layout.fruchtermanreingold(as_edgelist(fip_172.net ,names = FALSE), 
                                              vcount = vcount(fip_172.net), 
                                              area = 10 * vcount(fip_172.net)^2)

ggraph(fip_172.net, layout = layout172) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(fip_172.net)$tissue), 
                    size = rescale_vertex_igraph(V(fip_172.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# fip_192 - median 34
fip_192.clustsizes <- data.frame(max.cluster = unlist(lapply(fip_192.samples, function(x) max(x$cluster_data$node_count))))
fip_192.clustsizes$sample <- row.names(fip_192.clustsizes)

# select rows with a max cluster size of 73
fip_192.clustsizes.median <- fip_192.clustsizes[fip_192.clustsizes$max.cluster == 34,]

# randomly select a representative sample
sample(fip_192.clustsizes.median$sample, 1)

# plot sample 368
fip_192.net <- fip_192.samples[[368]]$igraph
V(fip_192.net)$size <- fip_192.samples[[368]]$node_data$Proportion
V(fip_192.net)$tissue <- fip_192.samples[[368]]$node_data$Tissue

layout192 = qgraph.layout.fruchtermanreingold(as_edgelist(fip_192.net ,names = FALSE), 
                                              vcount = vcount(fip_192.net), 
                                              area = 10 * vcount(fip_192.net)^2)

ggraph(fip_192.net, layout = layout192) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(fip_192.net)$tissue), 
                    size = rescale_vertex_igraph(V(fip_192.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# fip_120 - median 34
fip_120.clustsizes <- data.frame(max.cluster = unlist(lapply(fip_120.samples, function(x) max(x$cluster_data$node_count))))
fip_120.clustsizes$sample <- row.names(fip_120.clustsizes)

# select rows with a max cluster size of 73
fip_120.clustsizes.median <- fip_120.clustsizes[fip_120.clustsizes$max.cluster == 34,]

# randomly select a representative sample
sample(fip_120.clustsizes.median$sample, 1)

# plot sample 111
fip_120.net <- fip_120.samples[[111]]$igraph
V(fip_120.net)$size <- fip_120.samples[[111]]$node_data$Proportion
V(fip_120.net)$tissue <- fip_120.samples[[111]]$node_data$Tissue

layout120 = qgraph.layout.fruchtermanreingold(as_edgelist(fip_120.net ,names = FALSE), 
                                              vcount = vcount(fip_120.net), 
                                              area = 10 * vcount(fip_120.net)^2)

ggraph(fip_120.net, layout = layout120) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(fip_120.net)$tissue), 
                    size = rescale_vertex_igraph(V(fip_120.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#033BF4')) +
    labs(color = '') +
    theme_void()

# fip_37 - median 36
fip_37.clustsizes <- data.frame(max.cluster = unlist(lapply(fip_37.samples, function(x) max(x$cluster_data$node_count))))
fip_37.clustsizes$sample <- row.names(fip_37.clustsizes)

# select rows with a max cluster size of 73
fip_37.clustsizes.median <- fip_37.clustsizes[fip_37.clustsizes$max.cluster == 36,]

# randomly select a representative sample
sample(fip_37.clustsizes.median$sample, 1)

# plot sample 738
fip_37.net <- fip_37.samples[[738]]$igraph
V(fip_37.net)$size <- fip_37.samples[[738]]$node_data$Proportion
V(fip_37.net)$tissue <- fip_37.samples[[738]]$node_data$Tissue

layout37 = qgraph.layout.fruchtermanreingold(as_edgelist(fip_37.net ,names = FALSE), 
                                              vcount = vcount(fip_37.net), 
                                              area = 10 * vcount(fip_37.net)^2)

ggraph(fip_37.net, layout = layout37) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(fip_37.net)$tissue), 
                    size = rescale_vertex_igraph(V(fip_37.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#033BF4')) +
    labs(color = '') +
    theme_void()

# fip_68 - median 57
fip_68.clustsizes <- data.frame(max.cluster = unlist(lapply(fip_68.samples, function(x) max(x$cluster_data$node_count))))
fip_68.clustsizes$sample <- row.names(fip_68.clustsizes)

# select rows with a max cluster size of 73
fip_68.clustsizes.median <- fip_68.clustsizes[fip_68.clustsizes$max.cluster == 57,]

# randomly select a representative sample
sample(fip_68.clustsizes.median$sample, 1)

# plot sample 832
fip_68.net <- fip_68.samples[[832]]$igraph
V(fip_68.net)$size <- fip_68.samples[[832]]$node_data$Proportion
V(fip_68.net)$tissue <- fip_68.samples[[832]]$node_data$Tissue

layout68 = qgraph.layout.fruchtermanreingold(as_edgelist(fip_68.net ,names = FALSE), 
                                              vcount = vcount(fip_68.net), 
                                              area = 10 * vcount(fip_68.net)^2)

ggraph(fip_68.net, layout = layout68) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(fip_68.net)$tissue), 
                    size = rescale_vertex_igraph(V(fip_68.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# fip_86 - median 65
fip_86.clustsizes <- data.frame(max.cluster = unlist(lapply(fip_86.samples, function(x) max(x$cluster_data$node_count))))
fip_86.clustsizes$sample <- row.names(fip_86.clustsizes)

# select rows with a max cluster size of 73
fip_86.clustsizes.median <- fip_86.clustsizes[fip_86.clustsizes$max.cluster == 65,]

# randomly select a representative sample
sample(fip_86.clustsizes.median$sample, 1)

# plot sample 840
fip_86.net <- fip_86.samples[[840]]$igraph
V(fip_86.net)$size <- fip_86.samples[[840]]$node_data$Proportion
V(fip_86.net)$tissue <- fip_86.samples[[840]]$node_data$Tissue

layout86 = qgraph.layout.fruchtermanreingold(as_edgelist(fip_86.net ,names = FALSE), 
                                              vcount = vcount(fip_86.net), 
                                              area = 10 * vcount(fip_86.net)^2)

ggraph(fip_86.net, layout = layout86) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(fip_86.net)$tissue), 
                    size = rescale_vertex_igraph(V(fip_86.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
    labs(color = '') +
    theme_void()

# fip_96 - median 26
fip_96.clustsizes <- data.frame(max.cluster = unlist(lapply(fip_96.samples, function(x) max(x$cluster_data$node_count))))
fip_96.clustsizes$sample <- row.names(fip_96.clustsizes)

# select rows with a max cluster size of 73
fip_96.clustsizes.median <- fip_96.clustsizes[fip_96.clustsizes$max.cluster == 26,]

# randomly select a representative sample
sample(fip_96.clustsizes.median$sample, 1)

# plot sample 138
fip_96.net <- fip_96.samples[[138]]$igraph
V(fip_96.net)$size <- fip_96.samples[[138]]$node_data$Proportion
V(fip_96.net)$tissue <- fip_96.samples[[138]]$node_data$Tissue

layout96 = qgraph.layout.fruchtermanreingold(as_edgelist(fip_96.net ,names = FALSE), 
                                              vcount = vcount(fip_96.net), 
                                              area = 10 * vcount(fip_96.net)^2)

ggraph(fip_96.net, layout = layout96) +
    geom_edge_link(color = "black", width = 0.3) +
    geom_node_point(aes(color = V(fip_96.net)$tissue), 
                    size = rescale_vertex_igraph(V(fip_96.net)$size, minmax.relative.size = c(0.02, 0.08), adjust = 100)) +
    scale_color_manual(values = c('#EA8131', '#033BF4')) +
    labs(color = '') +
    theme_void()
```

Construct a global network comprising all clones from all samples, and determine cluster membership from this network.

```{r nair_global_cluster_alt}
# create df for all clones
igh_mixcr.f.df <- bind_rows(igh_mixcr.f$data, .id = 'Sample')

# add tissue
igh_mixcr.f.df$Tissue <- ifelse(grepl('liver', igh_mixcr.f.df$Sample), 'Liver',
                              ifelse(grepl('lung', igh_mixcr.f.df$Sample), 'Lung', 'MLN'))

# build network
igh_mixcr.f.net <- buildNet(data = igh_mixcr.f.df,
                       seq_col = 'CDR3.aa',
                       count_col = 'Proportion',
                       color_nodes_by = 'Tissue',
                       size_nodes_by = 'Proportion',
                       drop_isolated_nodes = TRUE,
                       node_stats = TRUE,
                       cluster_stats = TRUE)

# split node_data by sample and determine overlap (by cluster ID)
split.list <- split(igh_mixcr.f.net$node_data, igh_mixcr.f.net$node_data$Sample)

split.clust <- lapply(split.list, function(x) {
  x <- x %>%
    select(cluster_id) %>%
    unique()
})

sample.order <- c("control_54_liver", "control_54_lung", "control_54_mln", "control_69_liver", "control_69_lung", "control_69_mln",
                  "control_135_liver", "control_135_lung", "control_135_mln", "control_136_liver", "control_136_lung", "control_136_mln",
                  "control_174_liver", "control_174_lung", "control_174_mln", "control_188_liver", "control_188_lung", "control_188_mln",
                  "control_164_liver", "control_164_lung", "control_164_mln", "fip_142_liver", "fip_142_lung", "fip_142_mln",
                  "fip_192_liver", "fip_192_lung", "fip_192_mln", "fip_37_liver", "fip_37_mln", "fip_96_liver", "fip_96_mln",
                  "fip_120_liver", "fip_120_mln", "fip_172_liver", "fip_172_lung", "fip_172_mln",
                  "fip_68_liver", "fip_68_lung", "fip_68_mln", "fip_86_liver", "fip_86_lung", "fip_86_mln")

split.ovl <- apply_symm(split.clust[sample.order], function(.x, .y) {
  .x <- collect(.x, n = Inf)
  .y <- collect(.y, n = Inf)
  intersection <- nrow(dplyr::intersect(.x, .y))
  intersection / (nrow(.x) + nrow(.y) - intersection)
})

# options for measure of overlap in function above, jaccard preferred
# # ovlap coef
# nrow(dplyr::intersect(.x, .y)) / min(nrow(.x), nrow(.y))
# # jaccard
# intersection <- nrow(dplyr::intersect(.x, .y))
# intersection / (nrow(.x) + nrow(.y) - intersection)

# melt and plot with geom_tile
split.ovl.melt <- reshape2::melt(split.ovl)

# reverse sample order
split.ovl.melt$Var2 <- factor(split.ovl.melt$Var2, levels = rev(unique(split.ovl.melt$Var2)))

ggplot(split.ovl.melt, aes(Var1, Var2, fill = value)) + 
  geom_tile() +
  labs(x = '', y = '', fill = 'Jaccard index') +
  scale_fill_gradientn(
   colors=scico::scico(1000, palette = 'vik'),
   values=scales::rescale(c(0,0.01,0.21)),
   limits=c(0,0.21)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.border = element_blank())

# plot within and between sample overlap values for FIP and Control
upper <- which(upper.tri(split.ovl, diag = F), arr.ind = T)

split.upper <- data.frame(first = dimnames(split.ovl)[[2]][upper[,2]],
                second = dimnames(split.ovl)[[1]][upper[,1]],
                correlation = split.ovl[upper])

# remove overlaps between FIP and control
split.upper$group <- ifelse(grepl('control', split.upper$first) & grepl('control', split.upper$second), 'Control',
                              ifelse(grepl('fip', split.upper$first) & grepl('fip', split.upper$second), 'FIP', 'NA'))

split.upper.filt <- split.upper[split.upper$group != 'NA',]

split.upper.filt$inout <- ifelse(str_split_i(split.upper.filt$first, '_', 2) == str_split_i(split.upper.filt$second, '_', 2), 'Within', 'Between')

library(ggplot2)
library(ggpubr)

ggplot(split.upper.filt, aes(inout, correlation, fill = group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_point(aes(x = inout, y = correlation, color = group), position = position_jitterdodge(dodge.width = 0.9, jitter.width = 0.1)) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  scale_y_continuous(limits = c(0,0.25)) +
  stat_compare_means(method = 'wilcox.test') +
  labs(x = 'Cat overlap', y = 'Overlap index (Jaccard)', fill = '', color = '') +
  theme_bw() +
  theme(panel.grid = element_blank())
```

Look at top clusters (by FIP sample membership) and get seq logos. 

```{r nair_global_cluster_top}
library(igraph)
library(ggraph)
library(netdiffuseR)
library(qgraph)
library(ggseqlogo)
library(ggplot2)
library(ggpubr)

clust.tab <- as.data.frame(reshape2::acast(igh_mixcr.f.net$node_data, cluster_id ~ Sample, fill = 0))
clust.tab$nsamples <- rowSums(clust.tab > 0)
clust.tab$cluster <- rownames(clust.tab)
clust.tab <- subset(clust.tab, select = c(cluster, nsamples, 1:42))

fip.clust <- clust.tab %>%
  filter_at(vars(c(3:8, 12:23)), all_vars(. == 0))

control.clust <- clust.tab %>%
  filter_at(vars(c(24:44)), all_vars(. == 0))

cats.fip.clust <- fip.clust %>%
  mutate(control_135 = select(., control_135_liver:control_135_mln) %>% rowSums(),
         control_136 = select(., control_136_liver:control_136_mln) %>% rowSums(),
         control_164 = select(., control_164_liver:control_164_mln) %>% rowSums(),
         control_174 = select(., control_174_liver:control_174_mln) %>% rowSums(),
         control_188 = select(., control_188_liver:control_188_mln) %>% rowSums(),
         control_54 = select(., control_54_liver:control_54_mln) %>% rowSums(),
         control_69 = select(., control_69_liver:control_69_mln) %>% rowSums(),
         fip_120 = select(., fip_120_liver:fip_120_mln) %>% rowSums(),
         fip_142 = select(., fip_142_liver:fip_142_mln) %>% rowSums(),
         fip_172 = select(., fip_172_liver:fip_172_mln) %>% rowSums(),
         fip_192 = select(., fip_192_liver:fip_192_mln) %>% rowSums(),
         fip_37 = select(., fip_37_liver:fip_37_mln) %>% rowSums(),
         fip_68 = select(., fip_68_liver:fip_68_mln) %>% rowSums(),
         fip_86 = select(., fip_86_liver:fip_86_mln) %>% rowSums(),
         fip_96 = select(., fip_96_liver:fip_96_mln) %>% rowSums())

cats.fip.clust <- subset(cats.fip.clust, select = c(1, 2, 45:59))
cats.fip.clust$ncats <- rowSums(cats.fip.clust[,3:17] > 0)

# compare CDR3 length of 'shared' clusters to all other clusters
# define 'shared' clusters as present in at least 2 FIP cats and zero controls
cats.fip.clust.filt <- cats.fip.clust[cats.fip.clust$ncats > 1,]

fip.shared.clust <- cats.fip.clust.filt$cluster
other.clust <- clust.tab$cluster[!clust.tab$cluster %in% fip.shared.clust]

cluster.dat <- igh_mixcr.f.net$cluster_data
cluster.dat$seqlength <- nchar(cluster.dat$seq_w_max_degree)
cluster.dat$group <- factor(ifelse(cluster.dat$cluster_id %in% fip.shared.clust, 'FIP cluster', 'Other'), levels = c('Other', 'FIP cluster'))

ggplot(cluster.dat, aes(group, seqlength, fill = group)) +
  geom_boxplot() +
  stat_compare_means() +
  theme_bw()

# similar clusters identified - 24, 50, 219, 80, 124
# add cat id to nodedata for plotting
igh_mixcr.f.net$node_data$Cat <- paste0(stringr::str_split_i(igh_mixcr.f.net$node_data$Sample, '_', 1), 
                                        '_', 
                                        stringr::str_split_i(igh_mixcr.f.net$node_data$Sample, '_', 2))

igh_mixcr.f.net$node_data$Cat <- factor(igh_mixcr.f.net$node_data$Cat, levels = c('control_54','control_69','control_135','control_136','control_174','control_188','control_164','fip_142','fip_192','fip_37','fip_96','fip_120','fip_172','fip_68','fip_86'))

# set colour scale
cols <- c('#32A7CD', '#D071AD', '#D071AD', '#f4eaf4', '#d2acd3', '#b16eb3', '#8f3092', '#642266', '#0e050f')
names(cols) <- c('control_164', 'fip_142', 'fip_192', 'fip_37', 'fip_96', 'fip_120', 'fip_172', 'fip_68', 'fip_86')

# plot some top FIP clusters by nsamples - 24, 50, 219, 80, 124
# cluster 24
cluster24 <- subset(igh_mixcr.f.net$node_data[igh_mixcr.f.net$node_data$cluster_id == 24,], select = c(1:38, 49))

cluster24.net <- buildNet(data = cluster24,
                          seq_col = 'CDR3.aa',
                          count_col = 'Proportion',
                          color_nodes_by = 'Cat',
                          size_nodes_by = 'Proportion',
                          drop_isolated_nodes = FALSE,
                          node_stats = TRUE,
                          cluster_stats = TRUE)

# nice plot
cluster24.ig <- cluster24.net$igraph
V(cluster24.ig)$size <- cluster24.net$node_data$Proportion
V(cluster24.ig)$cat <- cluster24.net$node_data$Cat

layout24 = qgraph.layout.fruchtermanreingold(as_edgelist(cluster24.ig, names = FALSE), 
                                             vcount = vcount(cluster24.ig),
                                             area = 30 * vcount(cluster24.ig)^2)

ggraph(cluster24.ig, layout = layout24) +
    geom_edge_link(color = "darkgrey", width = 0.2, alpha = 0.7) +
    geom_node_point(aes(fill = V(cluster24.ig)$cat), 
                    size = rescale_vertex_igraph(V(cluster24.ig)$size, minmax.relative.size = c(0.02, 0.08), adjust = 200),
                    shape = 21, show.legend = TRUE) +
    scale_fill_manual(values = cols) +
    labs(fill = '') +
    theme_void()

# seqlogo
ggseqlogo(cluster24.net$node_data$CDR3.aa[nchar(cluster24.net$node_data$CDR3.aa) == 11], method = 'bits')

# cluster 50
cluster50 <- subset(igh_mixcr.f.net$node_data[igh_mixcr.f.net$node_data$cluster_id == 50,], select = c(1:38, 49))

cluster50.net <- buildNet(data = cluster50,
                          seq_col = 'CDR3.aa',
                          count_col = 'Proportion',
                          color_nodes_by = 'Sample',
                          size_nodes_by = 'Proportion',
                          drop_isolated_nodes = FALSE,
                          node_stats = TRUE,
                          cluster_stats = TRUE)

# nice plot
cluster50.ig <- cluster50.net$igraph
V(cluster50.ig)$size <- cluster50.net$node_data$Proportion
V(cluster50.ig)$cat <- cluster50.net$node_data$Cat

layout50 = qgraph.layout.fruchtermanreingold(as_edgelist(cluster50.ig, names = FALSE), 
                                             vcount = vcount(cluster50.ig),
                                             area = 30 * vcount(cluster50.ig)^2)

ggraph(cluster50.ig, layout = layout50) +
    geom_edge_link(color = "darkgrey", width = 0.2, alpha = 0.7) +
    geom_node_point(aes(fill = V(cluster50.ig)$cat), 
                    size = rescale_vertex_igraph(V(cluster50.ig)$size, minmax.relative.size = c(0.02, 0.08), adjust = 200),
                    shape = 21, show.legend = FALSE) +
    scale_fill_manual(values = cols) +
    labs(color = '') +
    theme_void()

# seqlogo
ggseqlogo(cluster50.net$node_data$CDR3.aa, method = 'bits')

# cluster 80
cluster80 <- subset(igh_mixcr.f.net$node_data[igh_mixcr.f.net$node_data$cluster_id == 80,], select = c(1:38, 49))

cluster80.net <- buildNet(data = cluster80,
                          seq_col = 'CDR3.aa',
                          count_col = 'Proportion',
                          color_nodes_by = 'Sample',
                          size_nodes_by = 'Proportion',
                          drop_isolated_nodes = FALSE,
                          node_stats = TRUE,
                          cluster_stats = TRUE)

# nice plot
cluster80.ig <- cluster80.net$igraph
V(cluster80.ig)$size <- cluster80.net$node_data$Proportion
V(cluster80.ig)$cat <- cluster80.net$node_data$Cat

layout80 = qgraph.layout.fruchtermanreingold(as_edgelist(cluster80.ig, names = FALSE), 
                                             vcount = vcount(cluster80.ig),
                                             area = 30 * vcount(cluster80.ig)^2)

ggraph(cluster80.ig, layout = layout80) +
    geom_edge_link(color = "darkgrey", width = 0.2, alpha = 0.7) +
    geom_node_point(aes(fill = V(cluster80.ig)$cat), 
                    size = rescale_vertex_igraph(V(cluster80.ig)$size, minmax.relative.size = c(0.02, 0.08), adjust = 200),
                    shape = 21, show.legend = FALSE) +
    scale_fill_manual(values = cols) +
    labs(color = '') +
    theme_void()

# seqlogo
ggseqlogo(cluster80.net$node_data$CDR3.aa, method = 'bits')

# cluster 219
cluster219 <- subset(igh_mixcr.f.net$node_data[igh_mixcr.f.net$node_data$cluster_id == 219,], select = c(1:38, 49))

cluster219.net <- buildNet(data = cluster219,
                          seq_col = 'CDR3.aa',
                          count_col = 'Proportion',
                          color_nodes_by = 'Sample',
                          size_nodes_by = 'Proportion',
                          drop_isolated_nodes = FALSE,
                          node_stats = TRUE,
                          cluster_stats = TRUE)

# seqlogo
ggseqlogo(cluster219.net$node_data$CDR3.aa, method = 'bits')

# cluster 124
cluster124 <- subset(igh_mixcr.f.net$node_data[igh_mixcr.f.net$node_data$cluster_id == 124,], select = c(1:38, 49))

cluster124.net <- buildNet(data = cluster124,
                          seq_col = 'CDR3.aa',
                          count_col = 'Proportion',
                          color_nodes_by = 'Sample',
                          size_nodes_by = 'Proportion',
                          drop_isolated_nodes = FALSE,
                          node_stats = TRUE,
                          cluster_stats = TRUE)

# seqlogo
ggseqlogo(cluster124.net$node_data$CDR3.aa, method = 'bits')
```

## Somatic hypermutation analysis

To detect mutations in IGH VDJ sequences, clonotypes were re-assembled for all samples using MiXCR with the option `--assemble-contigs-by VDJRegion` mix-in option. Allelic variation was then accounted for by re-assigning clonotypes to alleles using MiXCR's `findAlleles` function:

```{r mixcr_find_alleles_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J mixcr_find_alleles.job 
#SBATCH -o mixcr_find_alleles.%A_%a.out 
#SBATCH -c 2

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`

cats=()
for cat in *_mln_merged.contigs.clns
do
        cat=${cat%_mln_merged.contigs.clns}
        cats[${#cats[@]}]=$cat
done

# create input file variables if the file exists, otherwise leave them empty
if [ -e "${cats[$i]}_liver_merged.contigs.clns" ]
then
	liverfile=$path/${cats[$i]}_liver_merged.contigs.clns
else
	liverfile=""
fi

if [ -e "${cats[$i]}_lung_merged.contigs.clns" ]
then
	lungfile=$path/${cats[$i]}_lung_merged.contigs.clns
else
	lungfile=""
fi

if [ -e "${cats[$i]}_mln_merged.contigs.clns" ]
then
	mlnfile=$path/${cats[$i]}_mln_merged.contigs.clns
else
	mlnfile=""
fi

outpath=findAlleles

#function
mixcr findAlleles \
      --report $outpath/${cats[$i]}.findAlleles.report.txt \
      --export-alleles-mutations $outpath/${cats[$i]}_alleles.tsv \
      --export-library $outpath/${cats[$i]}_alleles.json \
      --output-template $outpath/{file_name}.reassigned.clns \
      $liverfile $lungfile $mlnfile
```

```{r mixcr_find_alleles_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
cats=()
for cat in *_mln_merged.contigs.clns
do
        cat=${cat%_mln_merged.contigs.clns}
        cats[${#cats[@]}]=$cat
done
catcount=${#cats[@]}

sbatch --array 1-$catcount:1 mixcr_find_alleles_secondary.sh
```

Full-length clonotypes were then exported in AIRR format using MiXCR's `exportAirr`:

```{r mixcr_export_airr_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J mixcr_export_airr_secondary.job 
#SBATCH -o mixcr_export_airr_secondary.%A_%a.out 
#SBATCH -c 1

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`

files=()
for file in *_merged.contigs.reassigned.clns
do
        file=${file%_merged.contigs.reassigned.clns}
        files[${#files[@]}]=$file
done

clnsfile=$path/${files[$i]}_merged.contigs.reassigned.clns

outpath=airr

mixcr exportAirr \
    --imgt-gaps \
    $clnsfile \
    $outpath/${files[$i]}_airr_condatest.tsv
```

```{r mixcr_export_airr_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
files=()
for file in *_merged.contigs.reassigned.clns
do
        file=${file%_merged.contigs.reassigned.clns}
        files[${#files[@]}]=$file
done
filecount=${#files[@]}

sbatch --array 1-$filecount:1 mixcr_export_airr_secondary.sh
```

Analyse mutations using [Shazam](https://shazam.readthedocs.io/en/stable/).

```{r shazam_mutation_analysis}
# read in AIRR format data
airr.files <- list.files('mixcr/IGH/fulllength/findAlleles/airr', pattern = '.tsv', full.names = TRUE)
airr.id <- airr.files %>% substr(46, nchar(.)-19) %>% unique()
airr.list <- lapply(airr.files, function(x) read.csv(x, sep = '\t', quote = ""))
names(airr.list) <- airr.id

# calculate R(eplacement) and S(ilent) mutation counts for all samples
library(alakazam)
library(shazam)
library(tidyverse)
library(ggplot2)
library(ggpubr)

db_obs <- lapply(airr.list, function(x) {
  observedMutations(x, sequenceColumn = "sequence_alignment",
                    germlineColumn = "germline_alignment",
                    regionDefinition = NULL,
                    frequency = TRUE,
                    nproc = 2)
})

# extract mutation freqs for all samples
mut.freq <- lapply(db_obs, function(x) {
  x %>% select(sequence_id, c_call, starts_with("mu_freq"))
})

# collapse IGHG subtypes
mut.freq <- mut.freq %>%
  map(~ .x %>% mutate(c_call = gsub('IGHG1|IGHG2|IGHG3', 'IGHG', c_call)))

# get mean frequency in each sample and collapse to single dataframe
# add metadata for plotting
mut.freq.summ <- lapply(mut.freq, function(x) {
  x %>%
  summarise(.by = c_call, meanRmut = mean(mu_freq_seq_r), meanSmut = mean(mu_freq_seq_s)) %>%
  complete(c_call = c('IGHM', 'IGHA', 'IGHD', 'IGHE', 'IGHG', ''), fill = list(meanRmut = 0, meanSmut = 0))
})

mut.freq.all <- bind_rows(mut.freq.summ, .id = 'Sample')

mut.freq.all$Group <- ifelse(grepl('control', mut.freq.all$Sample), 'Control', 'FIP')
mut.freq.all$Tissue <- ifelse(grepl('liver', mut.freq.all$Sample), 'Liver', 
                              ifelse(grepl('lung', mut.freq.all$Sample), 'Lung', 
                                     'MLN'))
mut.freq.all$Group2 <- ifelse(grepl('142|192', mut.freq.all$Sample), 'FIP (focal renal)', mut.freq.all$Group)

# plot mean freq in FIP vs Control for IGHG
# R mutations
r.plot <- ggplot(mut.freq.all[mut.freq.all$c_call == 'IGHG',], aes(Group, meanRmut, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Group2), width = 0.1) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092', '#D071AD')) +
  scale_y_continuous(limits = c(0,0.06), breaks = seq(0,0.06,0.01)) +
  labs(x = '', y = 'Mean replacement mutation frequency', fill = '') +
  stat_compare_means() +
  facet_grid(rows = vars(c_call), cols = vars(Tissue)) +
  ggtitle('IGHG') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5))

# S mutations
s.plot <- ggplot(mut.freq.all[mut.freq.all$c_call == 'IGHG',], aes(Group, meanSmut, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5, show.legend = FALSE) +
  geom_jitter(aes(color = Group2), width = 0.1, show.legend = TRUE) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092', '#D071AD')) +
  scale_y_continuous(limits = c(0,0.035), breaks = seq(0,0.04,0.01)) +
  labs(x = '', y = 'Mean silent mutation frequency', fill = '') +
  stat_compare_means() +
  facet_wrap(vars(Tissue), nrow = 1) +
  ggtitle('IGHG') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5))

compare_means(meanSmut ~ Group, data = mut.freq.all[mut.freq.all$c_call == 'IGHG',], group.by = 'Tissue', method = 'wilcox.test', p.adjust.method = 'bonferroni')

library(cowplot)

plot_grid(r.plot, s.plot, nrow = 2)

# should plot this by FCoV positivity, as expect the relationship to be even stronger...
# convert fcov data to annotations
fcov$virPos <- ifelse(fcov$value > -6.38, TRUE, FALSE)

mut.freq.all.virPos <- merge(mut.freq.all, fcov[,4:5], by = 'Sample', all.x = TRUE)
# replace NAs
mut.freq.all.virPos$virPos[is.na(mut.freq.all.virPos$virPos)] <- FALSE

# plot
# R mutations
ggplot(mut.freq.all.virPos[mut.freq.all.virPos$c_call == 'IGHG',], aes(virPos, meanRmut, fill = virPos)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Group), width = 0.1) +
  scale_fill_manual(values = c('white', 'grey')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  stat_compare_means() +
  facet_wrap(vars(Tissue), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())

# S mutations
ggplot(mut.freq.all.virPos[mut.freq.all.virPos$c_call == 'IGHG',], aes(virPos, meanSmut, fill = virPos)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = Group), width = 0.1) +
  scale_fill_manual(values = c('white', 'grey')) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  stat_compare_means() +
  facet_wrap(vars(Tissue), nrow = 1) +
  theme_bw() +
  theme(panel.grid = element_blank())

# plot the density of mutations in FIP and Control tissue samples
mut.freq.dens <- bind_rows(mut.freq, .id = 'Sample')

# add metadata
mut.freq.dens$Group <- ifelse(grepl('control', mut.freq.dens$Sample), 'Control', 'FIP')
mut.freq.dens$Tissue <- ifelse(grepl('liver', mut.freq.dens$Sample), 'Liver', 
                              ifelse(grepl('lung', mut.freq.dens$Sample), 'Lung', 
                                     'MLN'))
mut.freq.dens$Cat <- str_split_i(mut.freq.dens$Sample, '_', 2)
mut.freq.dens$Cat <- factor(mut.freq.dens$Cat, levels = c(54, 69, 135, 136, 164, 174, 188, 37, 68, 86, 96, 120, 142, 172, 192))

# set factor levels
mut.freq.dens$c_call <- factor(mut.freq.dens$c_call, levels = c('IGHM', 'IGHG', 'IGHA'))

ggplot(mut.freq.dens[mut.freq.dens$c_call %in% c('IGHM', 'IGHG', 'IGHA'),], aes(x = mu_freq_seq_s, colour = Group)) +
  geom_density(bw = 0.005) +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,60)) +
  facet_grid(rows = vars(Tissue), cols = vars(c_call)) +
  labs(x = 'Silent mutation frequency', y = 'Density', color = '') +
  theme_bw() +
  theme(panel.grid = element_blank())

# alt boxplots
ggplot(mut.freq.dens[mut.freq.dens$c_call %in% c('IGHM', 'IGHG', 'IGHA'),], aes(x = Group, y = mu_freq_seq_r)) +
  geom_boxplot(aes(fill = Group), alpha = 0.5, outlier.shape = NA) +
  geom_jitter(aes(colour = Group), width = 0.1, show.legend = F) +
  stat_compare_means() +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  facet_grid(rows = vars(c_call), cols = vars(Tissue)) +
  theme_bw()

ggplot(mut.freq.dens[mut.freq.dens$c_call %in% c('IGHM', 'IGHG', 'IGHA'),], aes(x = Group, y = mu_freq_seq_s)) +
  geom_boxplot(aes(fill = Group), alpha = 0.5, outlier.shape = NA) +
  geom_jitter(aes(colour = Group), width = 0.1, show.legend = F) +
  stat_compare_means() +
  scale_color_manual(values = c('#32A7CD', '#8F3092')) +
  scale_fill_manual(values = c('#32A7CD', '#8F3092')) +
  facet_grid(rows = vars(c_call), cols = vars(Tissue)) +
  theme_bw()

# add age
control.igg$cat <- str_split_i(control.igg$Sample, pattern = '_', i = 2)

cat.age <- read.csv(file = 'tables/cat_ages.txt', sep = '\t', header = TRUE)
cat.age$cat <- str_split_i(cat.age$cat, pattern = '_', i = 2)

control.igg.age <- merge(control.igg, cat.age, by = 'cat', all.x = TRUE)

# MLN only
control.igg.age <- control.igg.age[control.igg.age$Tissue == 'MLN',]

ggplot(control.igg.age, aes(age, meanSmut)) +
  geom_point(color = '#32A7CD') +
  geom_smooth(method = 'lm', se = FALSE, color = '#32A7CD') +
  scale_x_continuous(expand = c(0,0), limits = c(0,200)) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 0.03)) +
  labs(x = 'Age (months)', y = 'Mean silent mutation frequency') +
  theme_bw() +
  theme(panel.grid = element_blank())

cor.test(control.igg.age$meanSmut, control.igg.age$age)

summary(lm(meanSmut ~ age + cat, data = control.igg.age))

# FIP cats
fip.igg <- mut.freq.all[mut.freq.all$Group == 'FIP' & mut.freq.all$c_call == 'IGHG',]

# add age
fip.igg$cat <- str_split_i(fip.igg$Sample, pattern = '_', i = 2)

cat.age <- read.csv(file = 'tables/cat_ages.txt', sep = '\t', header = TRUE)
cat.age$cat <- str_split_i(cat.age$cat, pattern = '_', i = 2)

fip.igg.age <- merge(fip.igg, cat.age, by = 'cat', all.x = TRUE)

# MLN only
fip.igg.age <- fip.igg.age[fip.igg.age$Tissue == 'MLN',]
# exclude 37
fip.igg.age <- fip.igg.age[fip.igg.age$cat != 37,]

ggplot(fip.igg.age, aes(age, meanSmut)) +
  geom_point(color = '#8F3092') +
  geom_smooth(method = 'lm', se = FALSE, color = '#8F3092') +
  scale_x_continuous(expand = c(0,0), limits = c(0,40)) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 0.02)) +
  labs(x = 'Age (months)', y = 'Mean silent mutation frequency') +
  theme_bw() +
  theme(panel.grid = element_blank())

cor.test(fip.igg.age$meanSmut, fip.igg.age$age)

summary(lm(meanSmut ~ age + Tissue + cat, data = fip.igg.age))
```

Estimate selection scores.

```{r shazam_selection}
# use all clones from all control samples (collapsed clones) to infer full targeting model
controls.airr <- airr.list[grepl('control_', names(airr.list))]
# add names as sample column
for (i in seq_along(controls.airr)) {
  controls.airr[[i]]$Sample <- names(controls.airr[i])
}

# filter down to IGHG, remove rows with blank sequence_alignment, reformat clone_id and sequence_id
controls.igg.airr <- lapply(controls.airr, function(x) {
  x[x$c_call %in% c('IGHG1', 'IGHG2', 'IGHG3'),]
  x[!x$sequence_alignment == '',]
  x[!x$germline_alignment == '',]
  x[!x$v_call == '',]
  x$clone_id = paste(x$Sample, x$clone_id, sep = '.')
  x$sequence_id <- paste(x$Sample, x$sequence_id, sep = '.')
  x
})

# subset list by cat id
c135.igg.airr <- controls.igg.airr[1:3]
c136.igg.airr <- controls.igg.airr[4:6]
c164.igg.airr <- controls.igg.airr[7:9]
c174.igg.airr <- controls.igg.airr[10:12]
c188.igg.airr <- controls.igg.airr[13:15]
c54.igg.airr <- controls.igg.airr[16:18]
c69.igg.airr <- controls.igg.airr[19:21]

cats.igg.airr <- list(c135.igg.airr, c136.igg.airr, c164.igg.airr, c174.igg.airr, c188.igg.airr, c54.igg.airr, c69.igg.airr)

# collapse each sublist into single dataframe
cats.igg.comb <- lapply(cats.igg.airr, function(x) {
  bind_rows(x, .id = 'Sample')
})

names(cats.igg.comb) <- c('control_135', 'control_136', 'control_164', 'control_174', 'control_188', 'control_54', 'control_69')

cats.igg.clns <- lapply(cats.igg.comb, function(x) {
  collapseClones(x, cloneColumn = "clone_id", 
                    sequenceColumn = "sequence_alignment",
                    germlineColumn = "germline_alignment",
                    method="thresholdedFreq", minimumFrequency=0.6,
                    includeAmbiguous=FALSE, breakTiesStochastic=FALSE,
                    nproc = 2)
})

# create single df
all.igg.clns <- bind_rows(cats.igg.clns)

# remove rows with missing info
all.igg.clns <- all.igg.clns[all.igg.clns$clonal_sequence != '',]

# Create targeting model in one step using only silent mutations
# Use consensus sequence input and germline columns
model <- createTargetingModel(all.igg.clns, model="s", sequenceColumn="clonal_sequence", 
                              germlineColumn="clonal_germline", vCallColumn="v_call")

# Generate plots of mutability model
plotMutability(model, nucleotides="A", style="bar")
plotMutability(model, nucleotides="C", style="bar")
plotMutability(model, nucleotides="G", style="bar")
plotMutability(model, nucleotides="T", style="bar")

# apply model to all samples
all.airr <- airr.list
# add names as sample column
for (i in seq_along(all.airr)) {
  all.airr[[i]]$Sample <- names(all.airr[i])
}

# remove rows with blank sequence_alignment or c_call, reformat clone_id and sequence_id, collapse IGHG subtypes
all.airr <- lapply(all.airr, function(x) {
  x[!x$c_call == '',]
  x[!x$sequence_alignment == '',]
  x[!x$germline_alignment == '',]
  x[!x$v_call == '',]
  x$clone_id = paste(x$Sample, x$clone_id, sep = '.')
  x$sequence_id <- paste(x$Sample, x$sequence_id, sep = '.')
  x$c_call <- gsub('IGHG1|IGHG2|IGHG3', 'IGHG', x$c_call)
  x
})

# filter for IGHG only
ighg.airr <- lapply(all.airr, function(x) x[x$c_call == 'IGHG',])

# subset list by cat id
c135.airr <- ighg.airr[1:3]
c136.airr <- ighg.airr[4:6]
c164.airr <- ighg.airr[7:9]
c174.airr <- ighg.airr[10:12]
c188.airr <- ighg.airr[13:15]
c54.airr <- ighg.airr[16:18]
c69.airr <- ighg.airr[19:21]
f120.airr <- ighg.airr[22:23]
f142.airr <- ighg.airr[24:26]
f172.airr <- ighg.airr[27:29]
f192.airr <- ighg.airr[30:32]
f37.airr <- ighg.airr[33:34]
f68.airr <- ighg.airr[35:37]
f86.airr <- ighg.airr[38:40]
f96.airr <- ighg.airr[41:42]

cats.airr <- list(c135.airr, c136.airr, c164.airr, c174.airr, c188.airr, c54.airr, c69.airr,
                  f120.airr, f142.airr, f172.airr, f192.airr, f37.airr, f68.airr, f86.airr, f96.airr)

# collapse each sublist into single dataframe
cats.comb <- lapply(cats.airr, function(x) {
  bind_rows(x, .id = 'Sample')
})

names(cats.comb) <- c('control_135', 'control_136', 'control_164', 'control_174', 'control_188', 'control_54', 'control_69', 
                      'fip_120', 'fip_142', 'fip_172', 'fip_192', 'fip_37', 'fip_68', 'fip_86', 'fip_96')

cats.clns <- lapply(cats.comb, function(x) {
  collapseClones(x, cloneColumn = "clone_id", 
                    sequenceColumn = "sequence_alignment",
                    germlineColumn = "germline_alignment",
                    regionDefinition=IMGT_V, 
                    method="thresholdedFreq", minimumFrequency=0.6,
                    includeAmbiguous=FALSE, breakTiesStochastic=FALSE,
                    nproc = 2)
})

# create single df
all.clns <- bind_rows(cats.clns, .id = 'Sample')

# remove rows with missing info
all.clns <- all.clns[!all.clns$clonal_sequence == '',]

# Count observed mutations and append mu_count columns to the output
observed <- observedMutations(all.clns, 
                              sequenceColumn="clonal_sequence",
                              germlineColumn="clonal_germline",
                              regionDefinition=IMGT_V, nproc=1)

expected <- expectedMutations(observed, 
                              sequenceColumn="clonal_sequence",
                              germlineColumn="clonal_germline",
                              targetingModel=model,
                              regionDefinition=IMGT_V, nproc=1)

# add metadata
expected$Group <- ifelse(grepl('control', expected$Sample), 'Control', 'FIP')
expected$Tissue <- ifelse(grepl('liver', expected$clone_id), 'Liver',
                              ifelse(grepl('lung', expected$clone_id), 'Lung',
                                     'MLN'))

# Calculate selection scores using the output from expectedMutations
baseline <- calcBaseline(expected, testStatistic="focused", targetingModel = model,
                         regionDefinition=IMGT_V, nproc=1)

# First group by subject and status
subject_grouped <- groupBaseline(baseline, groupBy = c("Group", "Tissue"))

# Then group the output by status
status_grouped <- groupBaseline(subject_grouped, groupBy = "Group")

testBaseline(status_grouped, groupBy = "Group")

# Plot mean and confidence interval by group
# Group by CDR/FWR and facet by isotype
status_grouped@stats$Group <- factor(status_grouped@stats$Group, levels = c('FIP', 'Control'))

plotBaselineSummary(status_grouped, idColumn = "Group") + 
  coord_flip() +
  labs(y = 'Σ (selection score)', x = '') +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```


