---
title: "Fatal consequences of feline coronavirus infection are associated with virus persistence and a distinct adaptive immune repertoire - FCoV sequence analysis pipeline"
author: "Tom Hiron"
date: "2025"
output: 
  html_document:
    theme: "simplex"
    toc: true
    toc_float: true
    highlight: kate
    code_folding: show
    self_contained: true
---
```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      echo = TRUE, 
                      eval = TRUE,
                      fig.path = 'figures/',
                      fig.ext = 'png',
                      cache = FALSE,
                      cache.path = 'cache/')
```

Pipeline to reproduce feline coronavirus (FCoV) sequence analyses in the manuscript "Fatal consequences of feline coronavirus infection are associated with virus persistence and a distinct adaptive immune repertoire", Hiron et al. 2025, https://www.biorxiv.org/content/10.1101/2025.10.02.679983v1.

## Data availability

Sequencing data supporting the findings of this study have been deposited in the National Center for Biotechnology Information (NCBI) Sequence Read Archive under BioProject ID PRJNA1330231.

## Alignment to FCoV C1Je (DQ848678)

STAR v2.7.11b.

Prepare STAR index for hybrid assembly of felCat9 and FCoV C1Je (DQ848678). A FASTA file for the hybrid assembly was prepared by concatenating the FCoV C1Je genome sequence and the felCat9 genome FASTA used for the main RNA-seq analysis. A hybrid assembly GTF file was prepared by adding the following lines to the felCat9 genome GTF from Ensembl:

```
c1je_DQ848678.1 Prodigal_v2.6.3 gene    1       29255   .       +       .       gene_id "c1je_genome"
c1je_DQ848678.1 Prodigal_v2.6.3 exon    1       29255   .       +       .       gene_id "c1je_genome"
c1je_DQ848678.1 Prodigal_v2.6.3 CDS     1       29255   .       +       .       gene_id "c1je_genome"
```

```{r alignSTAR_fcov_index, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J build_STAR_index_fc9_ensembl111_c1je.job 
#SBATCH -o build_STAR_index_fc9_ensembl111_c1je.out 
#SBATCH -c 8

refdir=felCat9_c1je

STAR --runThreadN 8 --runMode genomeGenerate \
--genomeDir $refdir/fc9_ens111_c1je_STAR_index \
--genomeFastaFiles $refdir/Felis_catus.Felis_catus_9.0.c1je.dna.toplevel.fa \
--sjdbGTFfile $refdir/Felis_catus.Felis_catus_9.0.111.c1je.gtf \
--sjdbOverhang 149
```

Mapping:

```{r alignSTAR_fcov_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J align_STAR_feline_c1je.job 
#SBATCH -o align_STAR_feline_c1je.%A_%a.out 
#SBATCH -c 2

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
inpath=fastq
outpath=fcov_reads/alignments

cd $inpath

fnames=()
for fname in *_1.trimmed.fastq.gz
do
	fname=${fname%_1.trimmed.fastq.gz}
	fnames[${#fnames[@]}]=$fname
done

cd $path

#input variables
seqfile1=$inpath/${fnames[$i]}_1.trimmed.fastq.gz
seqfile2=$inpath/${fnames[$i]}_2.trimmed.fastq.gz

#other variables
outprefix=$outpath/${fnames[$i]}_
refdir=fc9_ens111_c1je_STAR_index

STAR --runThreadN 2 --genomeDir $refdir --readFilesIn $seqfile1 $seqfile2 --outFileNamePrefix $outprefix --outSAMunmapped Within \
    --outFilterMultimapNmax 10 --alignSJoverhangMin 5 --alignSJDBoverhangMin 3 --outFilterMismatchNmax 10 --outFilterMismatchNoverLmax 0.04 \
    --alignIntronMin 21 --alignIntronMax 0 --alignMatesGapMax 0 --outFilterIntronMotifs RemoveNoncanonical --outSAMtype BAM SortedByCoordinate \
    --readFilesCommand zcat
```

```{r alignSTAR_fcov_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
inpath=fastq

cd $inpath

fnames=()
for fname in *_1.trimmed.fastq.gz
do
	fname=${fname%_1.trimmed.fastq.gz}
	fnames[${#fnames[@]}]=$fname
done
fnamecount=${#fnames[@]}

cd $path

sbatch --array 1-$fnamecount:1 align_STAR_feline_c1je_secondary.sh
```

## Filtering

Samtools v1.21.

Filter alignments to retain uniquely aligned and properly paired reads. 

```{r bam_filter_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J bam_filter.job 
#SBATCH -o bam_filter.%A_%a.out 
#SBATCH -c 2

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
outpath=filtered

fnames=()
for fname in *.sortedByCoord.out.bam
do
        fname=${fname%.sortedByCoord.out.bam}
        fnames[${#fnames[@]}]=$fname
done

infile=$path/${fnames[$i]}.sortedByCoord.out.bam
outfile=$outpath/${fnames[$i]}.filtered.bam

#function
samtools view -h -b -f 2 -q 255 $infile > $outfile
samtools index $outfile
```

```{r bam_filter_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
fnames=()
for fname in *.sortedByCoord.out.bam
do
        fname=${fname%.sortedByCoord.out.bam}
        fnames[${#fnames[@]}]=$fname
done
fnamecount=${#fnames[@]}

sbatch --array 1-$fnamecount:1 bam_filter_secondary.sh
```

## Read summarisation

Samtools v1.21.

Sort alignments by name: 

```{r bam_nameSort_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash
#SBATCH -J bam_nameSort.job 
#SBATCH -o bam_nameSort.%A_%a.out 
#SBATCH -c 4

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`

bams=()
for bam in *filtered.bam
do
        bam=${bam%.bam}
        bams[${#bams[@]}]=$bam
done

#define variables
infile=$path/${bams[$i]}.bam
outfile=$path/${bams[$i]}.nameSorted.bam

#function
samtools sort -n -@ 4 $infile -o $outfile
```

```{r bam_nameSort_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
bams=()
for bam in *filtered.bam
do
        bam=${bam%.bam}
        bams[${#bams[@]}]=$bam
done
bamcount=${#bams[@]}

sbatch --array 1-$bamcount:1 bam_nameSort_secondary.sh
```

Rsubread v2.20.

Get gene-level counts:

```{r read_summ, eval = FALSE}
bam.files <- list.files(pattern = "*nameSorted.bam$", full.names = TRUE)

library(Rsubread)

ens111_c1je_fc <- featureCounts(files = bam.files, 
                                annot.ext = "Felis_catus.Felis_catus_9.0.111.c1je.gtf", 
                                isGTFAnnotationFile = TRUE, 
                                GTF.attrType = 'gene_id', 
                                isPairedEnd = TRUE, 
                                nthreads = 4)
```

## FCoV abundance heatmap

```{r fcov_counts_hmap}
library(edgeR)
library(stringr)
library(reshape2)

# get normalised FCoV counts
fcov.counts <- ens111_c1je_fc$counts
colnames(fcov.counts) <- str_remove(colnames(fcov.counts), 'feline_')
colnames(fcov.counts) <- str_remove(colnames(fcov.counts), '_Aligned.filtered.nameSorted.bam')

fcov.counts <- as.data.frame(fcov.counts)
fcov.counts$Gene <- rownames(fcov.counts)
rownames(fcov.counts) <- NULL
fcov.counts <- subset(fcov.counts, select = c(43, 1:42))

z <- DGEList(counts = fcov.counts[,2:43], genes = fcov.counts$Gene, remove.zeros = TRUE) 

z$samples$group <- factor(c(rep('Control', 21), rep('FIP', 21)), levels = c('Control', 'FIP'))
z$samples$tissue <- ifelse(grepl('liver', rownames(z$samples)), 'Liver', ifelse(grepl('lung', rownames(z$samples)), 'Lung', 'MLN'))

# recalculate library sizes
z$samples$lib.size <- colSums(z$counts)

# use Ensembl IDs as row names
rownames(z$counts) <- rownames(z$genes) <- z$genes$genes

# TMM Normalisation
z <- calcNormFactors(z)
z$samples

# design matrix
Group <- z$samples$group
Tissue <- z$samples$tissue
data.frame(Sample = colnames(z), Group, Tissue)

design <- model.matrix(~Tissue+Group)
rownames(design) <- colnames(z)
design

fcov.log.base.counts <- cpm(z, prior.count = 1, log = TRUE, normalized.lib.sizes = TRUE)

fcov.df <- as.data.frame(fcov.log.base.counts)
fcov.df$gene <- rownames(fcov.df)

fcov.df <- fcov.df[fcov.df$gene == 'c1je_genome',]
fcov.df$gene <- NULL

fcov.melt <- melt(fcov.df)
fcov.melt$group <- z$samples$group
fcov.melt$tissue <- z$sample$tissue

# create matrix for heatmap 
fcov.hmap <- fcov.melt
rownames(fcov.hmap) <- fcov.hmap$variable
fcov.hmap$variable <- NULL
fcov.hmap$group <- NULL
fcov.hmap$cat <- stringr::str_extract(rownames(fcov.hmap), "[^_]*_[^_]*")
# export
write.table(fcov.hmap, file = 'fcov_per_sample.txt', sep = '\t', quote = FALSE)

fcov.mat <- acast(fcov.hmap, cat ~ tissue)

# set row order
mat.col.order <- c('control_54', 'control_69', 'control_135', 'control_136', 'control_164', 'control_174', 'control_188', 'fip_37', 'fip_68', 'fip_86', 'fip_96', 'fip_120', 'fip_172', 'fip_142', 'fip_192')
fcov.mat <- fcov.mat[mat.col.order,]

# plot
library(pheatmap)
library(RColorBrewer)
library(viridis)

breaks = seq(-7, 12, 0.1)
hmcols = colorRampPalette(brewer.pal(9, 'OrRd'))(length(breaks))

pheatmap(fcov.mat,
         breaks = breaks,
         color = hmcols,
         na_col = 'darkgrey',
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         angle_col = 45,
         cellheight = 20,
         cellwidth = 20,
         gaps_row = 7)
```

## FCoV abundance vs gene set scores

```{r fcov_vs_genesetscores}
library(ggplot2)

# read in gene set scores
scoredf.ifn <- read.csv('interferon_gene_set_scores.txt', sep = '\t', header = TRUE)
scoredf.cytokine <- read.csv('cytokines_chemokines_gene_set_scores.txt', sep = '\t', header = TRUE)
scoredf.adaptive <- read.csv('adaptive_immune_response_gene_set_scores.txt', sep = '\t', header = TRUE)

# interferon stimulated genes
scoredf.ifn$fcov <- fcov.hmap$value

ggplot(scoredf.ifn, aes(fcov, TotalScore)) +
  geom_smooth(method = 'lm', color = 'black') +
  geom_point(aes(fill = tissue, shape = group), size = 2.5) +
  scale_shape_manual(values = c(21, 23)) +
  scale_fill_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
  labs(x = expression('FCoV '*log[2]*' CPM'), y = 'Interferon alpha response score', fill = '', shape = '') +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  theme_bw() +
  theme(panel.grid = element_blank())

summary(lm(fcov ~ TotalScore, data = scoredf.ifn))
cor.test(scoredf.ifn$fcov, scoredf.ifn$TotalScore)

# cytokines/chemokines
scoredf.cytokine$fcov <- fcov.hmap$value

ggplot(scoredf.cytokine, aes(fcov, TotalScore)) +
  geom_smooth(method = 'lm', color = 'black') +
  geom_point(aes(fill = tissue, shape = group), size = 2.5) +
  scale_shape_manual(values = c(21, 23)) +
  scale_fill_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
  labs(x = expression('FCoV '*log[2]*' CPM'), y = 'Cytokine/chemokine score', fill = '', shape = '') +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  theme_bw() +
  theme(panel.grid = element_blank())

summary(lm(fcov ~ TotalScore, data = scoredf.cytokine))
cor.test(scoredf.cytokine$fcov, scoredf.cytokine$TotalScore)

# adaptive immune response
scoredf.adaptive$fcov <- fcov.hmap$value

ggplot(scoredf.adaptive, aes(fcov, TotalScore)) +
  geom_smooth(method = 'lm', color = 'black') +
  geom_point(aes(fill = tissue, shape = group), size = 2.5) +
  scale_shape_manual(values = c(21, 23)) +
  scale_fill_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
  labs(x = expression('FCoV '*log[2]*' CPM'), y = 'Adaptive immune response score', fill = '', shape = '') +
  guides(fill = guide_legend(override.aes = list(shape = 21))) +
  theme_bw() +
  theme(panel.grid = element_blank())

summary(lm(fcov ~ TotalScore, data = scoredf.adaptive))
cor.test(scoredf.adaptive$fcov, scoredf.adaptive$TotalScore)
```

## FCoV genome assembly

Samtools v1.21.

Extract unmapped reads following alignment to felCat9.

```{r extract_unmapped_reads_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash
#SBATCH -J extract_unmapped_reads.job 
#SBATCH -o extract_unmapped_reads.%A_%a.out 
#SBATCH -c 2

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
outpath=unmapped

bams=()
for bam in *.bam
do
        bam=${bam%.bam}
        bams[${#bams[@]}]=$bam
done

# define variables
infile=$path/${bams[$i]}.bam
outfile1=$outpath/${bams[$i]}.unmapped.bam
outfile2=$outpath/${bams[$i]}.unmapped.sorted.bam
fq1=$path/${bams[$i]}.1.fq
fq2=$path/${bams[$i]}.2.fq

samtools view -@ 2 -b -f 4 $infile > $outfile1
samtools sort -n -@ 2 $outfile1 -o $outfile2
bamToFastq -i $outfile2 -fq $fq1 -fq2 $fq2
```

```{r extract_unmapped_reads_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
bams=()
for bam in *.bam
do
        bam=${bam%.bam}
        bams[${#bams[@]}]=$bam
done
bamcount=${#bams[@]}

sbatch --array 1-$bamcount:1 extract_unmapped_reads_secondary.sh
```

coronaSPAdes v3.15.4.

Assemble unmapped reads using coronaSPAdes with default settings. 

```{r coronaspades_assemble_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash
#SBATCH -J coronaspades_assemble.job 
#SBATCH -o coronaspades_assemble.%A_%a.out 
#SBATCH -c 2

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
inpath=fastq
outpath=coronaspades

cd $inpath

fqs=()
for fq in *.merged.unmapped.1.fastq
do
        fq=${fq%.merged.unmapped.1.fastq}
        fqs[${#fqs[@]}]=$fq
done

cd $path

#define variables
fq1=$inpath/${fqs[$i]}.merged.unmapped.1.fastq
fq2=$inpath/${fqs[$i]}.merged.unmapped.2.fastq
outfile=$outpath/${fqs[$i]}

coronaspades.py -o $outfile -1 $fq1 -2 $fq2
```

```{r coronaspades_assemble_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
inpath=fastq

cd $inpath

fqs=()
for fq in *.merged.unmapped.1.fastq
do
        fq=${fq%.merged.unmapped.1.fastq}
        fqs[${#fqs[@]}]=$fq
done
fqcount=${#fqs[@]}

cd $path

sbatch --array 1-$fqcount:1 coronaspades_assemble_secondary.sh
```

Split assemblies by scaffold using UCSC `faSplit` (installed via conda).

```{r coronaspades_split_by_scaffold_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash
#SBATCH -J coronaspades_split_by_scaffold.job 
#SBATCH -o coronaspades_split_by_scaffold.%A_%a.out 
#SBATCH -c 1

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`

dirs=()
for dir in *
do
        dir=${dir}
        dirs[${#dirs[@]}]=$dir
done

#define variables
fasta=$path/${dirs[$i]}/scaffolds.fasta
outpath=$path/${dirs[$i]}/split_by_scaffold/

#function
mkdir $outpath
faSplit byname $fasta $outpath
```

```{r coronaspades_split_by_scaffold_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
dirs=()
for dir in *
do
        dir=${dir}
        dirs[${#dirs[@]}]=$dir
done
dircount=${#dirs[@]}

sbatch --array 1-$dircount:1 coronaspades_split_by_scaffold_secondary.sh
```

Tabulate longest scaffold lengths for all samples and plot. 

```{r coronaspades_longest_scaffolds}
library(tidyr)
library(dplyr)
library(ggplot2)

# read in data
assembled.stats <- read.csv('coronaspades_longest_scaffolds.txt', sep = '\t', header = TRUE, row.names = NULL)
colnames(assembled.stats)
# [1] "Sample"                    "Group"                    
# [3] "Tissue"                    "Longest.scaffold"

# order within groups
assembled.stats.grpranked <- assembled.stats %>% arrange(Group, desc(Longest.scaffold))
assembled.stats.grpranked$Sample <- factor(assembled.stats.grpranked$Sample, levels = assembled.stats.grpranked$Sample)

ggplot(assembled.stats.grpranked, aes(Sample, Longest.scaffold, fill = Tissue)) +
  geom_bar(stat = 'identity') +
  scale_fill_manual(values = c('#EA8131', '#5EC64F', '#033BF4')) +
  scale_y_continuous(expand = c(0,0), limits = c(0,30000), breaks = seq(0,30000,5000)) +
  facet_wrap(vars(Group), scales = 'free') +
  labs(x = '', y = 'Scaffold length (nt)') +
  theme_bw() +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6),
        panel.grid = element_blank())
```

## Phylogenetic tree

Full-length FCoV genome sequences were retrieved from NCBI Virus and aligned with coronaSPAdes-assembled FCoV genomes from this study. Sequences were aligned using MUSCLE (v3.8.31) in AliView (v1.28). Maximum-likelihood (ML) trees were generated from the aligned sequences using RAxML with the GTRGAMMAIX model (selected using [modeltest](https://github.com/ddarriba/modeltest)). 

```{r fcov_tree_raxml, engine = 'bash', eval = FALSE}
raxmlHPC-PTHREADS -T 2 -m GTRGAMMAIX -s fig3_fipv_fecv_fcov2_fcovcandidates_ferretcovoutgroup_polyatrimmed_muscle.fa -n fig3_fipv_fecv_fcov2_fcovcandidates_ferretcovoutgroup_polyatrimmed_muscle.ML -o NC_030292.1 -p 11334 -k -f a -x 13243 -N 100 
```

Import best-scoring ML tree with bootstrap support and plot with annotation data.

```{r fcov_tree_plot}
library(ggtree)

# load tree data 
fcov.tree <- read.tree('RAxML_bipartitions.fig3_fipv_fecv_fcov2_fcovcandidates_ferretcovoutgroup_polyatrimmed_muscle.ML')

# remove all but sample id/group from names
fcov.tree$tip.label <- gsub('feline_', '', fcov.tree$tip.label)
fcov.tree$tip.label <- gsub('_node_1.*', '', fcov.tree$tip.label)
# add annotation data
fcov.tree.anno <- read.csv(file = 'fcov_tree_anno.txt', sep = '\t', header = TRUE)
colnames(fcov.tree.anno)
# [1] "Sequence" "Biotype"  "Serotype"

# build tree with annotation data
p <- ggtree(fcov.tree, layout = 'rectangular', ladderize = TRUE) +
  geom_treescale() +
  # geom_nodepoint(aes(color = as.numeric(label))) +
  # geom_nodelab(size = 1.5, nudge_x = 0.02, node = 'internal') +
  geom_tiplab(align = FALSE, as_ylab = TRUE, colour = 'black') 

p2 <- p %<+% fcov.tree.anno +
  geom_tippoint(aes(color = Biotype)) +
  scale_color_manual(values = c('#32A7CD', '#8F3092', 'grey')) +
  geom_point2(aes(label = label, subset = !is.na(as.numeric(label)) & as.numeric(label) > 80), shape = 23, fill = 'grey')
```

## Multiple sequence alignments

Open reading frames (ORFs) were predicted in the assembled FCoV genomes using Prodigal (v2.6.3), and predicted Spike proteins were aligned with Spike proteins from FECV sequences available in NCBI Virus using Clustal Omega (https://www.ebi.ac.uk/jdispatcher/msa/clustalo).

Visualise Spike protein alignments at key sites. 

```{r spike_alignment}
library(ggmsa)
library(Biostrings)

spike.align <- readAAMultipleAlignment(filepath = 'fcov_public_and_candidates_spike_clustal.aln',
                                       format = 'clustal')

# M1058L
ggmsa(spike.align, 1047, 1059, font = "mono", char_width = 0.5, seq_name = TRUE, border = NA, color = 'LETTER')

# S1/S2
ggmsa(spike.align, 786, 800, font = "mono", char_width = 0.5, seq_name = TRUE, border = NA, color = 'LETTER')
```

Visualise alignment of predicted ORF3c protein sequences. 

```{r orf3c_alignment}
library(ggmsa)
library(Biostrings)
library(aplot)

orf3c.nt.align <- readDNAMultipleAlignment(filepath = 'all_fcov_candidates_orf3c_nt_clustal.fa', format = 'fasta')

gg.ntmsa <- ggmsa(
  orf3c.nt.align,
  font = NULL,
  color = 'Chemistry_NT',
  seq_name = TRUE,
  border = 'white',
  show.legend = TRUE
) 

orf3c.aa.trunc.align <- readAAMultipleAlignment(filepath = 'all_fcov_candidates_orf3c_aa_clustal.fa', format = 'fasta')

gg.aamsa <- ggmsa(
  orf3c.aa.trunc.align,
  font = NULL,
  color = 'LETTER',
  seq_name = TRUE,
  border = 'white',
  show.legend = TRUE
) 

aplot::plot_list(gg.ntmsa, gg.aamsa, ncol = 1)
```

Visualise alignment of Spike domain 0 deletions.

```{r spike_domain0del_alignment}
library(ggmsa)
library(Biostrings)

domain0.nt.align <- readDNAMultipleAlignment(filepath = 'spike_domain0_nt_clustal.msa', format = 'fasta')

gg.sntmsa <- ggmsa(
  domain0.nt.align,
  font = NULL,
  color = 'Chemistry_NT',
  seq_name = TRUE,
  border = 'white',
  show.legend = TRUE
)
```

## FCoV C1Je genome coverage

Align the (unmapped) RNA-seq reads used for FCoV assembly with coronaSPAdes to the FCoV C1Je (DQ848678.1) genome using BWA-MEM and calculate depth at each position using Samtools. 

```{r c1je_align_unmapped_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash
#SBATCH -J c1je_align_unmapped.job 
#SBATCH -o c1je_align_unmapped.%A_%a.out 
#SBATCH -c 4

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
inpath=fastq
outpath=bam

cd $inpath

fqs=()
for fq in *1.fastq
do
        fq=${fq%.merged.unmapped.1.fastq}
        fqs[${#fqs[@]}]=$fq
done

cd $path

#define variables
fq1=$inpath/${fqs[$i]}.unmapped.1.fastq
fq2=$inpath/${fqs[$i]}.unmapped.2.fastq

refpath=DQ848678_c1je.fa

outfile=$outpath/${fqs[$i]}.unmapped.c1je
sortfile=$outpath/${fqs[$i]}.unmapped.c1je_sorted.bam

#function
bwa mem -t 4 -o $outfile.sam -a $refpath $fq1 $fq2
samtools sort -@ 4 -o $sortfile $outfile.sam
samtools index $sortfile
```

```{r c1je_align_unmapped_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
inpath=fastq

cd $inpath

fqs=()
for fq in *1.fastq
do
        fq=${fq%.merged.unmapped.1.fastq}
        fqs[${#fqs[@]}]=$fq
done
fqcount=${#fqs[@]}

cd $path

sbatch --array 1-$fqcount:1 c1je_align_unmapped_secondary.sh
```

Get read depth at each position in DQ848678.1 using `samtools depth`.

```{r c1je_depth_calc, engine = 'bash', eval = FALSE}
cat *c1je_sorted.bam > files.txt

samtools depth -a -H -f files.txt > all_samples_depth_c1je.txt
```

Read data into R and plot.

```{r c1je_depth_plot}
library(ggplot2)
library(reshape2)
library(viridis)

# read in depth
c1je.depth <- read.csv('all_samples_depth_c1je.txt', sep = '\t', header = T)
# melt
c1je.depth.melt <- melt(c1je.depth, id.vars = c('CHROM', 'POS'))
# set colours
cat.pal = viridis(n = 5, option = 'turbo')
# plot
ggplot(c1je.depth.melt, aes(POS, value, color = variable)) +
  geom_line() +
  scale_x_continuous(expand = c(0,0), limits = c(0, 29250), breaks = seq(0,30000, 5000)) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 1e6), breaks = c(0, 10, 100, 1000, 10000, 100000, 1e6), trans=scales::pseudo_log_trans(base = 10)) +
  scale_color_manual(values = cat.pal) +
  labs(x = '', y = 'Aligned read depth', color = '') +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## S gene percent identity

Align S gene sequences using e.g. Clustal Omega and calculate percent identity matrix.

```{r s_pid_matrix}
library(pheatmap)
library(viridis)
# hmcols = colorRampPalette(c("blue", "white"))(200)

breaks = seq(80, 100, 0.2)

# read in pid matrix
pim.spike <- read.csv('s_pid_matrix.txt', sep = '\t', header = TRUE, row.names = 1)

pheatmap(pim.spike,
         breaks = breaks,
         angle_col = 45,
         cellwidth = 25,
         cellheight = 25)
```

## Aligning reads to FCoV assemblies from MLN samples

For each cat, map reads used for FCoV assembly from all tissues to the FCoV assembly from MLN reads for analysis of intrahost single nucleotide variants (iSNVs). 

```{r mln_align_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash
#SBATCH -J mln_align.job 
#SBATCH -o mln_align.%A_%a.out 
#SBATCH -c 4

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
inpath=fastq
refpath=feline_fip_86_mln_bwa_index
outpath=mln_aln/86_mln

cd $inpath

fqs=()
for fq in *86*.1.fastq
do
        fq=${fq%.1.fastq}
        fqs[${#fqs[@]}]=$fq
done

cd $path

#find
fq1=$inpath/${fqs[$i]}.1.fastq
fq2=$inpath/${fqs[$i]}.2.fastq

#define variables
outfile=$outpath/${fqs[$i]}.mappedto86mln

#function
bwa mem -t 4 -o $outfile.sam -a $refpath $fq1 $fq2
samtools sort -@ 4 -o $outfile.sorted.bam $outfile.sam
samtools index $outfile.sorted.bam

# adapt for each cat
```

```{r mln_align_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
inpath=fastq

cd $inpath

fqs=()
for fq in *86*.1.fastq
do
        fq=${fq%.1.fastq}
        fqs[${#fqs[@]}]=$fq
done
fqcount=${#fqs[@]}

cd $path

sbatch --array 1-$fqcount:1 mln_align_secondary.sh
```

Calculate mean depth for each sample using `samtools coverage`.

```{r mln_align_coverage_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash
#SBATCH -J mln_align_coverage.job 
#SBATCH -o mln_align_coverage.%A_%a.out 
#SBATCH -c 2

let i=$SLURM_ARRAY_TASK_ID-1

inpath=`pwd`

bams=()
for bam in *.bam
do
        bam=${bam%.bam}
        bams[${#bams[@]}]=$bam
done

infile=$inpath/${bams[$i]}.bam
outfile=$inpath/coverage/${bams[$i]}_coverage.txt

samtools coverage $infile -o $outfile
```

```{r mln_align_coverage_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
bams=()
for bam in *.bam
do
        bam=${bam%.bam}
        bams[${#bams[@]}]=$bam
done
bamcount=${#bams[@]}

sbatch --array 1-$bamcount:1 mln_align_coverage_secondary.sh
```

Read in data and plot.

```{r mln_align_coverage}
library(tidyverse)
library(stringr)
library(ggplot2)

# read in samtools coverage output
file.list <- list.files(path = 'mln_aln/coverage', full.names = TRUE)

cov.df <- file.list %>%
  map_df(read.table)

cov.names <- str_remove(list.files(path = 'mln_aln/coverage'), '.unmapped.*')
cov.names <- str_remove(cov.names, '.merged.*')
cov.names <- str_remove(cov.names, 'feline_')

colnames(cov.df) <- c(
  'rname',
  'startpos',
  'endpos',
  'numreads',
  'covbases',
  'coverage',
  'meandepth',
  'meanbaseq',
  'meanmapq'
)
rownames(cov.df) <- cov.names
cov.df$rname <- NULL
cov.df$cat <- as.character(c(120, 120, 172, 172, 172, 68, 68, 68, 86, 86, 86, 96, 96))
cov.df$tissue <- ifelse(grepl('mln', rownames(cov.df)), 'MLN', ifelse(grepl('liver', rownames(cov.df)), 'Liver', 'Lung'))

write.table(cov.df, file = 'mln_mapped_coverage.txt', sep = '\t')
# missing samples manually added for plotting, i.e. zeros in place of missing assemblies/tissues
cov.df.plot <- read.csv(file = 'mln_mapped_coverage_with_missing_tissues.txt', sep = '\t', header = TRUE)

cov.df.plot$cat <- factor(cov.df.plot$cat, levels = c(68, 86, 96, 120, 172))

# plot bar chart
tissue.pal <- c('#EA8131', '#5EC64F', '#033BF4')

ggplot(cov.df.plot, aes(cat, log10(meandepth + 1), fill = tissue)) +
  geom_col(position = position_dodge(preserve = 'single')) +
  # geom_hline(yintercept = 2, lty = 'dashed', colour = 'black') +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 5)) +
  scale_fill_manual(values = tissue.pal) +
  facet_grid( ~ cat, scales = 'free_x') +
  labs(x = '',
       y = expression(log[10] * ' mean coverage'),
       fill = 'Tissue') +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank()
  )
```

## Call iSNVs with LoFreq

Using LoFreq (v2.1.5), call variants in viral reads for each sample relative to the MLN assembly for the same cat.

First, add indel qualities.

```{r lofreq_dindel_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J lofreq_dindel.job 
#SBATCH -o lofreq_dindel.%A_%a.out 
#SBATCH -c 8

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
refpath=fasta
bampath=bams
outpath=dindel

cd $refpath

fnames=()
for fname in *mln.fa
do
        fname=${fname%_mln.fa}
        fnames[${#fnames[@]}]=$fname
done

cd $path

reffile=$refpath/${fnames[$i]}_mln.fa

# create input file variables if the file exists, otherwise leave them empty
if [ -e "$bampath/${fnames[$i]}_mln.bam" ]
then
	mlnfile=$bampath/${fnames[$i]}_mln.bam
else
	mlnfile=""
fi

if [ -e "$bampath/${fnames[$i]}_liver.bam" ]
then
	liverfile=$bampath/${fnames[$i]}_liver.bam
else
	liverfile=""
fi

if [ -e "$bampath/${fnames[$i]}_lung.bam" ]
then
	lungfile=$bampath/${fnames[$i]}_lung.bam
else
	lungfile=""
fi

#function
lofreq indelqual --dindel -f $reffile -o $outpath/${fnames[$i]}_mln_dindel.bam $mlnfile
lofreq indelqual --dindel -f $reffile -o $outpath/${fnames[$i]}_liver_dindel.bam $liverfile
lofreq indelqual --dindel -f $reffile -o $outpath/${fnames[$i]}_lung_dindel.bam $lungfile
```

```{r lofreq_dindel_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
refpath=fasta

cd $refpath

fnames=()
for fname in *mln.fa
do
        fname=${fname%_mln.fa}
        fnames[${#fnames[@]}]=$fname
done
fnamecount=${#fnames[@]}

cd $path

sbatch --array 1-$fnamecount:1 lofreq_dindel_secondary.sh
```

Call variants.

```{r lofreq_call_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J lofreq_call.job 
#SBATCH -o lofreq_call.%A_%a.out 
#SBATCH -c 8

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
refpath=fasta
bampath=dindel
outpath=vcf

cd $refpath

fnames=()
for fname in *mln.fa
do
        fname=${fname%_mln.fa}
        fnames[${#fnames[@]}]=$fname
done

cd $path

reffile=$refpath/${fnames[$i]}_mln.fa

# create input file variables if the file exists, otherwise leave them empty
if [ -e "$bampath/${fnames[$i]}_mln_dindel.bam" ]
then
	mlnfile=$bampath/${fnames[$i]}_mln_dindel.bam
else
	mlnfile=""
fi

if [ -e "$bampath/${fnames[$i]}_liver_dindel.bam" ]
then
	liverfile=$bampath/${fnames[$i]}_liver_dindel.bam
else
	liverfile=""
fi

if [ -e "$bampath/${fnames[$i]}_lung_dindel.bam" ]
then
	lungfile=$bampath/${fnames[$i]}_lung_dindel.bam
else
	lungfile=""
fi

#function
lofreq call --call-indels -f $reffile -o ${fnames[$i]}_mln.vcf $mlnfile
lofreq call --call-indels -f $reffile -o ${fnames[$i]}_liver.vcf $liverfile
lofreq call --call-indels -f $reffile -o ${fnames[$i]}_lung.vcf $lungfile
```

```{r lofreq_call_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
refpath=fasta

cd $refpath

fnames=()
for fname in *mln.fa
do
        fname=${fname%_mln.fa}
        fnames[${#fnames[@]}]=$fname
done
fnamecount=${#fnames[@]}

cd $path

sbatch --array 1-$fnamecount:1 lofreq_call_secondary.sh
```

Annotate variants using SnpEff (v5.2). Custom databases were built for each MLN FCoV genome using the whole genome FASTA and GFF output from ORF prediction with Prodigal. 

```{r snpeff_annotate_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J snpeff_annotate.job 
#SBATCH -o snpeff_annotate.%A_%a.out 
#SBATCH -c 1

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
refpath=fasta
vcfpath=vcf
outpath=snpeff

cd $refpath

fnames=()
for fname in *mln.fa
do
        fname=${fname%_mln.fa}
        fnames[${#fnames[@]}]=$fname
done

cd $path

refname=${fnames[$i]}_mln

# create input file variables if the file exists, otherwise leave them empty
if [ -e "$vcfpath/${fnames[$i]}_mln.vcf" ]
then
	mlnfile=$vcfpath/${fnames[$i]}_mln.vcf
else
	mlnfile=""
fi

if [ -e "$vcfpath/${fnames[$i]}_liver.vcf" ]
then
	liverfile=$vcfpath/${fnames[$i]}_liver.vcf
else
	liverfile=""
fi

if [ -e "$vcfpath/${fnames[$i]}_lung.vcf" ]
then
	lungfile=$vcfpath/${fnames[$i]}_lung.vcf
else
	lungfile=""
fi

outmln=snpeff/${fnames[$i]}_mln_snpeff.vcf
outliver=snpeff/${fnames[$i]}_liver_snpeff.vcf
outlung=snpeff/${fnames[$i]}_lung_snpeff.vcf

#function
snpEff -no-downstream -no-intergenic -no-upstream -no-utr -v $refname $mlnfile > $outmln
snpEff -no-downstream -no-intergenic -no-upstream -no-utr -v $refname $liverile > $outliver
snpEff -no-downstream -no-intergenic -no-upstream -no-utr -v $refname $lungfile > $outlung
```

```{r snpeff_annotate_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
refpath=fasta

cd $refpath

fnames=()
for fname in *mln.fa
do
        fname=${fname%_mln.fa}
        fnames[${#fnames[@]}]=$fname
done
fnamecount=${#fnames[@]}

cd $path

sbatch --array 1-$fnamecount:1 snpeff_annotate_secondary.sh
```

Export variants with a single variant per line as tab-delimited text for reading into R.

```{r snpsift_export_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J snpsift_export.job 
#SBATCH -o snpsift_export.%A_%a.out 
#SBATCH -c 1

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`

fnames=()
for fname in *.vcf
do
        fname=${fname%.vcf}
        fnames[${#fnames[@]}]=$fname
done

infile=$path/${fnames[$i]}.vcf
outfile=$path/${fnames[$i]}.txt
snpeffpath=/snpeff-5.2-1/scripts

#function
cat $infile \
        | $snpeffpath/vcfEffOnePerLine.pl \
        | SnpSift extractFields -e "." - "CHROM" "POS" "ID" "REF" "ALT" "FILTER" "DP" "AF" "DP4" "ANN[*].ALLELE" "ANN[*].EFFECT" "ANN[*].IMPACT" "ANN[*].GENE" "ANN[*].BIOTYPE" "ANN[*].HGVS_C" "ANN[*].HGVS_P" "ANN[*].AA_POS" > $outfile
```

```{r snpsift_export_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
fnames=()
for fname in *.vcf
do
        fname=${fname%.vcf}
        fnames[${#fnames[@]}]=$fname
done
fnamecount=${#fnames[@]}

sbatch --array 1-$fnamecount:1 snpsift_export_secondary.sh
```

## Analysis of iSNVs

Read in annotated and re-formatted variants called with LoFreq. Apply preliminary filters (AAF < 0.02, ALT reads < 5).

```{r lofreq_filter}
library(tidyverse)
library(magrittr)

# read in variants
mln.vars.list <- list.files(path = 'mln_aln/lofreq',
                            pattern = '*\\.txt',
                            full.names = TRUE)
mln.vars.names <- list.files(path = 'mln_aln/lofreq',
                             pattern = '*_snpeff.txt',
                             full.names = FALSE)
mln.vars.names <- stringr::str_remove(mln.vars.names, '_snpeff.txt')

vars1 <- vector("list", length(mln.vars.list))
for (i in seq_along(mln.vars.list)) {
  y <- suppressMessages(readr::read_delim(mln.vars.list[i], delim = '\t', col_names = TRUE))
  vars1[[i]] <- y
  vars1[[i]]$SAMPLE <- mln.vars.names[i]
}
names(vars1) <- mln.vars.names

# add tissue and cat id for splitting
id <- stringr::str_remove(mln.vars.names, 'feline_fip_')
tissue <- stringr::str_remove(id, '[0-9].*_')
tissue <- sapply(tissue, function(x) {
  if (x == 'mln') {
    stringr::str_to_upper(x)
  } else {
    stringr::str_to_title(x)
  }
})
cat <- stringr::str_remove(id, '_.*')

# removed fip_96_liver - no variants

vars1 <- mapply(cbind,
                vars1,
                "cat" = cat,
                "tissue" = tissue,
                SIMPLIFY = FALSE)

# merge variants
vars2 <- Reduce(function(...)
  merge(..., all = TRUE), vars1)

# split DP4 column into separate fields - ref-forward bases, ref-reverse, alt-forward and alt-reverse
vars2[c('DP_REF_FWD', 'DP_REF_REV', 'DP_ALT_FWD', 'DP_ALT_REV')] <- stringr::str_split_fixed(vars2$DP4, ';', 4)

# filter variants in 5' and 3' UTRs
vars3 <- vars2[vars2$POS < 28968 & vars2$POS > 260, ]

# filter variants < 2%
vars3 <- vars3[vars3$AF >= 0.02, ]
table(vars3$cat, vars3$tissue)

# filter variants with < 5 ALT reads
vars3$DP_ALT <- as.numeric(vars3$DP_ALT_FWD) + as.numeric(vars3$DP_ALT_REV)
vars3 <- vars3[vars3$DP_ALT >= 5, ]
table(vars3$cat, vars3$tissue)

# triallelic sites
triallelic <- vars3 %>%
  group_by(SAMPLE, POS) %>%
  filter(n() > 1) %>%
  ungroup()
```

Stacked bar plots of annotated variant effects.

```{r lofreq_effects}
library(ggplot2)

# subset SNVs
snvs <- vars3[vars3$`ANN[*].EFFECT` %in% c('missense_variant', 'synonymous_variant', 'stop_gained'),]

# get number of SNVs in each gene in each sample
snvcount <- as.data.frame(table(snvs$cat, snvs$tissue, snvs$`ANN[*].GENE`, snvs$`ANN[*].EFFECT`))
colnames(snvcount) <- c('cat', 'tissue', 'gene', 'anno', 'count')
snvcount$gene <- factor(snvcount$gene, levels = c('ORF1a', 'ORF1b', 'S', '3a', '3b', '3c', 'E', 'M', 'N', '7a', '7b'))

# normalise SNV counts by ORF lengths
# write.table(snvcount, file = 'snv_count.txt', sep = '\t')

snvcount.norm <- read.csv(file = 'snv_count_lengthnorm.txt', sep = '\t', header = TRUE)
snvcount.norm$gene <- factor(snvcount.norm$gene, levels = c('ORF1a', 'ORF1b', 'S', '3a', '3b', '3c', 'E', 'M', 'N', '7a', '7b'))
snvcount.norm$tissue <- factor(snvcount.norm$tissue, levels = c('MLN', 'Liver', 'Lung'))

# remove cat96
snvcount.norm <- snvcount.norm[snvcount.norm$cat != 96,]

consequence.pal <- c('#35b779', '#fde725', '#31688e')

ggplot(snvcount.norm, aes(x = tissue, y = normcount, fill = anno, group = anno)) +
  geom_bar(stat = 'identity', position = 'stack') + 
  scale_fill_manual(values = consequence.pal) +
  labs(x = '', y = 'SNVs per kb', fill = '') +
  facet_grid(cat ~ gene) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(panel.grid = element_blank())
```

Plot mean depth vs no. SNVs for each sample. 

```{r lofreq_depth_vs_nosnvs}
nosnvs <- as.data.frame(table(snvs$cat, snvs$tissue))
nosnvs <- nosnvs[order(nosnvs$Var1),]

# store coverage stats from earlier and combine with snv counts
snvstats <- cov.df
snvstats$snvs <- nosnvs$Freq[-c(2,14)]

library(ggplot2)
library(ggsci)
library(viridis)

cat.pal = viridis(n = 5, option = 'turbo')

snv.stats.gg <- ggplot(snvstats, aes(log10(meandepth), snvs)) +
  # geom_smooth() +
  geom_point(aes(colour = cat, shape = tissue), size = 2.5) +
  scale_color_manual(values = cat.pal) +
  # geom_vline(xintercept = log10(500), color = 'black', lty = 'dashed') +
  labs(x = expression(log[10]*' mean depth'), y = 'Number of SNVs', color = 'Cat', shape = 'Tissue') + 
  theme_bw() +
  theme(panel.grid = element_blank())
```

Plot distribution of allele frequencies across tissues.

```{r lofreq_snv_freqdist}
library(ggplot2)

# remove 37 and 96 (not enough snvs)
vars4 <- droplevels(vars3[vars3$cat %in% c("68", "86", "120", "172"),])
vars4$cat <- factor(vars4$cat, levels = c('120', '172', '68', '86'))
vars4$tissue <- factor(vars4$tissue, levels = c('MLN', 'Liver', 'Lung'))

ggplot(vars4, aes(AF, fill = tissue)) +
  geom_histogram(bins = 60) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0), limits = c(0,125)) +
  facet_wrap(vars(cat), nrow = 1) +
  scale_fill_manual(values = tissue.pal[c(3,1,2)]) +
  labs(x = 'AAF', y = 'No. variants', fill = 'Tissue') +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Stability of variant calls - downsampling

Downsampled the 4 samples with highest mean depth (MLN from 68, 86, 120 and 172) to fixed mean depths (2000x - except 120, 1000x, 500x, 200x, 100x, 50x) using `samtools view` with `-s` and the value for [target_downsampled_depth / mean_depth] and called variants using lofreq with the same settings as for the main analysis. After filtering, see how many of the variants called with all reads are retained at each depth. 

```{r lofreq_downsampling}
library(ggplot2)

# read in data
ds.vars.list <- list.files(path = 'mln_aln/lofreq/downsampled', 
                        pattern = '*\\.txt',
                        full.names = TRUE)
ds.vars.names <- list.files(path = 'mln_aln/lofreq/downsampled', 
                        pattern = '*\\.txt',
                        full.names = FALSE)
ds.vars.names <- stringr::str_remove(ds.vars.names, '.txt')

dsvars1 <- vector("list", length(ds.vars.list))
for(i in seq_along(ds.vars.list)) {
  y <- suppressMessages(readr::read_delim(ds.vars.list[i], delim = '\t', col_names = TRUE))
  dsvars1[[i]] <- y
  dsvars1[[i]]$SAMPLE <- ds.vars.names[i]
}
names(dsvars1) <- ds.vars.names

# add depth and cat id for splitting
dsid <- stringr::str_remove(ds.vars.names, 'feline_fip_')
dsid <- stringr::str_remove(dsid, '_snpeff')
depth <- stringr::str_remove(dsid, '[0-9].*_mln.')
dscat <- stringr::str_remove(dsid, '_.*')

dsvars1 <- mapply(cbind, dsvars1, "cat"=dscat, "depth"=depth, SIMPLIFY = FALSE)

# merge variants
dsvars2 <- Reduce(function(...) merge(..., all = TRUE), dsvars1)

# split DP4 column into separate fields - ref-forward bases, ref-reverse, alt-forward and alt-reverse
dsvars2[c('DP_REF_FWD', 'DP_REF_REV', 'DP_ALT_FWD', 'DP_ALT_REV')] <- stringr::str_split_fixed(dsvars2$DP4, ';', 4)

# filter variants in 5' and 3' UTRs
dsvars3 <- dsvars2[dsvars2$POS < 28968 & dsvars2$POS > 260,]

# filter variants < 2%
dsvars3 <- dsvars3[dsvars3$AF >= 0.02,]
table(dsvars3$cat, dsvars3$depth)

# filter variants with < 5 ALT reads
dsvars3$DP_ALT <- as.numeric(dsvars3$DP_ALT_FWD) + as.numeric(dsvars3$DP_ALT_REV)
dsvars3 <- dsvars3[dsvars3$DP_ALT >= 5,]
ds.snv.tab <- table(dsvars3$cat, dsvars3$depth)

# divide downsampled snv counts by no. variants with all reads
norm.dsvars <- data.frame(rbind(ds.snv.tab / c(87, 491, 154, 81)))
norm.dsvars <- norm.dsvars[,c('X2000x', 'X1000x', 'X500x', 'X200x', 'X100x', 'X50x')]
norm.dsvars$Cat <- rownames(norm.dsvars)
colnames(norm.dsvars) <- c('2000x', '1000x', '500x', '200x', '100x', '50x', 'Cat')
# set 0 to na
norm.dsvars[norm.dsvars == 0] <- NA

norm.dsvars.melt <- reshape2::melt(norm.dsvars, id = 'Cat')

ds.box.gg <- ggplot(norm.dsvars.melt, aes(x = variable, y = value)) +
  geom_boxplot(fill = NA, color = 'darkgrey', outlier.shape = NA) +
  geom_point(aes(color = Cat), position = position_jitter(width = 0.1, height = 0), size = 2) +
  geom_hline(yintercept = 1, color = 'black', lty = 'dashed') +
  scale_y_continuous(expand = c(0,0), limits = c(0,2), breaks = seq(0,2,0.5)) +
  scale_color_manual(values = cat.pal[1:4]) +
  labs(x = 'Mean depth after downsampling', y = 'Variants identified / variants at max depth') +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Mutation spectrum

```{r lofreq_mutation_spectrum}
library(ggplot2)

# look at types of mutations and their frequencies
snvs <- vars3[nchar(vars3$REF) == 1 & nchar(vars3$ALT) == 1,]
snvs$mutation <- paste(snvs$REF,'>', snvs$ALT, sep = '')

# remove 37 and 96 (not enough snvs)
snvs <- droplevels(snvs[snvs$cat %in% c("68", "86", "120", "172"),])

# remove UTR variants
snvs <- snvs[!snvs$`ANN[*].GENE` == '.', ]

# summarise as proportion of variants for each type, and plot all cats together
# first, set mutation order
snvs$mutation <- factor(snvs$mutation, levels = c('C>A', 'G>T', 'C>G', 'G>C', 'C>T', 'G>A', 'T>A', 'A>T', 'T>C', 'A>G', 'T>G', 'A>C'))

snvs.summ <- snvs %>%
  group_by(cat, mutation) %>%
  summarise(n = n()) %>%
  complete(mutation = levels(snvs$mutation), fill = list(n = 0)) %>%
  mutate(freq = n/sum(n), .keep = 'all') 

snvs.summ$mutation <- factor(snvs.summ$mutation, levels = levels(snvs$mutation))

snvs.summ$group <- rep(c('T>G', 'T>C', 'T>A', 'C>A', 'C>G', 'C>T', 'C>T', 'C>G', 'C>A', 'T>A', 'T>C', 'T>G'), 4)
snvs.summ$strand <- rep(c(rep('rev', 3), rep('fwd', 3)), 8)

# cast to matrix and combine with mean data
snvs.cast <- reshape2::acast(snvs.summ, mutation~cat, value.var = 'freq')

snvs.mean <- snvs.summ %>%
  group_by(mutation) %>%
  summarise(meanfreq = mean(freq), sdfreq = sd(freq))

snvs.mean$group <- c(rep('C>A', 2), rep('C>G', 2), rep('C>T', 2), rep('T>A', 2), rep('T>C', 2), rep('T>G', 2))
snvs.mean$strand <- rep(c('fwd', 'rev'), 6)

snvs.comb <- cbind(snvs.mean, snvs.cast)

# melt for plotting
snvs.melt <- reshape2::melt(snvs.comb, id.vars = c('mutation', 'meanfreq', 'sdfreq', 'group', 'strand'))

ggplot(snvs.melt, aes(group, meanfreq, group = strand, color = variable)) +
  geom_bar(aes(fill = strand), stat = 'identity', position = 'dodge', color = NA) +
  geom_errorbar(aes(ymin = meanfreq-sdfreq, ymax = meanfreq+sdfreq), position = position_dodge(0.9), width = 0.2, color = 'black') +
  geom_point(aes(x = group, y = value), position = position_jitterdodge(dodge.width = 0.9, jitter.width = 0.1)) +
  scale_color_manual(values = cat.pal) +
  scale_fill_manual(values = c('lightgray', 'darkgrey')) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 0.3), breaks = seq(0,0.3,0.05)) +
  labs(x = 'Mutation type (plus strand)', y = 'Mutation frequency', color = 'Cat', fill = 'Strand') +
  theme_bw() +
  theme(panel.grid = element_blank())
```

## Selection analysis

Using the [dndscv](https://github.com/im3sanger/dndscv) package to get gene-level estimates of selection.  

```{r dndscv_selection}
# following the tutorial for building refs and estimating selection pressure - note the tutorial is for human data, but the tool has been used for SARS-CoV-2 data in this paper: https://elifesciences.org/articles/66857#s4
library(dndscv)

# start with 86_mln
# create CDS table to use as input (based on GFF from Prodigal)
# note - cds must be divisible by 3
# https://github.com/im3sanger/dndscv/issues/55
buildref(cdsfile='dndscv/tx_tables/fip_86_mln.txt', genomefile='dndscv/feline_fip_86_mln.fa', outfile = 'dndscv/refcds/fip_86_mln.rda', numcode = 11)

# inspect ref
load('dndscv/refcds/fip_86_mln.rda')
# looks fine

# create variant table for 86mln
mut.86mln <- subset(vars3[vars3$SAMPLE == 'feline_fip_86_mln',], select = c(SAMPLE, CHROM, POS, REF, ALT))
colnames(mut.86mln) <- c('sampleID', 'chr', 'pos', 'ref', 'mut')

# run dndscv
dndscv.86mln <- dndscv(mut.86mln, refdb = 'dndscv/refcds/fip_86_mln.rda', numcode = 11, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf, cv = NULL)

# now try all tissues for cat 86
mut.86 <- subset(vars3[vars3$CHROM == 'feline_fip_86_mln',], select = c(SAMPLE, CHROM, POS, REF, ALT))
colnames(mut.86) <- c('sampleID', 'chr', 'pos', 'ref', 'mut')

# filter shared variants
mut.86.dedup <- mut.86 %>%
  distinct(pos, mut, .keep_all = TRUE)

# run dndscv
dndscv.86 <- dndscv(mut.86.dedup, refdb = 'dndscv/refcds/fip_86_mln.rda', sm = '192r_3w', numcode = 11, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf, outmats = TRUE, constrain_wnon_wspl = TRUE)

# add missing line to mle_submodel
dndscv.86$mle_submodel[nrow(dndscv.86$mle_submodel)+1,] = list('wspl', 7.547607e-02, 0.0181692282, 0.31353214)

ci.86 <- geneci(dndscv.86)
ci.86$gene <- factor(ci.86$gene, levels = c('ORF1a', 'ORF1b', 'S', '3a', '3b', '3c', 'E', 'M', 'N', '7a', '7b'))

ggplot(ci.86, aes(gene, mis_mle)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(x = gene, ymin = mis_low, ymax = mis_high), width = 0.2) +
  theme_bw()

# now 68_mln
# create CDS table to use as input
# check that CDSs are divisible by 3
buildref(cdsfile='dndscv/tx_tables/fip_68_mln.txt', genomefile='dndscv/feline_fip_68_mln.fa', outfile = 'dndscv/refcds/fip_68_mln.rda', numcode = 11)

# now try all tissues for cat 68
mut.68 <- subset(vars3[vars3$CHROM == 'feline_fip_68_mln',], select = c(SAMPLE, CHROM, POS, REF, ALT))
colnames(mut.68) <- c('sampleID', 'chr', 'pos', 'ref', 'mut')

# filter shared variants
mut.68.dedup <- mut.68 %>%
  distinct(pos, mut, .keep_all = TRUE)

# run dndscv
dndscv.68 <- dndscv(mut.68.dedup, refdb = 'dndscv/refcds/fip_68_mln.rda', sm = '192r_3w', numcode = 11, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf, outmats = TRUE, constrain_wnon_wspl = TRUE)

# add missing line to mle_submodel
dndscv.68$mle_submodel[nrow(dndscv.68$mle_submodel)+1,] = list('wspl', 7.547607e-02, 0.0181692282, 0.31353214)

ci.68 <- geneci(dndscv.68)
ci.68$gene <- factor(ci.68$gene, levels = c('ORF1a', 'ORF1b', 'S', '3a', '3b', '3c', 'E', 'M', 'N', '7a', '7b'))

ggplot(ci.68, aes(gene, mis_mle)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(x = gene, ymin = mis_low, ymax = mis_high), width = 0.2) +
  theme_bw()

# 120_mln
# create CDS table to use as input
# check that CDSs are divisible by 3
buildref(cdsfile='dndscv/tx_tables/fip_120_mln.txt', genomefile='dndscv/feline_fip_120_mln.fa', outfile = 'dndscv/refcds/fip_120_mln.rda', numcode = 11)

# now try all tissues for cat 120
mut.120 <- subset(vars3[vars3$CHROM == 'feline_fip_120_mln',], select = c(SAMPLE, CHROM, POS, REF, ALT))
colnames(mut.120) <- c('sampleID', 'chr', 'pos', 'ref', 'mut')

# filter shared variants
mut.120.dedup <- mut.120 %>%
  distinct(pos, mut, .keep_all = TRUE)

# run dndscv
dndscv.120 <- dndscv(mut.120.dedup, refdb = 'dndscv/refcds/fip_120_mln.rda', sm = '192r_3w', numcode = 11, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf, outmats = TRUE, constrain_wnon_wspl = TRUE)

# add missing line to mle_submodel
dndscv.120$mle_submodel[nrow(dndscv.120$mle_submodel)+1,] = list('wspl', 7.547607e-02, 0.0181692282, 0.31353214)

ci.120 <- geneci(dndscv.120)
ci.120$gene <- factor(ci.120$gene, levels = c('ORF1a', 'ORF1b', 'S', '3a', '3b', '3c', 'E', 'M', 'N', '7a', '7b'))

ggplot(ci.120, aes(gene, mis_mle)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(x = gene, ymin = mis_low, ymax = mis_high), width = 0.2) +
  theme_bw()

# 172_mln
# create CDS table to use as input
# check that CDSs are divisible by 3
buildref(cdsfile='dndscv/tx_tables/fip_172_mln.txt', genomefile='dndscv/feline_fip_172_mln.fa', outfile = 'dndscv/refcds/fip_172_mln.rda', numcode = 11)

# now try all tissues for cat 172
mut.172 <- subset(vars3[vars3$CHROM == 'feline_fip_172_mln',], select = c(SAMPLE, CHROM, POS, REF, ALT))
colnames(mut.172) <- c('sampleID', 'chr', 'pos', 'ref', 'mut')

# filter shared variants
mut.172.dedup <- mut.172 %>%
  distinct(pos, mut, .keep_all = TRUE)

# run dndscv
dndscv.172 <- dndscv(mut.172.dedup, refdb = 'dndscv/refcds/fip_172_mln.rda', sm = '192r_3w', numcode = 11, max_muts_per_gene_per_sample = Inf, max_coding_muts_per_sample = Inf, outmats = TRUE, constrain_wnon_wspl = TRUE)

# add missing line to mle_submodel
dndscv.172$mle_submodel[nrow(dndscv.172$mle_submodel)+1,] = list('wspl', 7.547607e-02, 0.0181692282, 0.31353214)

ci.172 <- geneci(dndscv.172)
ci.172$gene <- factor(ci.172$gene, levels = c('ORF1a', 'ORF1b', 'S', '3a', '3b', '3c', 'E', 'M', 'N', '7a', '7b'))

ggplot(ci.172, aes(gene, mis_mle)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(x = gene, ymin = mis_low, ymax = mis_high), width = 0.2) +
  theme_bw()

# combine global selection scores for plotting
dndscv.all <- list(dndscv.68$globaldnds, dndscv.86$globaldnds, dndscv.120$globaldnds, dndscv.172$globaldnds)
names(dndscv.all) <- c('68', '86', '120', '172')

dndscv.df <- bind_rows(dndscv.all, .id = 'Cat')

# remove NAs, wtru and wall
dndscv.df <- dndscv.df[dndscv.df$name %in% c('wmis', 'wnon'),]

# plot
ggplot(dndscv.df, aes(Cat, mle, fill = name)) +
  geom_bar(stat="identity",position=position_dodge(0.9,preserve='single')) + 
  geom_errorbar(stat='identity',aes(ymin=cilow, ymax=cihigh), position=position_dodge(0.9, preserve = "single"),width=0.2) +
  geom_hline(yintercept = 1, lty = 'dashed', color = 'black') +
  scale_fill_manual(values = c('lightgray', 'darkgrey')) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 1.5), breaks = seq(0,1.5,0.25)) +
  labs(x = 'Cat', y = 'dN/dS', fill = '') +
  theme_bw() +
  theme(panel.grid = element_blank())


# combine data for ORF1a, ORF1b, S and N for each cat, and produce grouped plot
ci.all <- list(ci.68, ci.86, ci.120, ci.172)
names(ci.all) <- c('68', '86', '120', '172')

ci.df <- bind_rows(ci.all, .id = 'Cat')

ci.df.filt <- ci.df[ci.df$gene %in% c('ORF1a', 'ORF1b', 'S', 'N'),]
ci.df.filt$gene <- factor(ci.df.filt$gene, levels = c('ORF1a', 'ORF1b', 'S', 'N'))

ggplot(ci.df.filt, aes(gene, mis_mle, fill = Cat)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(gene, ymin = mis_low, ymax = mis_high), width = 0.2, color = 'black') +
  geom_hline(yintercept = 1, lty = 'dashed', color = 'black') +
  scale_y_continuous(expand = c(0,0), limits = c(0, 2)) +
  scale_fill_manual(values = cat.pal) +
  facet_wrap(vars(Cat), nrow = 1) +
  labs(x = '', y = 'dN/dS (missense variants)') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# plot for alt genes - M and ORF7b
ci.df.filt2 <- ci.df[ci.df$gene %in% c('M', '7b'),]
ci.df.filt2$gene <- factor(ci.df.filt2$gene, levels = c('M', '7b'))

ggplot(ci.df.filt2, aes(gene, mis_mle, fill = Cat)) +
  geom_bar(stat = 'identity') +
  geom_errorbar(aes(gene, ymin = mis_low, ymax = mis_high), width = 0.2, color = 'black') +
  geom_hline(yintercept = 1, lty = 'dashed', color = 'black') +
  scale_y_continuous(expand = c(0,0), limits = c(0, 5)) +
  scale_fill_manual(values = cat.pal) +
  facet_wrap(vars(Cat), nrow = 1) +
  labs(x = '', y = 'dN/dS (missense variants)') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# trinucleotide seq context
library(ggseqlogo)

trinuc.86 <- dndscv.86$annotmuts
trinuc.86$sub <- paste0(trinuc.86$ref, '>', trinuc.86$mut)
# remove indels
trinuc.86 <- trinuc.86[nchar(trinuc.86$sub) == 3,]

ggplot(trinuc.86[trinuc.86$sub %in% c('C>T'),], aes(ref3_cod)) +
  geom_bar(stat = 'count') +
  # facet_wrap(vars(sampleID), scales = 'free_x', nrow = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

trinuc.68 <- dndscv.68$annotmuts
trinuc.68$sub <- paste0(trinuc.68$ref, '>', trinuc.68$mut)
# remove indels
trinuc.68 <- trinuc.68[nchar(trinuc.68$sub) == 3,]

ggplot(trinuc.68[trinuc.68$sub %in% c('C>T'),], aes(ref3_cod)) +
  geom_bar(stat = 'count') +
  # facet_wrap(vars(sampleID), scales = 'free_x', nrow = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

trinuc.172 <- dndscv.172$annotmuts
trinuc.172$sub <- paste0(trinuc.172$ref, '>', trinuc.172$mut)
# remove indels
trinuc.172 <- trinuc.172[nchar(trinuc.172$sub) == 3,]

ggplot(trinuc.172[trinuc.172$sub %in% c('C>T'),], aes(ref3_cod)) +
  geom_bar(stat = 'count') +
  # facet_wrap(vars(sampleID), scales = 'free_x', nrow = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

trinuc.120 <- dndscv.120$annotmuts
trinuc.120$sub <- paste0(trinuc.120$ref, '>', trinuc.120$mut)
# remove indels
trinuc.120 <- trinuc.120[nchar(trinuc.120$sub) == 3,]

ggplot(trinuc.120[trinuc.120$sub %in% c('C>T'),], aes(ref3_cod)) +
  geom_bar(stat = 'count') +
  # facet_wrap(vars(sub), scales = 'free_x', nrow = 1) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# combine data and plot summary
trinuc.all <- list(trinuc.68, trinuc.86, trinuc.120, trinuc.172)
names(trinuc.all) <- c('68', '86', '120', '172')
trinuc.df <- bind_rows(trinuc.all, .id = 'Cat')

# subset C>T mutations
ct.df <- trinuc.df[trinuc.df$sub == 'C>T',]

ct.summ <- ct.df %>%
  group_by(Cat, ref3_cod) %>%
  summarise(count = n()) %>%
  complete(ref3_cod = unique(ct.df$ref3_cod), fill = list(count = 0)) %>%
  mutate(prop = count/sum(count))

ggplot(ct.summ, aes(ref3_cod, prop, fill = Cat)) +
  geom_bar(stat = 'identity', show.legend = F) +
  scale_fill_manual(values = cat.pal) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 0.3), breaks = seq(0,0.3,0.05)) +
  labs(x = 'Trinucleotide sequence context', y = 'Proportion of C>T mutations') +
  facet_wrap(vars(Cat), nrow = 2) +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# seqlogos
ggseqlogo(trinuc.86[trinuc.86$sub %in% c('C>T'),]$ref3_cod, method = 'probability')
ggseqlogo(trinuc.68[trinuc.68$sub %in% c('C>T'),]$ref3_cod, method = 'probability')
ggseqlogo(trinuc.172[trinuc.172$sub %in% c('C>T'),]$ref3_cod, method = 'probability')
ggseqlogo(trinuc.120[trinuc.120$sub %in% c('C>T'),]$ref3_cod, method = 'probability')

# all C>T
ggseqlogo(c(trinuc.86[trinuc.86$sub %in% c('C>T'),]$ref3_cod,
            trinuc.68[trinuc.68$sub %in% c('C>T'),]$ref3_cod,
            trinuc.172[trinuc.172$sub %in% c('C>T'),]$ref3_cod,
            trinuc.120[trinuc.120$sub %in% c('C>T'),]$ref3_cod),
          method = 'probability')
```

## Cross-tissue variants

Look at variants in specific genes (mainly S and M) shared by FCoV isolates from different tissues of the same cat.  

```{r lofreq_genefocus}
# plot shared vars only, for specific genes
# get SNVs detected in > 1 tissue (this also retains multiallelic SNVs)
vars.shared <- vars3 %>%
  filter(`ANN[*].EFFECT` %in% c('missense_variant', 'stop_gained')) %>%
  group_by(cat, POS, ALT) %>%
  filter(n() > 1 | tissue != 'MLN' & AF > 0.5) %>%
  ungroup()

# alternative filter that does not retain variants with AF > 0.5 in a non-MLN tissue (so only shared variants plotted)
vars.shared2 <- vars3 %>%
  filter(`ANN[*].EFFECT` %in% c('missense_variant', 'stop_gained')) %>%
  group_by(cat, POS, ALT) %>%
  filter(n() > 1) %>%
  group_by(POS, ALT, `ANN[*].GENE`) %>%
  filter(n() != 1) %>%
  ungroup()

vars.shared2$tissue <- factor(vars.shared2$tissue, levels = c('MLN', 'Lung', 'Liver'))

# plot changes in AF between tissues
ggplot(vars.shared2, aes(tissue, AF)) +
  geom_boxplot(aes(group = tissue), outlier.shape = NA) +
  geom_point(aes(color = tissue, group = POS)) +
  geom_line(aes(group = POS), color = 'darkgrey', alpha = 0.3) +
  scale_color_manual(values = tissue.pal[c(3,2,1)]) +
  facet_wrap(vars(cat), nrow = 1) +
  labs(x = '', y = 'AAF (cross-tissue variants)') +
  theme_bw() +
  theme(panel.grid = element_blank())

# S/Spike
vars.shared.spike <- vars.shared[vars.shared$`ANN[*].GENE` == 'S',]
vars.shared.spike$`ANN[*].AA_POS` <- as.numeric(vars.shared.spike$`ANN[*].AA_POS`)
vars.shared.spike.ns <- vars.shared.spike[vars.shared.spike$`ANN[*].EFFECT` == 'missense_variant',]
write.table(vars.shared.spike.ns, file = 'vars_shared_spike_ns.txt', sep = '\t')

# correct AA positions based on alignment of the MLN spike seqs
# load in the corrected data and plot
vars.shared.spike.ns.corr <- read.csv('vars_shared_spike_ns_corrected.txt', sep = '\t', header = TRUE)

s.vars.gg <- ggplot(vars.shared.spike.ns.corr, aes(as.numeric(ANN....AA_POS_CORR), AF, colour = tissue)) +
  geom_point() +
  # geom_vline(xintercept = c(275, 540, 695, 800, 942, 1055, 1156), color = 'black', lty = 'dashed') +
  scale_color_manual(values = tissue.pal) +
  facet_wrap(vars(cat), ncol = 1) +
  scale_x_continuous(expand = c(0,0), limits = c(0,1467), breaks = seq(0,1400,200)) +
  labs(x = '', y = 'Alternate allele frequency', color = '') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())

# plot density of variants by codon position
library(GenomicRanges)

# first, get corrected codon positions of all SNVs - export and correct
# write.table(vars3[vars3$`ANN[*].GENE` == 'S',], file = 'all_vars_spike.txt', sep = '\t')

# read in corrected variants
all.vars.spike <- read.csv('all_vars_spike_corrected.txt', sep = '\t', header = TRUE)

# save as GRanges
s.snvs <- GRanges("S", IRanges(all.vars.spike$ANN....AA_POS_CORR, width=1))
s.cvg <- coverage(s.snvs)
s.r <- runsum(s.cvg, 10)

s.r.df <- as(s.r$S, 'data.frame')
s.r.df$pos <- as.numeric(rownames(s.r.df))
colnames(s.r.df) <- c('snvs', 'pos')

s.dens.gg <- ggplot(s.r.df, aes(pos, snvs)) +
  geom_line(color = '#D52D23', linewidth = 0.6) +
  labs(x = '', y = 'No. SNVs') +
  scale_x_continuous(expand = c(0,0), limits = c(0,1467)) +
  theme_bw()
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank())

# plot positions of spike domains
# need to make gff of spike domains based on PDB 6JX7
library(rtracklayer)

spike.gff <- import.gff3('spike_domains_pdb6JX7_forGFF.gff')
spike.domain.names <- spike.gff$ID
spike.df <- as.data.frame(spike.gff)

spike.domains <- ggplot(spike.df) +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 6, colour = 'darkgrey') +
  geom_text(aes(x = (start+end)/2, y = 1, label = ID), colour = 'black', size = 2.5) +
  scale_y_continuous(limits = c(0,3)) +
  scale_x_continuous(expand = c(0,0), breaks = seq(0,1500,200), limits = c(0, 1467)) +
  labs(x = 'Codon position in Spike alignment', y = '') +
  theme_bw() +
  theme(legend.position = 'none',
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        plot.margin = margin(t = 0, r = 5, b = 5, l = 5),
        panel.grid = element_blank())

# combine plots
cowplot::plot_grid(s.dens.gg, s.vars.gg, spike.domains, ncol = 1, axis = 'lrtb', align = 'v', rel_heights = c(0.2,1, 0.2))

# select HR1 mutations for annotation of structural models
vars.hr1 <- vars.shared.spike.ns.corr[vars.shared.spike.ns.corr$ANN....AA_POS_CORR >= 1100 & vars.shared.spike.ns.corr$ANN....AA_POS_CORR < 1120,]

# read in AF matrix and plot hmap for main figure
hr1.mat <- read.csv('hr1mut_af_mat.txt', sep = '\t', row.names = 1)

library(pheatmap)
library(RColorBrewer)
breaks = seq(0,1,0.01)
hmcols = colorRampPalette(rev(c("#bd0026", "#f03b20", "#fd8d3c", "#fecc5c", "#ffffb2")))(length(breaks)-1)

pheatmap(hr1.mat, color = hmcols, breaks = breaks, cluster_rows = FALSE, cluster_cols = FALSE,
         gaps_col = c(1,4,6), show_colnames = FALSE, cellheight = 15, cellwidth = 15)

# M/Membrane
vars.shared.membrane <- vars.shared[vars.shared$`ANN[*].GENE` == 'M',]
vars.shared.membrane$`ANN[*].AA_POS` <- as.numeric(vars.shared.membrane$`ANN[*].AA_POS`)
vars.shared.membrane.ns <- vars.shared.membrane[vars.shared.membrane$`ANN[*].EFFECT` == 'missense_variant',]
# write.table(vars.shared.membrane.ns, file = 'vars_shared_membrane_ns.txt', sep = '\t')

# correct AA positions based on alignment of the MLN membrane seqs
# load in the corrected data and plot
vars.shared.membrane.ns.corr <- read.csv('vars_shared_membrane_ns_corrected.txt', sep = '\t', header = TRUE)

m.vars.gg <- ggplot(vars.shared.membrane.ns.corr, aes(ANN....AA_POS_CORR, AF, colour = tissue)) +
  geom_point() +
  # geom_vline(xintercept = c(275, 540, 695, 800, 942, 1055, 1156), color = 'black', lty = 'dashed') +
  scale_color_manual(values = tissue.pal) +
  facet_wrap(vars(cat), ncol = 1) +
  scale_x_continuous(expand = c(0,0), limits = c(0,274), breaks = seq(0,274,50)) +
  labs(x = '', y = 'Alternate allele frequency', color = '') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank())

# plot density of variants by codon position
# write.table(vars3[vars3$`ANN[*].GENE` == 'M',], file = 'all_vars_membrane.txt', sep = '\t')

# read in corrected variants
all.vars.membrane <- read.csv('all_vars_membrane_corrected.txt', sep = '\t', header = TRUE)

# save as GRanges
m.snvs <- GRanges("M", IRanges(as.numeric(all.vars.membrane$ANN....AA_POS_CORR), width=1))
m.cvg <- coverage(m.snvs)
m.r <- runsum(m.cvg, 10)

m.r.df <- as(m.r$M, 'data.frame')
m.r.df$pos <- as.numeric(rownames(m.r.df))
colnames(m.r.df) <- c('snvs', 'pos')
# append positions with no snvs
m.r.df.add <- data.frame(snvs = c(rep(1, 9), rep(0,20)), pos = seq(246,274,1))
m.r.df <- rbind(m.r.df, m.r.df.add)

m.dens.gg <- ggplot(m.r.df, aes(pos, snvs)) +
  geom_line(color = '#D52D23', linewidth = 0.6) +
  labs(x = '', y = 'No. SNVs') +
  scale_x_continuous(expand = c(0,0), limits = c(0,274), breaks = seq(0,300,50)) +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        panel.grid = element_blank())

# plot positions of membrane domains
# need to make gff of membrane domains based on alignment with SARS-CoV-2 membrane protein
library(rtracklayer)

membrane.gff <- import.gff3('membrane_domains_alphafold_forGFF.gff')
membrane.domain.names <- membrane.gff$ID
membrane.df <- as.data.frame(membrane.gff)

membrane.domains <- ggplot(membrane.df) +
  geom_segment(aes(x = start, xend = end, y = 1, yend = 1), linewidth = 6, colour = 'darkgrey') +
  geom_text(aes(x = (start+end)/2, y = 1, label = ID), colour = 'black', size = 2.5) +
  scale_y_continuous(limits = c(0,3)) +
  scale_x_continuous(expand = c(0,0), breaks = seq(0,300,50), limits = c(0, 274)) +
  labs(x = 'Codon position in Membrane alignment', y = '') +
  theme_bw() +
  theme(legend.position = 'none',
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        plot.margin = margin(t = 0, r = 5, b = 5, l = 5),
        panel.grid = element_blank())

# combine plots
cowplot::plot_grid(m.dens.gg, m.vars.gg, membrane.domains, ncol = 1, axis = 'lrtb', align = 'v', rel_heights = c(0.2,1, 0.2))
```
