---
title: "Fatal consequences of feline coronavirus infection are associated with virus persistence and a distinct adaptive immune repertoire - TCR repertoire analysis pipeline"
author: "Tom Hiron"
date: "2025"
output: 
  html_document:
    theme: "simplex"
    toc: true
    toc_float: true
    highlight: kate
    code_folding: show
    self_contained: true
---
```{r global_options, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      echo = TRUE, 
                      eval = TRUE,
                      fig.path = 'figures/',
                      fig.ext = 'png',
                      cache = FALSE,
                      cache.path = 'cache/')
```

Pipeline to reproduce T Cell Receptor (TCR) repertoire analysis in the manuscript "Fatal consequences of feline coronavirus infection are associated with virus persistence and a distinct adaptive immune repertoire", Hiron et al. 2025, https://www.biorxiv.org/content/10.1101/2025.10.02.679983v1.

## Data availability

Sequencing data supporting the findings of this study have been deposited in the National Center for Biotechnology Information (NCBI) Sequence Read Archive under BioProject ID PRJNA1330231.

## Preparation of feline V(D)J sequence libraries

Sequences for feline V(D)J gene segments available from [IMGT](https://www.imgt.org/download/V-QUEST/IMGT_V-QUEST_reference_directory/Felis_catus/). JSON libraries built using [repseqio](https://github.com/repseqio/repseqio).

```{r repseqio_buildJSONs, engine = 'bash', eval = FALSE}
#!/bin/bash
#SBATCH -J repseqio_buildJSONs.job 
#SBATCH -o repseqio_buildJSONs.out 
#SBATCH -c 2

repseqio fromPaddedFasta -t 9685 -c TRA -j 3 -n 1 -g V -s cat \
    -PFR1Begin=0 -PCDR1Begin=78 -PFR2Begin=114 -PCDR2Begin=165 \
    -PFR3Begin=195 -PCDR3Begin=309 -PVEnd=-1 \
    TRAV.fasta TRAV.json.fasta TRAV.json

repseqio fromPaddedFasta -t 9685 -c TRA -j 3 -n 1 -g J -s cat \
    -PJBegin=0 -PFR4Begin=-31 -LFR4Begin='[WF](G.G)' -PFR4End=-1 \
    TRAJ.fasta TRAJ.json.fasta TRAJ.json

repseqio merge TRAV.json TRAJ.json feline_TRA.json

repseqio fromPaddedFasta -t 9685 -c TRB -j 3 -n 1 -g V -s cat \
    -PFR1Begin=0 -PCDR1Begin=78 -PFR2Begin=114 -PCDR2Begin=165 \
    -PFR3Begin=195 -PCDR3Begin=309 -PVEnd=-1 \
    TRBV.fasta TRBV.json.fasta TRBV.json

repseqio fromPaddedFasta -t 9685 -c TRB -j 3 -n 1 -g D -s cat \
    -PDBegin=0 -PDEnd=-1 \
    TRBD.fasta TRBD.json.fasta TRBD.json

repseqio fromPaddedFasta -t 9685 -c TRB -j 3 -n 1 -g J -s cat \
    -PJBegin=0 -PFR4Begin=-31 -LFR4Begin='[WF](G.G)' -PFR4End=-1 \
    TRBJ.fasta TRBJ.json.fasta TRBJ.json

repseqio merge TRBV.json TRBD.json TRBJ.json feline_TRB.json
```

## TCR repertoire analysis with MiXCR

MiXCR v4.7.0.

```{r mixcr_tcr_secondary, engine = 'bash', eval = FALSE}
#!/bin/bash 
#SBATCH -J mixcr_tcr_alpha_secondary.job 
#SBATCH -o mixcr_tcr_alpha_secondary.%A_%a.out 
#SBATCH -c 4

let i=$SLURM_ARRAY_TASK_ID-1

path=`pwd`
inpath=fastq
traout=TRA
trbout=TRB

cd $inpath

fastqs=()
for fastq in *_1.fastq.gz
do
        fastq=${fastq%_1.fastq.gz}
        fastqs[${#fastqs[@]}]=$fastq
done

cd $path

#put task here
file1=$inpath/${fastqs[$i]}_1.fastq.gz
file2=$inpath/${fastqs[$i]}_2.fastq.gz

mixcr analyze generic-amplicon \
  --species cat \
  --library feline_TRA \
  --rna \
  --rigid-left-alignment-boundary \
  --floating-right-alignment-boundary C \
  --assemble-clonotypes-by CDR3 \
  $file1 $file2 \
  $traout/${fastqs[$i]}
  
mixcr analyze generic-amplicon \
  --species cat \
  --library feline_TRB \
  --rna \
  --rigid-left-alignment-boundary \
  --floating-right-alignment-boundary C \
  --assemble-clonotypes-by CDR3 \
  $file1 $file2 \
  $trbout/${fastqs[$i]}
```

```{r mixcr_tcr_primary, engine = 'bash', eval = FALSE}
#!/bin/bash
path=`pwd`
inpath=fastq

cd $inpath

fastqs=()
for fastq in *_1.fastq.gz
do
        fastq=${fastq%_1.fastq.gz}
        fastqs[${#fastqs[@]}]=$fastq
done
fastqcount=${#fastqs[@]}

cd $path

sbatch --array 1-$fastqcount:1 mixcr_tcr_alpha_secondary.sh
```

## Clonotype analysis

Immunarch (https://immunarch.com/).

Load MiXCR data.

```{r immunarch_load}
library(immunarch)
library(stringr)

tra_mixcr <- repLoad('mixcr/TRA')
trb_mixcr <- repLoad('mixcr/TRB')

# TRA - include metadata file in directory with sample, group and IHC result
names(tra_mixcr$data) <- str_remove(names(tra_mixcr$data), '_tcr.clones_TRA')
tra_mixcr$meta$Sample <- str_remove(tra_mixcr$meta$Sample, '_tcr.clones_TRA')

# TRB - include metadata file in directory with sample, group and IHC result
names(trb_mixcr$data) <- str_remove(names(trb_mixcr$data), '_tcr.clones_TRB')
trb_mixcr$meta$Sample <- str_remove(trb_mixcr$meta$Sample, '_tcr.clones_TRB')
```

Basic analysis - alpha chain.

```{r immunarch_basicTRA}
# TRA
# first, plot the number of clones detected in each sample, faceted by tissue
tra.clns <- data.frame(lapply(tra_mixcr$data, function(x) sum(x$Clones)))
tra.clns <- as.data.frame(t(tra.clns))
tra.clns$sample <- rownames(tra.clns)
tra.clns$group <- tra_mixcr$meta$Group
colnames(tra.clns) <- c('Clones', 'Sample', 'Group')

library(immunarch)
library(ggplot2)
library(ggrepel)
library(cowplot)

tra.clones.gg <- ggplot(tra.clns, aes(Sample, log10(Clones), fill = Group)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  geom_hline(yintercept = c(3, 3.30103, 3.69897), color = 'black', lty = 'dashed') +
  scale_y_continuous(expand = c(0,0), limits = c(0, 5)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# consider a reasonable threshold for downsampling
# plotted above are 1000, 2000 and 5000 clones
# all samples are retained at the 2000 cut off
# 4 samples dropped at 5000 (control_115, control_69, fip_176, fip_83)
# this might be too strict, but ensures sufficient clones for a robust analysis of diversity... 
# set a threshold of 7618, with control_90 at the minimum

# number of clonotypes
tra.exp_vol <- repExplore(tra_mixcr$data, .method = 'volume')
tra.exp_vol$Group <- tra_mixcr$meta$Group
tra.clntypes.gg <- ggplot(tra.exp_vol, aes(Sample, log10(Volume), fill = Group)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 4.5)) +
  labs(x = 'Sample', y = 'Clonotypes') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

plot_grid(tra.clones.gg, tra.clntypes.gg,
          align = 'v', 
          ncol = 1,
          axis = 'lr')

# plot clones vs clonotypes
tra.clns.clntypes <- tra.clns
tra.clns.clntypes$Volume <- tra.exp_vol$Volume
tra.clns.clntypes$IHC <- trb_mixcr$meta$IHC

write.table(tra.clns.clntypes, file = 'repseq_tra_clntypes.txt', sep = '\t')

ggplot(tra.clns.clntypes, aes(Clones, Volume, color = IHC)) +
  geom_point() +
  geom_text_repel(aes(label = Sample)) +
  theme_bw()

# CDR3 lengths
tra.exp_len <- repExplore(tra_mixcr$data, .method = "len", .col = "aa")
vis(tra.exp_len, .by = c('Group'), .meta = tra_mixcr$meta)

# clearly some very long CDR3 sequences, so some filtering required (filter before downsampling to avoid biasing libraries)

# get count of clones from clonotypes with > 1 read for comparison with RNA-seq
tra_mixcr.clnfilt <- repFilter(tra_mixcr, .method = "by.clonotype", .query = list(Clones = morethan(1)))

tra.exp_vol <- repExplore(tra_mixcr.clnfilt$data, .method = 'volume')
tra.exp_vol$Group <- tra_mixcr.clnfilt$meta$Group
ggplot(tra.exp_vol, aes(Sample, log10(Volume), fill = Group)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 4.5)) +
  labs(x = 'Sample', y = 'Clonotypes') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

write.table(tra.exp_vol, 'repseq_tra_mincount1_clntypes.txt', sep = '\t')
```

Basic analysis - beta chain.

```{r immunarch_basicTRB}
# TRB
# first, plot the number of clones detected in each sample, faceted by tissue
trb.clns <- data.frame(lapply(trb_mixcr$data, function(x) sum(x$Clones)))
trb.clns <- as.data.frame(t(trb.clns))
trb.clns$sample <- rownames(trb.clns)
trb.clns$group <- trb_mixcr$meta$Group
colnames(trb.clns) <- c('Clones', 'Sample', 'Group')

library(immunarch)
library(ggplot2)
library(ggrepel)
library(cowplot)

trb.clones.gg <- ggplot(trb.clns, aes(Sample, log10(Clones), fill = Group)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  geom_hline(yintercept = c(3, 3.30103, 3.69897), color = 'black', lty = 'dashed') +
  scale_y_continuous(expand = c(0,0), limits = c(0, 5)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

# consider a reasonable threshold for downsampling, as for TRA
# similar threshold probably suitable, as at 5000 the same 4 samples are dropped
# for TRB, 10539 is suitable, with fip_28 at the minimum

# number of clonotypes
trb.exp_vol <- repExplore(trb_mixcr$data, .method = 'volume')
trb.exp_vol$Group <- trb_mixcr$meta$Group
trb.clntypes.gg <- ggplot(trb.exp_vol, aes(Sample, log10(Volume), fill = Group)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 4.5)) +
  labs(x = 'Sample', y = 'Clonotypes') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

plot_grid(trb.clones.gg, trb.clntypes.gg,
          align = 'v', 
          ncol = 1,
          axis = 'lr')

# plot clones vs clonotypes
trb.clns.clntypes <- trb.clns
trb.clns.clntypes$Volume <- trb.exp_vol$Volume

write.table(trb.clns.clntypes, file = 'repseq_trb_clntypes.txt', sep = '\t')

ggplot(trb.clns.clntypes, aes(Clones, Volume, color = Group)) +
  geom_point() +
  geom_text_repel(aes(label = Sample)) +
  theme_bw()

# CDR3 lengths
trb.exp_len <- repExplore(trb_mixcr$data, .method = "len", .col = "aa")
vis(trb.exp_len, .by = c('Group'), .meta = trb_mixcr$meta)

# again, some very long CDR3 sequences, so some filtering required (filter before downsampling to avoid biasing libraries)

# get count of clones from clonotypes with > 1 read for comparison with RNA-seq
trb_mixcr.clnfilt <- repFilter(trb_mixcr, .method = "by.clonotype", .query = list(Clones = morethan(1)))

trb.exp_vol <- repExplore(trb_mixcr.clnfilt$data, .method = 'volume')
trb.exp_vol$Group <- trb_mixcr.clnfilt$meta$Group
ggplot(trb.exp_vol, aes(Sample, log10(Volume), fill = Group)) +
  geom_bar(stat = 'identity', show.legend = FALSE) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 4.5)) +
  labs(x = 'Sample', y = 'Clonotypes') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())

write.table(trb.exp_vol, 'repseq_trb_mincount1_clntypes.txt', sep = '\t')
```

Filter out CDR3s of aberrant length and samples with low counts (prior to downsampling). 


```{r immunarch_filtering}
library(immunarch)

# filter clonotypes by CDR3 length 
# TRA
for(i in seq_along(tra_mixcr$data)) {
  tra_mixcr$data[[i]]$CDR3.length <- nchar(tra_mixcr$data[[i]]$CDR3.aa)
}

tra_mixcr.f <- repFilter(tra_mixcr, .method = 'by.clonotype', .query = list(CDR3.length = lessthan(30)))

# visualise new CDR3 aa length distribution
tra.filt_len <- repExplore(tra_mixcr.f$data, .method = "len", .col = "aa")
vis(tra.filt_len, .by = 'Group', .meta = tra_mixcr.f$meta)

# TRB
for(i in seq_along(trb_mixcr$data)) {
  trb_mixcr$data[[i]]$CDR3.length <- nchar(trb_mixcr$data[[i]]$CDR3.aa)
}

trb_mixcr.f <- repFilter(trb_mixcr, .method = 'by.clonotype', .query = list(CDR3.length = lessthan(30)))

# visualise new CDR3 aa length distribution
trb.filt_len <- repExplore(trb_mixcr.f$data, .method = "len", .col = "aa")
vis(trb.filt_len, .by = 'Group', .meta = trb_mixcr.f$meta)

# drop samples with low counts (control_115, control_69, fip_176, fip_83)
tra_mixcr.f <- repFilter(tra_mixcr, .method = 'by.meta', .query = list(Sample = exclude('control_115', 'control_69', 'fip_176', 'fip_83')))
trb_mixcr.f <- repFilter(trb_mixcr, .method = 'by.meta', .query = list(Sample = exclude('control_115', 'control_69', 'fip_176', 'fip_83')))

# TRA samples
table(tra_mixcr.f$meta$Group)
tra.f.exp_vol <- repExplore(tra_mixcr.f$data, .method = "volume")
vis(tra.f.exp_vol)
# write.table(tra.f.exp_vol, file = 'tra_clonotypes.txt', sep = '\t')

# TRB samples
table(trb_mixcr.f$meta$Group)
trb.f.exp_vol <- repExplore(trb_mixcr.f$data, .method = "volume")
vis(trb.f.exp_vol)
# write.table(trb.f.exp_vol, file = 'trb_clonotypes.txt', sep = '\t')
```

Downsample repertoires.

```{r immunarch_downsample}
library(immunarch)

# aim is to even out coverage across samples
# need to manually drop the samples to be filtered out
# TRA - set to 7618
tra_mixcr.f.ds <- tra_mixcr.f
tra_mixcr.f.ds$data <- repSample(tra_mixcr.f$data, .method = 'downsample', .n = 7618)

# TRB - set to 10539
trb_mixcr.f.ds <- trb_mixcr.f
trb_mixcr.f.ds$data <- repSample(trb_mixcr.f$data, .method = 'downsample', .n = 10539)
```

Sample-level stats - alpha chain.

```{r immunarch_statsTRA}
library(immunarch)
library(ggplot2)
library(ggpubr)
library(cowplot)

# exploratory analysis
tra.exp_cnt <- repExplore(tra_mixcr.f.ds$data, .method = "count")

vis(tra.exp_cnt, .by = 'IHC', .meta = tra_mixcr.f.ds$meta)

# clonality
tra_pr <- repClonality(tra_mixcr.f.ds$data, .method = 'clonal.prop')
tra_top <- repClonality(tra_mixcr.f.ds$data, .method = 'top', .head = c(10, 100, 1000, 3000, 10000))
tra_rare <- repClonality(tra_mixcr.f.ds$data, .method = 'rare')
tra_hom <- repClonality(tra_mixcr.f.ds$data,
  .method = "homeo",
  .clone.types = c(Small = .0001, Medium = .001, Large = .01, Hyperexpanded = 1)
)

vis(tra_pr, .by = 'IHC', .meta = tra_mixcr.f.ds$meta)

vis(tra_top)
vis(tra_top, .by = 'IHC', .meta = tra_mixcr.f.ds$meta)

vis(tra_rare)
vis(tra_rare, .by = 'IHC', .meta = tra_mixcr.f.ds$meta)

# re-plot with new sample order based on IHC result
ihc.sample_order <- c("control_52", "control_54", "control_56", "control_57", "control_63", "control_78", "control_80", "control_84", "control_90",
                       "control_91", "control_102", "control_111", "control_123", "control_130", "control_135", "control_136", "control_141", "control_144",
                       "control_164", "control_174", "control_182", "control_188", "fip_81", "fip_92", "fip_96", "fip_142", "fip_146", "fip_151", "fip_192",
                       "fip_34", "fip_35", "fip_36", "fip_68", "fip_86", "fip_120", "fip_172", "fip_28", "fip_37", "fip_108", "fip_129", "fip_173")

tra_rare <- tra_rare[ihc.sample_order,]

# borrowed preprocessing code from immunarch github - https://github.com/immunomind/immunarch/blob/master/R/vis.R
  tmp <- tra_rare
  if (is.null(dim(tmp))) {
    tmp <- t(as.matrix(tmp))
    tra_rare <- t(as.matrix(tra_rare))
  }

  colnames_new <- c()
  prev_value <- "1"
  start_index <- 1
  if (colnames(tra_rare)[1] == "1") {
    colnames_new <- "1"
    prev_value <- "2"
    start_index <- 2
  }
  for (i in start_index:ncol(tra_rare)) {
    colnames_new <- c(colnames_new, paste0(prev_value, " - ", colnames(tra_rare)[i]))
    if (colnames(tra_rare)[i] != "MAX") {
      prev_value <- as.character(as.integer(colnames(tra_rare)[i]) + 1)
    }
  }
  colnames(tmp) <- colnames_new

  for (i in 2:ncol(tra_rare)) {
    tmp[, i] <- tra_rare[, i] - tra_rare[, i - 1]
  }
  tra.res <- tmp

  tra.res <- as.data.frame(tra.res)
  tra.res$Sample <- row.names(tra.res)
  tra.res <- reshape2::melt(tra.res)
  colnames(tra.res) <- c("Sample", "Counts", "Value")
  
  tra.res$Sample <- factor(tra.res$Sample, levels = ihc.sample_order)

ggplot(tra.res, aes(Sample, Value, group = Counts, fill = Counts)) +
  geom_bar(stat = 'identity', position = 'stack', color = 'black', width = 1) +
  scale_y_continuous(expand = c(0,0), labels = seq(0,100,25)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c('#FF4B20','#FF9F30','#DCDFA2','#A8E6F3','#63ABEC','#0348A6')) +
  labs(x = '', y = 'Occupied repertoire space (%)', fill = 'Clonotype counts') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

vis(tra_hom)
vis(tra_hom, .by = 'IHC', .meta = tra_mixcr.f.ds$meta)

# diversity estimates
# chao
alpha.chao <- repDiversity(tra_mixcr.f.ds$data, 'chao1')

vis(alpha.chao, .by = 'IHC', .meta = tra_mixcr.f.ds$meta)

# nice boxplots
alpha.chao.df <- as.data.frame(alpha.chao)
alpha.chao.df$IHC <- tra_mixcr.f.ds$meta$IHC

compare_means(Estimator ~ IHC, data = alpha.chao.df)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

alpha.chao.gg <- ggplot(alpha.chao.df, aes(IHC, Estimator, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 40000)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 37000) +
  labs(x = '', y = 'Chao1 index') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# gini
alpha.div_gini <- repDiversity(tra_mixcr.f.ds$data, 'gini')
alpha.div_gini <- as.data.frame(alpha.div_gini)
alpha.div_gini$group <- tra_mixcr.f.ds$meta$Group
alpha.div_gini$IHC <- tra_mixcr.f.ds$meta$IHC

# get p values
compare_means(V1 ~ IHC, data = alpha.div_gini)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

alpha.gini.gg <- ggplot(alpha.div_gini, aes(IHC, V1, group = IHC, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, show.legend = TRUE, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 1)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 1) +
  labs(x = '', y = 'Gini coefficient') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# align chao and gini plots to get equal sized plots
plot_grid(alpha.chao.gg, alpha.gini.gg,
          align = 'v', 
          ncol = 1,
          axis = 'lr')
```

Sample-level stats - beta chain.

```{r immunarch_statsTRB}
library(immunarch)
library(ggplot2)
library(ggpubr)
library(cowplot)

# exploratory analysis
trb.exp_cnt <- repExplore(trb_mixcr.f.ds$data, .method = "count")

vis(trb.exp_cnt, .by = 'IHC', .meta = trb_mixcr.f.ds$meta)

# clonality
trb_pr <- repClonality(trb_mixcr.f.ds$data, .method = 'clonal.prop')
trb_top <- repClonality(trb_mixcr.f.ds$data, .method = 'top', .head = c(10, 100, 1000, 3000, 10000))
trb_rare <- repClonality(trb_mixcr.f.ds$data, .method = 'rare')
trb_hom <- repClonality(trb_mixcr.f.ds$data,
  .method = "homeo",
  .clone.types = c(Small = .0001, Medium = .001, Large = .01, Hyperexpanded = 1)
)

vis(trb_pr, .by = 'IHC', .meta = trb_mixcr.f.ds$meta)

vis(trb_top)
vis(trb_top, .by = 'IHC', .meta = trb_mixcr.f.ds$meta)

vis(trb_rare)
vis(trb_rare, .by = 'IHC', .meta = trb_mixcr.f.ds$meta)

# re-plot with new sample order based on IHC result
ihc.sample_order <- c("control_52", "control_54", "control_56", "control_57", "control_63", "control_78", "control_80", "control_84", "control_90",
                       "control_91", "control_102", "control_111", "control_123", "control_130", "control_135", "control_136", "control_141", "control_144",
                       "control_164", "control_174", "control_182", "control_188", "fip_81", "fip_92", "fip_96", "fip_142", "fip_146", "fip_151", "fip_192",
                       "fip_34", "fip_35", "fip_36", "fip_68", "fip_86", "fip_120", "fip_172", "fip_28", "fip_37", "fip_108", "fip_129", "fip_173")

trb_rare <- trb_rare[ihc.sample_order,]

# borrowed preprocessing code from immunarch github - https://github.com/immunomind/immunarch/blob/master/R/vis.R
  tmp <- trb_rare
  if (is.null(dim(tmp))) {
    tmp <- t(as.matrix(tmp))
    trb_rare <- t(as.matrix(trb_rare))
  }

  colnames_new <- c()
  prev_value <- "1"
  start_index <- 1
  if (colnames(trb_rare)[1] == "1") {
    colnames_new <- "1"
    prev_value <- "2"
    start_index <- 2
  }
  for (i in start_index:ncol(trb_rare)) {
    colnames_new <- c(colnames_new, paste0(prev_value, " - ", colnames(trb_rare)[i]))
    if (colnames(trb_rare)[i] != "MAX") {
      prev_value <- as.character(as.integer(colnames(trb_rare)[i]) + 1)
    }
  }
  colnames(tmp) <- colnames_new

  for (i in 2:ncol(trb_rare)) {
    tmp[, i] <- trb_rare[, i] - trb_rare[, i - 1]
  }
  trb.res <- tmp

  trb.res <- as.data.frame(trb.res)
  trb.res$Sample <- row.names(trb.res)
  trb.res <- reshape2::melt(trb.res)
  colnames(trb.res) <- c("Sample", "Counts", "Value")
  
  trb.res$Sample <- factor(trb.res$Sample, levels = ihc.sample_order)

ggplot(trb.res, aes(Sample, Value, group = Counts, fill = Counts)) +
  geom_bar(stat = 'identity', position = 'stack', color = 'black', width = 1) +
  scale_y_continuous(expand = c(0,0), labels = seq(0,100,25)) +
  scale_x_discrete(expand = c(0,0)) +
  scale_fill_manual(values = c('#FF4B20','#FF9F30','#DCDFA2','#A8E6F3','#63ABEC','#0348A6')) +
  labs(x = '', y = 'Occupied repertoire space (%)', fill = 'Clonotype counts') +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

vis(trb_hom)
vis(trb_hom, .by = 'IHC', .meta = trb_mixcr.f.ds$meta)

# diversity estimates
# chao
beta.chao <- repDiversity(trb_mixcr.f.ds$data, 'chao1')

vis(beta.chao, .by = 'IHC', .meta = trb_mixcr.f.ds$meta)

# nice boxplots
beta.chao.df <- as.data.frame(beta.chao)
beta.chao.df$IHC <- trb_mixcr.f.ds$meta$IHC

compare_means(Estimator ~ IHC, data = beta.chao.df)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

beta.chao.gg <- ggplot(beta.chao.df, aes(IHC, Estimator, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 60000)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 57000) +
  labs(x = '', y = 'Chao1 index') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# gini
beta.div_gini <- repDiversity(trb_mixcr.f.ds$data, 'gini')
beta.div_gini <- as.data.frame(beta.div_gini)
beta.div_gini$group <- trb_mixcr.f.ds$meta$Group
beta.div_gini$IHC <- trb_mixcr.f.ds$meta$IHC

# get p values
compare_means(V1 ~ IHC, data = beta.div_gini)
my.comparisons <- list(c('Control', 'FCoV -'), c('FCoV -', 'FCoV +'), c('Control', 'FCoV +'))

beta.gini.gg <- ggplot(beta.div_gini, aes(IHC, V1, group = IHC, fill = IHC)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE, alpha = 0.5) +
  geom_jitter(aes(color = IHC), width = 0.1, show.legend = FALSE) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_color_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  scale_y_continuous(limits = c(0, 1)) +
  stat_compare_means(comparisons = my.comparisons) +
  stat_compare_means(label.y = 1) +
  labs(x = '', y = 'Gini coefficient') +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

# align chao and gini plots to get equal sized plots
plot_grid(beta.chao.gg, beta.gini.gg,
          align = 'v', 
          ncol = 1,
          axis = 'lr')
```

Repertoire overlap and public clonotypes.

```{r immunarch_overlap}
library(immunarch)

# TRA
# public - simple count of public clonotypes
tra_ov1 <- repOverlap(tra_mixcr.f.ds$data[ihc.sample_order], .method = 'public', .verbose = F)
vis(tra_ov1, 'heatmap2', cluster_rows = FALSE, cluster_cols = FALSE)
vis(tra_ov1, .text.size = 0) +
  labs(x = '', y = '', title = 'Public clonotypes')

# jaccard - jaccard index, size of intersection/size of union
tra_ov2 <- repOverlap(tra_mixcr.f.ds$data[ihc.sample_order], .method = 'jaccard', .verbose = F)
vis(tra_ov2, 'heatmap2', cluster_rows = FALSE, cluster_cols = FALSE)
vis(tra_ov2, .text.size = 0) +
  labs(x = '', y = '', title = 'Jaccard index')

# public repertoire
tra.pr.aa <- pubRep(tra_mixcr.f.ds$data, "aa", .verbose = F)

# TRB
# public - simple count of public clonotypes
trb_ov1 <- repOverlap(trb_mixcr.f.ds$data[ihc.sample_order], .method = 'public', .verbose = F)
vis(trb_ov1, 'heatmap2', cluster_rows = FALSE, cluster_cols = FALSE)
vis(trb_ov1, .text.size = 0) +
  labs(x = '', y = '', title = 'Public clonotypes')

# jaccard - jaccard index, size of intersection/size of union
trb_ov2 <- repOverlap(trb_mixcr.f.ds$data[ihc.sample_order], .method = 'jaccard', .verbose = F)
vis(trb_ov2, 'heatmap2', cluster_rows = FALSE, cluster_cols = FALSE)
vis(trb_ov2, .text.size = 0) +
  labs(x = '', y = '', title = 'Jaccard index')

# public repertoire
trb.pr.aa <- pubRep(trb_mixcr.f.ds$data, "aa", .verbose = F)
```

Gene usage.

First fix the functions distributed with immunarch.

```{r immunarch_gene_usage_altfunctions}
geneUsage2 <- function(.data, # df, list, MonetDB
                      .gene = c("hs.trbv", "HomoSapiens.TRBJ", "macmul.IGHV"),
                      .quant = c(NA, "count"),
                      .ambig = c("inc", "exc", "maj"),
                      .type = c("segment", "allele", "family"),
                      .norm = FALSE) {
  .type <- .type[1]
  .ambig <- .ambig[1]
  .quant <- .quant[1]
  .gene <- .gene[1]
  
  if (immunarch:::has_class(.data, "list")) {
    if (length(.data) == 1) {
      return(geneUsage2(.data[[1]], .gene = .gene, .quant = .quant, .ambig = .ambig, .norm = .norm))
    } else {
      genedata <- lapply(.data, geneUsage2,
                         .gene = .gene, .quant = .quant,
                         .ambig = .ambig, .type = .type, .norm = .norm
      )
      
      # Add check if list has 1 element
      for (i in 2:(length(genedata))) {
        if (i > 2) {
          allres <- full_join(allres, genedata[[i]], by = "Names")
          names(allres)[i + 1] <- names(genedata)[i]
        } else {
          allres <- full_join(genedata[[1]], genedata[[2]], by = "Names")
          names(allres)[2] <- names(genedata)[1]
          names(allres)[3] <- names(genedata)[2]
        }
      }
      allres <- allres[order(allres$Names), ]
      
      return(immunarch:::add_class(allres, "immunr_gene_usage"))
    }
  }
  
  "%!in%" <- function(x, y) !("%in%"(x, y))
  
  which_gene <- strsplit(.gene, ".", TRUE)[[1]][2]
  gene_col <- tolower(substr(which_gene, 4, 4))
  if (gene_col == "v") {
    gene_col <- immunarch:::IMMCOL$v
  } else if (gene_col == "d") {
    gene_col <- immunarch:::IMMCOL$d
  } else if (gene_col == "j") {
    gene_col <- immunarch:::IMMCOL$j
  } else {
    stop("The entered gene_col name is invalid")
  }
  
  .gene <- get_genes2(.gene, .type)
  
  # Bad Cases:
  # X.name is segments, .type == "alleles"
  # X.name is families, .type == "segments"
  # X.name is families, .type == "alleles"
  
  if (.ambig == "inc") {
    if (.type == "segment") {
      .data[[gene_col]] <- immunarch:::return_segments(.data[[gene_col]])
    } else if (.type == "family") {
      .data[[gene_col]] <- immunarch:::return_families(.data[[gene_col]])
    }
  }
  
  if (immunarch:::has_class(.data, "data.table")) {
    .data <- .data %>% lazy_dt()
  }
  dataset <- .data %>%
    select(Gene = gene_col, immunarch:::IMMCOL$count) %>%
    group_by(Gene) %>%
    rename(Quant = immunarch:::IMMCOL$count)
  
  if (is.na(.quant)) {
    dataset <- dataset %>%
      summarise(count = n())
  } else {
    dataset <- dataset %>%
      summarise(count = sum(Quant))
  }
  
  dataset <- dataset %>%
    collect(n = Inf)
  
  if (.ambig == "inc") {
    res <- dataset
    names(res) <- c("Names", immunarch:::IMMCOL$count)
  } else {
    gene_col_split <- strsplit(dataset[["Gene"]], ", ", fixed = TRUE, useBytes = TRUE)
    counts <- dataset$count
    names(counts) <- dataset[["Gene"]]
    
    if (.ambig == "exc") {
      included_names <- names(counts) %in% gene_col_split[sapply(gene_col_split, length) == 1]
      chosen <- counts[included_names]
      chosen <- chosen[names(chosen) %in% .gene]
      other <- counts[names(chosen) %!in% .gene]
      res <- data.frame(
        Names = names(chosen),
        Count = unlist(chosen), stringsAsFactors = FALSE
      )
    } else if (.ambig == "maj") {
      maj_counts <- data.frame(Names = sapply(gene_col_split, "[[", 1), Counts = counts, stringsAsFactors = FALSE)
      row.names(maj_counts) <- NULL
      res <- maj_counts %>%
        group_by(Names) %>%
        summarise(Counts = sum(Counts))
    }
  }
  if (length(.gene %in% res[[1]]) < length(.gene)) {
    missing <- data.frame(.gene[!.gene %in% res[[1]]],
                          0,
                          stringsAsFactors = FALSE
    )
    names(missing) <- names(res)
    res <- rbind(res, missing)
  }
  row.names(res) <- NULL
  res <- res[order(res[[1]]), ]
  if (.norm) {
    res[, 2] <- res[, 2] / sum(res[, 2])
  }
  res <- as_tibble(as.data.frame(res))
  immunarch:::add_class(res, "immunr_gene_usage")
}



get_genes2 <- function(.gene = c("hs.trbv", "HomoSapiens.trbj", "macmul.ighv"), .type = c("segment", "allele", "family")) {
  .gene <- .gene[1]
  .type <- .type[1]
  
  # Species
  .gene <- tolower(.gene)
  which_species <- strsplit(.gene, ".", TRUE)[[1]][1]
  if (which_species %in% GENE_SEGMENTS2$alias) {
    which_species_col <- "alias"
  } else {
    if (which_species %in% tolower(GENE_SEGMENTS2$species)) {
      which_species_col <- "species"
    } else {
      stop("Unknown species name / alias")
    }
  }
  
  # Genes
  which_gene <- strsplit(.gene, ".", TRUE)[[1]][2]
  if (!(which_gene %in% GENE_SEGMENTS2[["gene"]][tolower(GENE_SEGMENTS2[[which_species_col]]) == which_species])) {
    stop("Unknown gene name")
  }
  
  # Id's type
  which_type <- paste0(.type, "_id")
  
  sort(unique(GENE_SEGMENTS2[[which_type]][tolower(GENE_SEGMENTS2[[which_species_col]]) == which_species & GENE_SEGMENTS2[["gene"]] == which_gene]))
}
```

```{r immunarch_gene_usage}
tra_mixcr.f.gu <- tra_mixcr.f.ds

library(immunarch)
library(tidyverse)
library(pheatmap)
library(RColorBrewer)

tra_mixcr.f.gu$data <- tra_mixcr.f$data %>% 
  map(~ .x %>% mutate(V.name = gsub('\\*01\\(.{2,7}\\)', '', V.name), 
                      J.name = gsub('\\*01\\(.{2,7}\\)', '', J.name)))

# constructed GENE_SEGMENTS entries for feline VDJ genes - formatted as follows
# alias	species	gene	family_id	segment_id	allele_id	reference_point	sequence
# fc	FelisCatus	traj	TRAJ1	TRAJ1	TRAJ1*01		
# fc	FelisCatus	traj	TRAJ10	TRAJ10	TRAJ10*01		
# fc	FelisCatus	traj	TRAJ11	TRAJ11	TRAJ11*01		
# fc	FelisCatus	traj	TRAJ12	TRAJ12	TRAJ12*01		

# read in and rbind
feline.gene_segments <- read.csv(file = 'gene_segments_cat.txt', sep = '\t', header = T)
GENE_SEGMENTS2 <- rbind(GENE_SEGMENTS, feline.gene_segments)

# calculate and plot gene usage
# TRAV
trav_gu <- geneUsage2(tra_mixcr.f.gu$data, "fc.trav", .quant = 'count', .ambig = 'exc', .norm = T)
vis(trav_gu, .by = "Group", .plot = 'box', .meta = tra_mixcr.f.gu$meta)
vis(trav_gu, .by = "IHC", .plot = 'box', .meta = tra_mixcr.f.gu$meta)

trav_gu2 <- as.data.frame(trav_gu)
rownames(trav_gu2) <- trav_gu2$Names
trav_gu2$Names <- NULL
trav_gu2 <- trav_gu2[,ihc.sample_order]
trav.breaks = seq(0,0.2,0.005)
trav.cols = colorRampPalette(brewer.pal(9, 'YlGnBu'))(length(trav.breaks))
pheatmap(trav_gu2, cluster_cols = FALSE, breaks = trav.breaks, color = trav.cols)

trav_gu_cor <- geneUsageAnalysis(trav_gu, .method = "cor", .verbose = F)
vis(trav_gu_cor, .title = "Gene usage correlation", .leg.title = "Cor", .text.size = 0)

# TRAJ
traj_gu <- geneUsage2(tra_mixcr.f.gu$data, "fc.traj", .quant = 'count', .ambig = 'exc', .norm = T)
vis(traj_gu, .by = "Group", .plot = 'box', .meta = tra_mixcr.f.gu$meta)
vis(traj_gu, .by = "IHC", .plot = 'box', .meta = tra_mixcr.f.gu$meta)

traj_gu2 <- as.data.frame(traj_gu)
rownames(traj_gu2) <- traj_gu2$Names
traj_gu2$Names <- NULL
traj_gu2 <- traj_gu2[,ihc.sample_order]
traj.breaks = seq(0,0.2,0.005)
traj.cols = colorRampPalette(brewer.pal(9, 'YlGnBu'))(length(traj.breaks))
pheatmap(traj_gu2, cluster_rows = FALSE, cluster_cols = FALSE, breaks = traj.breaks, color = traj.cols)

traj_gu_cor <- geneUsageAnalysis(traj_gu, .method = "cor", .verbose = F)
vis(traj_gu_cor, .title = "Gene usage correlation", .leg.title = "Cor", .text.size = 0)

# prepare trb
trb_mixcr.f.gu <- trb_mixcr.f.ds

trb_mixcr.f.gu$data <- trb_mixcr.f$data %>% 
  map(~ .x %>% mutate(V.name = gsub('\\*01\\(.{2,7}\\)', '', V.name), 
                      J.name = gsub('\\*01\\(.{2,7}\\)', '', J.name)))

# TRBV
trbv_gu <- geneUsage2(trb_mixcr.f.gu$data, "fc.trbv", .quant = 'count', .ambig = 'exc', .norm = T)
vis(trbv_gu, .by = "Group", .plot = 'box', .meta = trb_mixcr.f.gu$meta, .test = T)
vis(trbv_gu, .by = "IHC", .plot = 'box', .meta = trb_mixcr.f.gu$meta, .test = T)

trbv_gu2 <- as.data.frame(trbv_gu)
rownames(trbv_gu2) <- trbv_gu2$Names
trbv_gu2$Names <- NULL
trbv_gu2 <- trbv_gu2[,ihc.sample_order]
trbv.breaks = seq(0,0.3,0.005)
trbv.cols = colorRampPalette(brewer.pal(9, 'YlGnBu'))(length(trbv.breaks))
pheatmap(trbv_gu2, cluster_cols = FALSE, breaks = trbv.breaks, color = trbv.cols)

trbv_gu_cor <- geneUsageAnalysis(trbv_gu, .method = "cor", .verbose = F)
vis(trbv_gu_cor, .title = "Gene usage correlation", .leg.title = "Cor", .text.size = 0)

# TRBJ
trbj_gu <- geneUsage2(trb_mixcr.f.gu$data, "fc.trbj", .quant = 'count', .ambig = 'exc', .norm = T)
vis(trbj_gu, .by = "Group", .meta = trb_mixcr.f.gu$meta, .test = T)
vis(trbj_gu, .by = "IHC", .meta = trb_mixcr.f.gu$meta, .test = T)

trbj_gu2 <- as.data.frame(trbj_gu)
rownames(trbj_gu2) <- trbj_gu2$Names
trbj_gu2$Names <- NULL
trbj_gu2 <- trbj_gu2[,ihc.sample_order]
trbj.breaks = seq(0,0.4,0.005)
trbj.cols = colorRampPalette(brewer.pal(9, 'YlGnBu'))(length(trbj.breaks))
pheatmap(trbj_gu2, cluster_cols = FALSE, breaks = trbj.breaks, color = trbj.cols)

trbj_gu_cor <- geneUsageAnalysis(trbj_gu, .method = "cor", .verbose = F)
vis(trbj_gu_cor, .title = "Gene usage correlation", .leg.title = "Cor", .text.size = 0)
```

## Network analysis (NAIR)

https://mlizhangx.github.io/Network-Analysis-for-Repertoire-Sequencing-/

```{r nair_fipclusters_alpha}
library(tidyverse)
library(NAIR)
library(igraph)
library(ggraph)

# stash data
alpha.dat <- tra_mixcr.f.ds$data
control.dat <- tra_mixcr.f.ds$data[1:22]
fcovneg.samples <- c("fip_81", "fip_92", "fip_96", "fip_142", "fip_146", "fip_151", "fip_192")
fcovplus.samples <- c("fip_34", "fip_35", "fip_36", "fip_68", "fip_86", "fip_120", "fip_172")

# save each sample as rds
lapply(names(alpha.dat), function(x) saveRDS(alpha.dat[[x]], file = paste0('nair/tcr_manuscript/alpha/', paste0(x, '.rds'))))

# define input files
alpha.input.files <- list.files(path = 'nair/tcr_manuscript/alpha', pattern = '.rds', full.names = TRUE)
group_labels <- c(rep('Control', 22), rep('FIP', 19))

# find associated seqs - takes a while to run so results saved and loaded from a txt file
# alpha.associated_seqs <- findAssociatedSeqs(file_list = alpha.input.files,
#                                       input_type = 'rds',
#                                       group_ids = group_labels,
#                                       seq_col = 'CDR3.aa',
#                                       freq_col = 'Proportion',
#                                       min_seq_length = NULL,
#                                       drop_matches = NULL,
#                                       min_sample_membership = NULL,
#                                       pval_cutoff = 0.05)

# write.table(alpha.associated_seqs, file = 'alpha_associated_seqs_tcr_manuscript.txt', sep = '\t')
alpha.associated_seqs <- read.csv('alpha_associated_seqs_tcr_manuscript.txt', sep = '\t')

# find associated clones (top 20 seqs)
findAssociatedClones(file_list = alpha.input.files,
                     input_type = 'rds',
                     group_ids = group_labels,
                     seq_col = 'CDR3.aa',
                     assoc_seqs = alpha.associated_seqs$ReceptorSeq,
                     min_seq_length = NULL,
                     drop_matches = NULL,
                     output_dir = 'nair/tcr_manuscript/alpha/assoc_seq_nbds',
                     sample_ids = names(alpha.dat))

# combine neighbourhoods and perform network analysis
alpha.nbd.input <- list.files('nair/tcr_manuscript/alpha/assoc_seq_nbds', full.names = TRUE)

alpha.all_clusters <- buildAssociatedClusterNetwork(file_list = alpha.nbd.input,
                                                    seq_col = 'CDR3.aa',
                                                    size_nodes_by = 'Proportion',
                                                    print_plots = TRUE,
                                                    drop_isolated_nodes = FALSE)

# export CDR3s to get pGens with OLGA
alpha.all_clusters.cdr3 <- data.frame(unique(alpha.all_clusters$node_data$CDR3.aa))
# write.table(alpha.all_clusters.cdr3, file = 'alpha_all_clusters_unique_cdr3s_tcr_manuscript.txt', sep = '\t')

# add IHC group to node data
control.samples <- names(control.dat)
alpha.all_clusters$node_data$IHC <- ifelse(alpha.all_clusters$node_data$SampleID %in% control.samples, 'Control', 
                                           ifelse(alpha.all_clusters$node_data$SampleID %in% fcovneg.samples, 'FCoV -',
                                                  ifelse(alpha.all_clusters$node_data$SampleID %in% fcovplus.samples, 'FCoV +', 'FCoV ND')))
# export network for visualisation in gephi
alpha.all_clusters.elist <- igraph::as_edgelist(alpha.all_clusters$igraph, names = T)
write.table(alpha.all_clusters.elist, file = 'alpha_all_clusters_edgelist_tcr_manuscript.csv', sep = ',')
write.table(alpha.all_clusters$node_data, file = 'alpha_all_clusters_nodedata_tcr_manuscript.csv', sep = ',')
```

Prepare network visualisation in Gephi (https://gephi.org/).

```{r nair_fipclusters_hmap_alpha}
# for all clusters
alpha.all <- alpha.all_clusters$node_data
# count of clonotypes
clntyp.count.matrix <- unclass(table(alpha.all$AssocSeq, alpha.all$SampleID))
# mean frequency
clntyp.freq.matrix <- reshape2::acast(alpha.all, cluster_id ~ SampleID, value.var = 'Proportion', fun.aggregate = mean, fill = 0)
# change column order to group by IHC
clntyp.freq.ihc <- clntyp.freq.matrix[,ihc.sample_order]

# set row order by delta mean freq
clntyp.freq.ihc.df <- as.data.frame(clntyp.freq.ihc)
clntyp.freq.ihc.df$control_mean <- rowMeans(subset(clntyp.freq.ihc.df, select = c(1:22)))
clntyp.freq.ihc.df$fip_mean <- rowMeans(subset(clntyp.freq.ihc.df, select = c(23:41)))
clntyp.freq.ihc.df$delta <- clntyp.freq.ihc.df$fip_mean - clntyp.freq.ihc.df$control_mean
clntyp.freq.ihc.df <- clntyp.freq.ihc.df[order(clntyp.freq.ihc.df$delta, decreasing = FALSE),]
# add sample count for alt plot
clntyp.freq.ihc.df$nsamples <- rowSums(clntyp.freq.ihc.df[,1:41] > 0)
# and control/fip-exclusive clusters
clntyp.freq.ihc.df$control_excl <- clntyp.freq.ihc.df$fip_mean == 0
clntyp.freq.ihc.df$fip_excl <- clntyp.freq.ihc.df$control_mean == 0
# convert to matrix for hmap
clntyp.freq.ihc.mat <- as.matrix(clntyp.freq.ihc.df[,1:41])

library(pheatmap)
library(RColorBrewer)
breaks = seq(0, 0.001, 0.00001)
hmcols = colorRampPalette(brewer.pal(9, 'OrRd'))(length(breaks))
pheatmap(clntyp.freq.ihc.mat,
         breaks = breaks,
         color = hmcols,
         border_color = NA,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = FALSE,
         fontsize_row = 5,
         angle_col = 45,
         gaps_col = c(22,29,36))

# plot delta mean freq to add to main hmap in illustrator
paletteLength <- 1000
delta.cols <- colorRampPalette(c('#32A7CD', 'white', '#8F3092'))(paletteLength)
# length(breaks) == length(paletteLength) + 1
# use floor and ceiling to deal with even/odd length pallettelengths
myBreaks <- c(seq(min(clntyp.freq.ihc.df$delta), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(clntyp.freq.ihc.df$delta)/paletteLength, max(clntyp.freq.ihc.df$delta), length.out=floor(paletteLength/2)))

pheatmap(as.matrix(clntyp.freq.ihc.df$delta), 
         breaks = myBreaks, 
         color = delta.cols, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE, 
         border_color = NA)
```

Calculate TCR generation probabilities (pGen) using OLGA (https://github.com/statbiophys/OLGA).

First, generate recombination models using Pygor (https://github.com/statbiophys/pygor3) with non-functional CDR3 sequences.

```{r pgen_models, engine = 'bash', eval = FALSE}
pygor model-create -M models/Felis+catus/TRA/ -t VJ
pygor igor-infer -M models/Felis+catus/TRA/ -i nonfunc/all_samples_nonfunc_cdr3_seqsonly_TRA.txt -o new_TRA

pygor model-create -M models/Felis+catus/TRB/ -t VDJ
pygor igor-infer -M models/Felis+catus/TRB/ -i nonfunc/all_samples_nonfunc_cdr3_seqsonly_TRB.txt -o new_TRB
```

Calculate pGens for FIP-associated clonotypes.

```{r pgen_calc, engine = 'bash', eval = FALSE}
olga-compute_pgen --set_custom_model_VJ models/TRA/ -i unique_cdr3s/alpha_all_clusters_unique_cdr3s_tcr_manuscript.txt -o output/alpha_all_clusters_unique_cdr3s_tcr_manuscript_pgens.tsv

olga-compute_pgen --set_custom_model_VDJ models/TRB/ -i unique_cdr3s/beta_all_clusters_unique_cdr3s_tcr_manuscript.txt -o output/beta_all_clusters_unique_cdr3s_tcr_manuscript_pgens.tsv
```

```{r nair_fipclusters_pgenplots_alpha}
library(ggrepel)

# read in pGens from OLGA and merge with clone node data
alpha.all_clusters.pgens <- read.csv('alpha_all_clusters_unique_cdr3s_tcr_manuscript_pgens.tsv', sep = '\t', header = TRUE)
alpha_pgens <- merge(alpha.all_clusters$node_data, alpha.all_clusters.pgens, by.x = 'CDR3.aa', by.y = 'Seq', all.x = TRUE)

# calculate median cluster pGen
alpha.assoccluster.pgen <- alpha_pgens %>%
  group_by(cluster_id) %>%
  summarise(medianpgen = median(Score, na.rm = TRUE))

# get cluster data for plotting
clntyp.freq.ihc.df$cluster_id <- rownames(clntyp.freq.ihc.df)
freq.pgen.plot <- subset(clntyp.freq.ihc.df, select = c(cluster_id, delta, nsamples, control_excl, fip_excl))
freq.pgen.plot <- merge(freq.pgen.plot, alpha.assoccluster.pgen, by = 'cluster_id', all.x = TRUE)

# add seq_w_max_degree for labelling
alpha.clusterstat <- subset(alpha.all_clusters$cluster_data, select = c(cluster_id, seq_w_max_degree))
freq.pgen.plot <- merge(freq.pgen.plot, alpha.clusterstat, by = 'cluster_id', all.x = TRUE)

# plot
ggplot(freq.pgen.plot, aes(x = delta, y = medianpgen, size = nsamples)) +
  geom_vline(xintercept = 0, color = 'darkgrey', lty = 'dashed') +
  geom_point(shape = 21) +
  geom_point(data = freq.pgen.plot[freq.pgen.plot$fip_excl == TRUE,], aes(x = delta, y = medianpgen, size = nsamples), shape = 21, fill = '#8F3092') +
  geom_point(data = freq.pgen.plot[freq.pgen.plot$control_excl == TRUE,], aes(x = delta, y = medianpgen, size = nsamples), shape = 21, fill = '#32A7CD') +
  # geom_text_repel(aes(label = seq_w_max_degree, size = 5), show.legend = FALSE) +
  scale_x_continuous(limits = c(-0.0005,0.001), breaks = seq(-5e-04, 1e-03, 2.5e-04)) +
  scale_y_continuous(limits = c(50, 140), breaks = seq(50, 140, 10)) +
  scale_size_continuous(breaks = seq(5,25,10)) +
  labs(x = 'Δ cluster mean frequency (FIP - Control)', y = expression(-log[2]*' cluster median pGen')) +
  theme_bw() +
  theme(panel.grid = element_blank())

# plot zoomed view of selected clusters
# top 20?
clusters.plot <- alpha.associated_seqs[order(alpha.associated_seqs$fisher_pvalue, decreasing = FALSE),]$ReceptorSeq[1:20]

alpha.pgens.plot <- alpha_pgens[alpha_pgens$AssocSeq %in% clusters.plot,]

ggplot(alpha.pgens.plot, aes(Proportion, Score, fill = IHC)) +
  geom_point(shape = 21) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  facet_wrap(vars(AssocSeq), nrow = 4) +
  labs(x = 'Clonotype frequency', y = expression(-log[2]*' clonotype pGen')) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 7),
        axis.text.y = element_text(size = 7),
        panel.grid = element_blank(),
        strip.text = element_text(size = 7))
```

```{r nair_fipclusters_subset_alpha}
library(igraph)
library(ggraph)
library(ggseqlogo)

# subset and plot nodes from key clusters, i.e. top-ranked clusters from hmap (ranked by delta mean freq)
# Control clusters - 3, 44
# cluster 3
control3.net <- buildNet(data = alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 3,],
                      seq_col = 'CDR3.aa')

# set node size to frequency
V(control3.net$igraph)$size <- control3.net$node_data$Proportion
# set vertex group id
V(control3.net$igraph)$group <- control3.net$node_data$IHC
# get nice plot
control3.gg <- ggraph(control3.net$igraph, layout = 'fr', maxiter = 10000) +
  geom_edge_link(color = 'darkgrey', edge_width = 0.2) +
  geom_node_point(aes(size = size, fill = group), shape = 21) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', 'darkgray')) +
  scale_size_continuous(limits = c(0,0.18), range = c(3, 8)) +
  labs(size = '', title = '', fill = '') +
  theme(panel.background = element_blank(),
        plot.title = element_text(hjust = 0.5))

# cluster 44
control44.net <- buildNet(data = alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 44,],
                      seq_col = 'CDR3.aa')

# set node size to frequency
V(control44.net$igraph)$size <- control44.net$node_data$Proportion
# set vertex group id
V(control44.net$igraph)$group <- control44.net$node_data$IHC
# get nice plot
control44.gg <- ggraph(control44.net$igraph, layout = 'fr', maxiter = 10000) +
  geom_edge_link(color = 'darkgrey', edge_width = 0.2) +
  geom_node_point(aes(size = size, fill = group), shape = 21) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', 'darkgray')) +
  scale_size_continuous(limits = c(0,0.18), range = c(3, 8)) +
  labs(size = '', title = '', fill = '') +
  theme(panel.background = element_blank(),
        plot.title = element_text(hjust = 0.5))

# FIP clusters - 106, 86, 96, 80
# cluster 106 - max freq 0.0002625361
fip106.net <- buildNet(data = alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 106,],
                      seq_col = 'CDR3.aa')

# set node size to frequency
V(fip106.net$igraph)$size <- fip106.net$node_data$Proportion
# set vertex group id
V(fip106.net$igraph)$group <- fip106.net$node_data$IHC
# get nice plot
fip106.gg <- ggraph(fip106.net$igraph, layout = 'fr', maxiter = 10000) +
  geom_edge_link(color = 'darkgrey', edge_width = 0.2) +
  geom_node_point(aes(size = size, fill = group), shape = 21) +
  scale_fill_manual(values = c('#D071AD', 'darkgray')) +
  scale_size_continuous(limits = c(0,0.18), range = c(3, 8)) +
  labs(size = '', title = '', fill = '') +
  theme(panel.background = element_blank(),
        plot.title = element_text(hjust = 0.5))

# cluster 86 - max freq 0.0055132581
fip86.net <- buildNet(data = alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 86,],
                      seq_col = 'CDR3.aa')

# set node size to frequency
V(fip86.net$igraph)$size <- fip86.net$node_data$Proportion
# set vertex group id
V(fip86.net$igraph)$group <- fip86.net$node_data$IHC
# get nice plot
fip86.gg <- ggraph(fip86.net$igraph, layout = 'fr', maxiter = 10000) +
  geom_edge_link(color = 'darkgrey', edge_width = 0.2) +
  geom_node_point(aes(size = size, fill = group), shape = 21) +
  scale_fill_manual(values = c('#8F3092', 'darkgray')) +
  scale_size_continuous(limits = c(0,0.18), range = c(3, 8)) +
  labs(size = '', title = '', fill = '') +
  theme(panel.background = element_blank(),
        plot.title = element_text(hjust = 0.5))

# cluster 96 - max freq 0.1034392229
fip96.net <- buildNet(data = alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 96,],
                      seq_col = 'CDR3.aa')

# set node size to frequency
V(fip96.net$igraph)$size <- fip96.net$node_data$Proportion
# set vertex group id
V(fip96.net$igraph)$group <- fip96.net$node_data$IHC
# get nice plot
fip96.gg <- ggraph(fip96.net$igraph, layout = 'fr', maxiter = 10000) +
  geom_edge_link(color = 'darkgrey', edge_width = 0.2) +
  geom_node_point(aes(size = size, fill = group), shape = 21) +
  scale_fill_manual(values = c('#D071AD', '#8F3092')) +
  scale_size_continuous(limits = c(0,0.18), range = c(3, 8)) +
  labs(size = '', title = '', fill = '') +
  theme(panel.background = element_blank(),
        plot.title = element_text(hjust = 0.5))

# cluster 80 - max freq 0.1671042268
fip80.net <- buildNet(data = alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 80,],
                      seq_col = 'CDR3.aa')

# set node size to frequency
V(fip80.net$igraph)$size <- fip80.net$node_data$Proportion
# set vertex group id
V(fip80.net$igraph)$group <- fip80.net$node_data$IHC
# get nice plot
fip80.gg <- ggraph(fip80.net$igraph, layout = 'fr', maxiter = 10000) +
  geom_edge_link(color = 'darkgrey', edge_width = 0.2) +
  geom_node_point(aes(size = size, fill = group), shape = 21) +
  scale_fill_manual(values = c('#D071AD', '#8F3092')) +
  scale_size_continuous(limits = c(0,0.18), range = c(3, 8)) +
  labs(size = '', title = '', fill = '') +
  theme(panel.background = element_blank(),
        plot.title = element_text(hjust = 0.5))

plot_grid(control3.gg, fip106.gg, fip96.gg, control44.gg, fip86.gg, fip80.gg, nrow = 2)


# seqlogo
control3.seq <- alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 3,]
control3.logo <- ggseqlogo(control3.seq$CDR3.aa, method = 'probability')

control44.seq <- alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 44,]
control44.logo <- ggseqlogo(control44.seq$CDR3.aa, method = 'probability')

fip106.seq <- alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 106,]
fip106.logo <- ggseqlogo(fip106.seq$CDR3.aa, method = 'probability')

fip86.seq <- alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 86,]
fip86.logo <- ggseqlogo(fip86.seq$CDR3.aa, method = 'probability')

fip96.seq <- alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 96,]
fip96.logo <- ggseqlogo(fip96.seq$CDR3.aa, method = 'probability')

fip80.seq <- alpha.all_clusters$node_data[alpha.all_clusters$node_data$cluster_id == 80,]
fip80.logo <- ggseqlogo(fip80.seq$CDR3.aa, method = 'probability')

plot_grid(control3.logo, fip106.logo, fip96.logo, control44.logo, fip86.logo, fip80.logo, nrow = 2)
```

```{r nair_fipclusters_beta}
# stash data
beta.dat <- trb_mixcr.f.ds$data
control.dat <- trb_mixcr.f.ds$data[1:22]
fcovneg.samples <- c("fip_81", "fip_92", "fip_96", "fip_142", "fip_146", "fip_151", "fip_192")
fcovplus.samples <- c("fip_34", "fip_35", "fip_36", "fip_68", "fip_86", "fip_120", "fip_172")

# save each sample as rds
lapply(names(beta.dat), function(x) saveRDS(beta.dat[[x]], file = paste0('nair/tcr_manuscript/beta/', paste0(x, '.rds'))))

# define input files
beta.input.files <- list.files(path = 'nair/tcr_manuscript/beta', pattern = '.rds', full.names = TRUE)
group_labels <- c(rep('Control', 22), rep('FIP', 19))

# find associated seqs - takes a while to run so results saved and loaded from a txt file
# beta.associated_seqs <- findAssociatedSeqs(file_list = beta.input.files,
#                                       input_type = 'rds',
#                                       group_ids = group_labels,
#                                       seq_col = 'CDR3.aa',
#                                       freq_col = 'Proportion',
#                                       min_seq_length = NULL,
#                                       drop_matches = NULL,
#                                       min_sample_membership = NULL,
#                                       pval_cutoff = 0.05)

# write.table(beta.associated_seqs, file = 'beta_associated_seqs_tcr_manuscript.txt', sep = '\t')
beta.associated_seqs <- read.csv('beta_associated_seqs_tcr_manuscript.txt', sep = '\t')

# find associated clones (top 20 seqs)
findAssociatedClones(file_list = beta.input.files,
                     input_type = 'rds',
                     group_ids = group_labels,
                     seq_col = 'CDR3.aa',
                     assoc_seqs = beta.associated_seqs$ReceptorSeq,
                     min_seq_length = NULL,
                     drop_matches = NULL,
                     output_dir = 'nair/tcr_manuscript/beta/assoc_seq_nbds',
                     sample_ids = names(beta.dat))

# combine neighbourhoods and perform network analysis
beta.nbd.input <- list.files('nair/tcr_manuscript/beta/assoc_seq_nbds', full.names = TRUE)

beta.all_clusters <- buildAssociatedClusterNetwork(file_list = beta.nbd.input,
                                                    seq_col = 'CDR3.aa',
                                                    size_nodes_by = 'Proportion',
                                                    print_plots = TRUE,
                                                    drop_isolated_nodes = FALSE)

# export CDR3s to get pGens with OLGA
beta.all_clusters.cdr3 <- data.frame(unique(beta.all_clusters$node_data$CDR3.aa))
# write.table(beta.all_clusters.cdr3, file = 'beta_all_clusters_unique_cdr3s_tcr_manuscript.txt', sep = '\t')

# add IHC group to node data
control.samples <- names(control.dat)
beta.all_clusters$node_data$IHC <- ifelse(beta.all_clusters$node_data$SampleID %in% control.samples, 'Control', 
                                           ifelse(beta.all_clusters$node_data$SampleID %in% fcovneg.samples, 'FCoV -',
                                                  ifelse(beta.all_clusters$node_data$SampleID %in% fcovplus.samples, 'FCoV +', 'FCoV ND')))
# export network for visualisation in gephi
beta.all_clusters.elist <- igraph::as_edgelist(beta.all_clusters$igraph, names = T)
write.table(beta.all_clusters.elist, file = 'beta_all_clusters_edgelist_tcr_manuscript.csv', sep = ',')
write.table(beta.all_clusters$node_data, file = 'beta_all_clusters_nodedata_tcr_manuscript.csv', sep = ',')
```

Prepare network in Gephi.

```{r nair_fipclusters_hmap_beta}
# for all clusters
beta.all <- beta.all_clusters$node_data
# count of clonotypes
clntyp.count.matrix <- unclass(table(beta.all$AssocSeq, beta.all$SampleID))
# mean frequency
clntyp.freq.matrix <- reshape2::acast(beta.all, cluster_id ~ SampleID, value.var = 'Proportion', fun.aggregate = mean, fill = 0)
# add columns for missing samples
clntyp.freq.matrix <- as.data.frame(clntyp.freq.matrix)
clntyp.freq.matrix$control_56 <- 0
clntyp.freq.matrix$control_57 <- 0
clntyp.freq.matrix$control_78 <- 0
clntyp.freq.matrix$control_90 <- 0
clntyp.freq.matrix$control_91 <- 0
clntyp.freq.matrix$control_102 <- 0
clntyp.freq.matrix$control_111 <- 0
clntyp.freq.matrix$control_130 <- 0
clntyp.freq.matrix$control_135 <- 0
clntyp.freq.matrix$control_188 <- 0
clntyp.freq.matrix$fip_129 <- 0
clntyp.freq.matrix <- as.matrix(clntyp.freq.matrix)

# change column order to group by IHC
clntyp.freq.ihc <- clntyp.freq.matrix[,ihc.sample_order]

# set row order by delta mean freq
clntyp.freq.ihc.df <- as.data.frame(clntyp.freq.ihc)
clntyp.freq.ihc.df$control_mean <- rowMeans(subset(clntyp.freq.ihc.df, select = c(1:22)))
clntyp.freq.ihc.df$fip_mean <- rowMeans(subset(clntyp.freq.ihc.df, select = c(23:41)))
clntyp.freq.ihc.df$delta <- clntyp.freq.ihc.df$fip_mean - clntyp.freq.ihc.df$control_mean
clntyp.freq.ihc.df <- clntyp.freq.ihc.df[order(clntyp.freq.ihc.df$delta, decreasing = FALSE),]
# add sample count for alt plot
clntyp.freq.ihc.df$nsamples <- rowSums(clntyp.freq.ihc.df[,1:41] > 0)
# and control/fip-exclusive clusters
clntyp.freq.ihc.df$control_excl <- clntyp.freq.ihc.df$fip_mean == 0
clntyp.freq.ihc.df$fip_excl <- clntyp.freq.ihc.df$control_mean == 0
# convert to matrix for hmap
clntyp.freq.ihc.mat <- as.matrix(clntyp.freq.ihc.df[,1:41])

library(pheatmap)
library(RColorBrewer)
breaks = seq(0, 0.0007, 0.00001)
hmcols = colorRampPalette(brewer.pal(9, 'OrRd'))(length(breaks))
pheatmap(clntyp.freq.ihc.mat,
         breaks = breaks,
         color = hmcols,
         border_color = NA,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         show_rownames = TRUE,
         fontsize_row = 7,
         angle_col = 45,
         gaps_col = c(22,29,36))

# plot delta mean freq to add to main hmap in illustrator
paletteLength <- 1000
delta.cols <- colorRampPalette(c('#32A7CD', 'white', '#8F3092'))(paletteLength)
# length(breaks) == length(paletteLength) + 1
# use floor and ceiling to deal with even/odd length pallettelengths
myBreaks <- c(seq(min(clntyp.freq.ihc.df$delta), 0, length.out=ceiling(paletteLength/2) + 1), 
              seq(max(clntyp.freq.ihc.df$delta)/paletteLength, max(clntyp.freq.ihc.df$delta), length.out=floor(paletteLength/2)))

pheatmap(as.matrix(clntyp.freq.ihc.df$delta), 
         breaks = myBreaks, 
         color = delta.cols, 
         cluster_rows = FALSE, 
         cluster_cols = FALSE, 
         border_color = NA)
```


```{r nair_fipclusters_pgenplots_beta}
library(ggrepel)

# read in pGens from OLGA and merge with clone node data
beta.all_clusters.pgens <- read.csv('beta_all_clusters_unique_cdr3s_tcr_manuscript_pgens.tsv', sep = '\t', header = TRUE)
beta_pgens <- merge(beta.all_clusters$node_data, beta.all_clusters.pgens, by.x = 'CDR3.aa', by.y = 'Seq', all.x = TRUE)

# calculate median cluster pGen
beta.assoccluster.pgen <- beta_pgens %>%
  group_by(cluster_id) %>%
  summarise(medianpgen = median(Score, na.rm = TRUE))

# get cluster data for plotting
clntyp.freq.ihc.df$cluster_id <- rownames(clntyp.freq.ihc.df)
freq.pgen.plot <- subset(clntyp.freq.ihc.df, select = c(cluster_id, delta, nsamples, control_excl, fip_excl))
freq.pgen.plot <- merge(freq.pgen.plot, beta.assoccluster.pgen, by = 'cluster_id', all.x = TRUE)

# add seq_w_max_degree for labelling
beta.clusterstat <- subset(beta.all_clusters$cluster_data, select = c(cluster_id, seq_w_max_degree))
freq.pgen.plot <- merge(freq.pgen.plot, beta.clusterstat, by = 'cluster_id', all.x = TRUE)

# plot
ggplot(freq.pgen.plot, aes(x = delta, y = medianpgen, size = nsamples)) +
  geom_vline(xintercept = 0, color = 'darkgrey', lty = 'dashed') +
  geom_point(shape = 21) +
  geom_point(data = freq.pgen.plot[freq.pgen.plot$fip_excl == TRUE,], aes(x = delta, y = medianpgen, size = nsamples), shape = 21, fill = '#8F3092') +
  geom_point(data = freq.pgen.plot[freq.pgen.plot$control_excl == TRUE,], aes(x = delta, y = medianpgen, size = nsamples), shape = 21, fill = '#32A7CD') +
  # geom_text_repel(aes(label = seq_w_max_degree, size = 5), show.legend = FALSE) +
  scale_x_continuous(limits = c(-0.0001,0.0001), breaks = seq(-0.0001, 0.0001, 2.5e-05)) +
  # scale_y_continuous(limits = c(50, 140), breaks = seq(50, 140, 10)) +
  scale_size_continuous(breaks = seq(5,25,10)) +
  labs(x = 'Δ cluster mean frequency (FIP - Control)', y = expression(-log[2]*' cluster median pGen')) +
  theme_bw() +
  theme(panel.grid = element_blank())

# plot zoomed view of selected clusters
# top 20?
clusters.plot <- beta.associated_seqs[order(beta.associated_seqs$fisher_pvalue, decreasing = FALSE),]$ReceptorSeq[1:20]

beta.pgens.plot <- beta_pgens[beta_pgens$AssocSeq %in% clusters.plot,]

ggplot(beta.pgens.plot, aes(Proportion, Score, fill = IHC)) +
  geom_point(shape = 21) +
  scale_fill_manual(values = c('#32A7CD', '#D071AD', '#8F3092', 'darkgray')) +
  facet_wrap(vars(AssocSeq), nrow = 4) +
  labs(x = 'Clonotype frequency', y = expression(-log[2]*' clonotype pGen')) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 7),
        axis.text.y = element_text(size = 7),
        panel.grid = element_blank(),
        strip.text = element_text(size = 7))
```

```{r nair_fipclusters_subset_beta}
library(igraph)
library(ggraph)
library(ggseqlogo)

# subset and plot nodes from key clusters, i.e. top-ranked clusters from hmap (ranked by delta mean freq)
# cluster 12
fip12.net <- buildNet(data = beta.all_clusters$node_data[beta.all_clusters$node_data$cluster_id == 12,],
                      seq_col = 'CDR3.aa')

# set node size to frequency
V(fip12.net$igraph)$size <- fip12.net$node_data$Proportion
# set vertex group id
V(fip12.net$igraph)$group <- fip12.net$node_data$IHC
# get nice plot
fip12.gg <- ggraph(fip12.net$igraph, layout = 'fr', maxiter = 10000) +
  geom_edge_link(color = 'darkgrey', edge_width = 0.2) +
  geom_node_point(aes(size = size, fill = group), shape = 21) +
  scale_fill_manual(values = c('#D071AD', '#8F3092', 'darkgray')) +
  scale_size_continuous(limits = c(0,0.0011), range = c(2, 8)) +
  labs(size = '', title = '', fill = '') +
  theme(panel.background = element_blank(),
        plot.title = element_text(hjust = 0.5))

# seqlogo
fip12.seq <- beta.all_clusters$node_data[beta.all_clusters$node_data$cluster_id == 12,]
fip12.logo <- ggseqlogo(fip12.seq$CDR3.aa, method = 'probability')
```

## Additional plots from RNA-seq repertoire analysis

Correlation between TCR-seq and RNA-seq.

```{r corr}
# read in data
corr <- read.csv('top_clonotypes_tra_targetvsrnaseq.txt', sep = '\t', header = TRUE)

topclon.gg <- ggplot(corr, aes(sqrt(rnaseq), sqrt(tcrseq), color = detection)) +
  geom_point(size = 2) +
  scale_color_manual(values = c('navyblue', 'lightgrey')) +
  labs(x = 'Square root top TRA clonotype proportion (RNA-seq)', y = 'Square root top TRA clonotype proportion (targeted)', color = 'Detected in') +
  theme_bw() +
  theme(panel.grid = element_blank())

# clonotype correlation
clntype.corr <- read.csv('rnaseq_tcrseq_clntype_corr.txt', sep = '\t', header = TRUE)
clntype.corr$Group <- c(rep('Control', 7), 'FIP', 'FIP (focal renal)', 'FIP', 'FIP (focal renal)', rep('FIP', 4))

clntyp.gg <- ggplot(clntype.corr, aes(rnaseq, tcrseq, color = Group)) +
  geom_point(size = 2) +
  scale_x_continuous(limits = c(1,4)) +
  scale_y_continuous(limits = c(1,4)) +
  scale_color_manual(values = c('#32A7CD', '#8F3092', '#D071AD')) +
  geom_abline(slope = 1, lty = 'dashed', color = 'darkgrey') +
  labs(x = 'log10 TRA clonotypes (RNA-seq)', y = 'log10 TRA clonotypes (targeted)', color = '') +
  theme_bw() +
  theme(panel.grid = element_blank())
```


